---
title: Differential intron retention in breast cancer cell lines treated with splicing
  inhibitors
author: "Monique White"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(data.table)
library(ggpubr)
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
library(readxl)
source("ir_helpers.R")   # for make_ir_id()
library(ggsignif)

metadata_sc5903_mw <- read_csv("metadata_sc5903_mw.csv")
metadata_sc5903_mw <- metadata_sc5903_mw |> 
  mutate(conc_uM = factor(conc_uM, ordered = TRUE))
```

# 1. Download data
```{bash eval=FALSE, echo=FALSE}
#!/usr/bin/env bash
# =============================================================================
# Master FASTQ download script
# Studies: SRP112607 (4 runs), SRP141379 (128 runs), E-MTAB-14664 (6 runs)
# Total: 138 runs — FTP paths are fetched live from ENA API at runtime
# Resumable: re-run to skip completed files and resume partial transfers
# Integrity-checked: gzip-tested after download, MD5 verified at the end
# =============================================================================

set -euo pipefail

BASE_DIR="/data/moniquew/BrCa-Special-Topics/raw_data"
LOG_DIR="/data/moniquew/BrCa-Special-Topics/logs"
mkdir -p "$LOG_DIR"

# =============================================================================
# Core helpers
# =============================================================================

_is_complete() {
  local f="$1"
  [ -s "$f" ] || return 1
  [ -f "$f.done" ] && [ ! "$f" -nt "$f.done" ] && return 0
  case "$f" in
  *.gz) gzip -t "$f" 2>/dev/null ;;
*) return 0 ;;
esac
}

_human() { numfmt --to=iec --suffix=B --format='%.1f' "$1" 2>/dev/null || echo "${1}B"; }

_done_count=0
_total_count=0

download_one() {
  local url="$1" out="$2" expected="${3:-0}"
  _done_count=$((_done_count + 1))
  local prefix="[$_done_count/$_total_count] $(basename $out)"
  local sz=0
  [ -f "$out" ] && sz=$(stat -c %s "$out" 2>/dev/null || echo 0)
  local sz_msg=""
  [ "$expected" -gt 0 ] && sz_msg=" ($(_human $sz) / $(_human $expected))"
  printf '%s: checking%s... ' "$prefix" "$sz_msg" >&2
  
  if [ "$expected" -gt 0 ] && [ "$sz" -gt "$expected" ]; then
  echo "oversized, redownloading" >&2
  rm -f "$out" "$out.done"
  elif _is_complete "$out"; then
  : > "$out.done"
  echo "valid, skip" >&2
  return 0
  elif [ "$sz" -gt 0 ]; then
  echo "partial, resuming" >&2
  rm -f "$out.done"
  wget -c -q --show-progress --tries=10 --timeout=60 --retry-connrefused -O "$out" "$url" || true
  if _is_complete "$out"; then
  : > "$out.done"
  echo "$prefix: resumed OK" >&2
  return 0
  fi
  echo "$prefix: WARN integrity check failed after resume; redownloading from scratch" >&2
  rm -f "$out" "$out.done"
  else
    echo "missing, downloading" >&2
  rm -f "$out.done"
  fi
  
  mkdir -p "${out%/*}"
  wget -q --show-progress --tries=10 --timeout=60 --retry-connrefused -O "$out" "$url"
  _is_complete "$out" || {
    echo "$prefix: ERR corrupted after download (re-run script)" >&2
    rm -f "$out" "$out.done"
    return 1
  }
  : > "$out.done"
  echo "$prefix: done" >&2
}

# Fetch FTP paths + MD5 + sizes from ENA API for a given accession
# Prints: run_accession <TAB> ftp_url1;ftp_url2 <TAB> md5_1;md5_2 <TAB> bytes_1;bytes_2
fetch_ena_info() {
  local accession="$1"
  curl -sS --retry 5 --retry-delay 3 \
  "https://www.ebi.ac.uk/ena/portal/api/filereport?accession=${accession}&result=read_run&fields=run_accession,fastq_ftp,fastq_md5,fastq_bytes&format=tsv" \
  | tail -n +2   # strip header
}

# =============================================================================
# Download one study worth of runs
# Arguments: study_id  base_subdir  [list of run accessions...]
# =============================================================================
download_study() {
  local study="$1"
  local subdir="$2"
  shift 2
  local runs=("$@")
  
  echo ""
  echo "============================================================"
  echo "  Study: $study  (${#runs[@]} runs)"
  echo "  Output: $BASE_DIR/$subdir"
  echo "============================================================"
  
  # Collect all file info up front
  declare -a all_urls all_outs all_sizes all_md5s
  all_urls=(); all_outs=(); all_sizes=(); all_md5s=()
  
  echo "Querying ENA API for FTP paths..." >&2
  for run in "${runs[@]}"; do
  local info
  info=$(fetch_ena_info "$run")
  if [ -z "$info" ]; then
  echo "WARN: No ENA data found for $run — skipping" >&2
  continue
  fi
  
  while IFS=$'\t' read -r run_acc ftp_urls md5s bytes; do
  [ -z "$ftp_urls" ] && continue
  IFS=';' read -ra urls  <<< "$ftp_urls"
  IFS=';' read -ra mds   <<< "$md5s"
  IFS=';' read -ra szs   <<< "$bytes"
  for i in "${!urls[@]}"; do
  local url="ftp://${urls[$i]}"
  local fname; fname=$(basename "${urls[$i]}")
  local out="$BASE_DIR/$subdir/$run_acc/$fname"
  local sz="${szs[$i]:-0}"
  local md="${mds[$i]:-}"
  all_urls+=("$url")
  all_outs+=("$out")
  all_sizes+=("$sz")
  all_md5s+=("$md  $out")
  done
  done <<< "$info"
  done
  
  _total_count=${#all_urls[@]}
    _done_count=0
    
    # Resume estimate
    local present=0 local_bytes=0 total_bytes=0
    for i in "${!all_outs[@]}"; do
    local exp=${all_sizes[$i]:-0}
    total_bytes=$((total_bytes + exp))
    local f=${all_outs[$i]}
    if [ -f "$f" ]; then
    local sz; sz=$(stat -c %s "$f" 2>/dev/null || echo 0)
    [ "$sz" -gt "$exp" ] && sz=0
    local_bytes=$((local_bytes + sz))
    [ "$sz" -ge "$exp" ] && [ "$exp" -gt 0 ] && present=$((present + 1))
    fi
    done
    local pct=$(( total_bytes > 0 ? local_bytes * 100 / total_bytes : 0 ))
    echo "Resume estimate: $present/$_total_count files complete ($(_human $local_bytes) / $(_human $total_bytes) = ${pct}%)."
    
    # Download
    for i in "${!all_urls[@]}"; do
    download_one "${all_urls[$i]}" "${all_outs[$i]}" "${all_sizes[$i]:-0}"
    done
    
    # MD5 check
    echo ""
    echo "Verifying MD5 checksums for $study..."
    local md5_input=""
    for entry in "${all_md5s[@]}"; do
    [ -n "$entry" ] && md5_input+="$entry"$'\n'
    done
    if [ -n "$md5_input" ]; then
    echo "$md5_input" | md5sum -c --quiet && echo "All checksums OK for $study" || echo "WARN: Some checksums failed for $study — check above"
    else
      echo "No MD5s available to check for $study"
    fi
    
    echo "Done: $study → $BASE_DIR/$subdir"
  }
  
  # =============================================================================
  # Study definitions — run accessions from metadata CSV
  # =============================================================================
  
  # --- SRP112607 (4 runs, 4 experiments) ---
  SRP112607_RUNS=(
    SRR5833360
    SRR5833361
    SRR5833362
    SRR5833363
  )
  
  # --- SRP141379 (128 runs, 16 experiments) ---
  SRP141379_RUNS=(
    SRR7048256 SRR7048257 SRR7048258 SRR7048259
    SRR7048260 SRR7048261 SRR7048262 SRR7048263
    SRR7048264 SRR7048265 SRR7048266 SRR7048267
    SRR7048268 SRR7048269 SRR7048270 SRR7048271
    SRR7048272 SRR7048273 SRR7048274 SRR7048275
    SRR7048276 SRR7048277 SRR7048278 SRR7048279
    SRR7048280 SRR7048281 SRR7048282 SRR7048283
    SRR7048284 SRR7048285 SRR7048286 SRR7048287
    SRR7048288 SRR7048289 SRR7048290 SRR7048291
    SRR7048292 SRR7048293 SRR7048294 SRR7048295
    SRR7048296 SRR7048297 SRR7048298 SRR7048299
    SRR7048300 SRR7048301 SRR7048302 SRR7048303
    SRR7048304 SRR7048305 SRR7048306 SRR7048307
    SRR7048308 SRR7048309 SRR7048310 SRR7048311
    SRR7048312 SRR7048313 SRR7048314 SRR7048315
    SRR7048316 SRR7048317 SRR7048318 SRR7048319
    SRR7048320 SRR7048321 SRR7048322 SRR7048323
    SRR7048324 SRR7048325 SRR7048326 SRR7048327
    SRR7048328 SRR7048329 SRR7048330 SRR7048331
    SRR7048332 SRR7048333 SRR7048334 SRR7048335
    SRR7048336 SRR7048337 SRR7048338 SRR7048339
    SRR7048340 SRR7048341 SRR7048342 SRR7048343
    SRR7048344 SRR7048345 SRR7048346 SRR7048347
    SRR7048348 SRR7048349 SRR7048350 SRR7048351
    SRR7048352 SRR7048353 SRR7048354 SRR7048355
    SRR7048356 SRR7048357 SRR7048358 SRR7048359
    SRR7048360 SRR7048361 SRR7048362 SRR7048363
    SRR7048364 SRR7048365 SRR7048366 SRR7048367
    SRR7048368 SRR7048369 SRR7048370 SRR7048371
    SRR7048372 SRR7048373 SRR7048374 SRR7048375
    SRR7048376 SRR7048377 SRR7048378 SRR7048379
    SRR7048380 SRR7048381 SRR7048382 SRR7048383
  )
  
  # --- E-MTAB-14664 (6 runs) — already handled by download_ERP171667.sh ---
  # Included here for completeness / in case you want one unified script.
  # Comment this out if ERP171667 is already downloading separately.
  EMTAB14664_RUNS=(
    ERR14856082
    ERR14856083
    ERR14856084
    ERR14856085
    ERR14856086
    ERR14856087
  )
  
  # =============================================================================
  # Run all studies sequentially
  # =============================================================================
  
  echo "Starting downloads: $(date)"
  echo "Output root: $BASE_DIR"
  
  download_study "SRP112607"   "SRP112607"   "${SRP112607_RUNS[@]}"
  download_study "SRP141379"   "SRP141379"   "${SRP141379_RUNS[@]}"
  download_study "E-MTAB-14664" "E-MTAB-14664/ERP171667" "${EMTAB14664_RUNS[@]}"
  
  echo ""
  echo "============================================================"
  echo "  All studies complete: $(date)"
  echo "  Files saved under: $BASE_DIR"
  echo "============================================================"
```


# 2. Fastqc data
Aim: identify adaptors for trimming prior to running IRFinder.


# 3. Preparing data for IRFinder
## Trimming adaptors
Used cutadapt to trim the 15bp adaptors from all samples
```{bash eval=FALSE, echo=FALSE}
# --- Create output directory ---
#mkdir -p trimmed_data

# --- Loop through all _1.fastq.gz files ---
find /bioinformatics/moniquew/BrCa-Special-Topics/raw_data/SRP112607/ /bioinformatics/moniquew/BrCa-Special-Topics/raw_data/SRP141379/ /bioinformatics/moniquew/BrCa-Special-Topics/raw_data/E-MTAB-14664/ \
    -name "*_1.fastq.gz" ! -name "*.done" \
    -type f | while read R1; do

    # Derive the R2 filename by replacing _1.fastq.gz with _2.fastq.gz
    R2="${R1/_1.fastq.gz/_2.fastq.gz}"

    # Get the sample name (e.g. SRR5833363) from the filename
    SAMPLE=$(basename "$R1" _1.fastq.gz)

    # Define output filenames
    R1_OUT="trimmed_data/${SAMPLE}_1_trimmed.fastq.gz"
    R2_OUT="trimmed_data/${SAMPLE}_2_trimmed.fastq.gz"
    R1_UNPAIRED="trimmed_data/${SAMPLE}_1_unpaired.fastq.gz"
    R2_UNPAIRED="trimmed_data/${SAMPLE}_2_unpaired.fastq.gz"

    echo "Trimming $SAMPLE: $(date)"

    cutadapt \
        -u 15 \
        -U 15 \
        -j 8 \
        -o "$R1_OUT" \
        -p "$R2_OUT" \
        "$R1" \
        "$R2"


    echo "Done $SAMPLE: $(date)"

done

echo "Cutadapt pipeline finished: $(date)"
```


## Merging lanes
Zhu et al had samples spread across multiple lanes, and thus required merging
```{bash, eval=FALSE, echo=FALSE}
#!/usr/bin/env bash
# Merge per-lane trimmed FASTQs into per-biological-replicate FASTQs
# Study: SRP141379 (Zhu et al. 2018) — CAL51, CC-671 treatment
# 16 biological replicates across 4 conditions x 2 timepoints
# Each biological replicate was sequenced across 8 lanes → merged here into 1

set -euo pipefail

TRIMMED="/bioinformatics/moniquew/BrCa-Special-Topics/trimmed_data"
OUT="/bioinformatics/moniquew/BrCa-Special-Topics/merged_data"

mkdir -p "$OUT"

echo "Starting lane merging — $(date)"

# ---------------------------------------------------------------------------
# Helper function: merge R1 and R2 for a set of SRR runs into one output sample
# Usage: merge_sample <sample_name> <SRR1> <SRR2> ... 
# ---------------------------------------------------------------------------
merge_sample() {
  local name="$1"; shift
  local srrs=("$@")
  local r1_files=() r2_files=()

  for srr in "${srrs[@]}"; do
    r1_files+=("$TRIMMED/${srr}_1_trimmed.fastq.gz")
    r2_files+=("$TRIMMED/${srr}_2_trimmed.fastq.gz")
  done

  echo "  Merging ${#srrs[@]} lanes → ${name} ..."
  cat "${r1_files[@]}" > "$OUT/${name}_R1.fastq.gz"
  cat "${r2_files[@]}" > "$OUT/${name}_R2.fastq.gz"
  echo "  Done: ${name}"
}

# ---------------------------------------------------------------------------
# DMSO controls — 6hr (2 biological replicates)
# ---------------------------------------------------------------------------
echo "[1/16] DMSO_6hr_rep1 (SRX3979816: SRR7048256–263)"
merge_sample "DMSO_6hr_rep1" \
  SRR7048256 SRR7048257 SRR7048258 SRR7048259 \
  SRR7048260 SRR7048261 SRR7048262 SRR7048263

echo "[2/16] DMSO_6hr_rep2 (SRX3979817: SRR7048264–271)"
merge_sample "DMSO_6hr_rep2" \
  SRR7048264 SRR7048265 SRR7048266 SRR7048267 \
  SRR7048268 SRR7048269 SRR7048270 SRR7048271

# ---------------------------------------------------------------------------
# DMSO controls — 24hr (2 biological replicates)
# ---------------------------------------------------------------------------
echo "[3/16] DMSO_24hr_rep1 (SRX3979818: SRR7048272–279)"
merge_sample "DMSO_24hr_rep1" \
  SRR7048272 SRR7048273 SRR7048274 SRR7048275 \
  SRR7048276 SRR7048277 SRR7048278 SRR7048279

echo "[4/16] DMSO_24hr_rep2 (SRX3979819: SRR7048280–287)"
merge_sample "DMSO_24hr_rep2" \
  SRR7048280 SRR7048281 SRR7048282 SRR7048283 \
  SRR7048284 SRR7048285 SRR7048286 SRR7048287

# ---------------------------------------------------------------------------
# CC-671 3uM — 6hr (3 biological replicates)
# ---------------------------------------------------------------------------
echo "[5/16] CC671_3uM_6hr_rep1 (SRX3979820: SRR7048288–295)"
merge_sample "CC671_3uM_6hr_rep1" \
  SRR7048288 SRR7048289 SRR7048290 SRR7048291 \
  SRR7048292 SRR7048293 SRR7048294 SRR7048295

echo "[6/16] CC671_3uM_6hr_rep2 (SRX3979821: SRR7048296–303)"
merge_sample "CC671_3uM_6hr_rep2" \
  SRR7048296 SRR7048297 SRR7048298 SRR7048299 \
  SRR7048300 SRR7048301 SRR7048302 SRR7048303

echo "[7/16] CC671_3uM_6hr_rep3 (SRX3979822: SRR7048304–311)"
merge_sample "CC671_3uM_6hr_rep3" \
  SRR7048304 SRR7048305 SRR7048306 SRR7048307 \
  SRR7048308 SRR7048309 SRR7048310 SRR7048311

# ---------------------------------------------------------------------------
# CC-671 3uM — 24hr (3 biological replicates)
# SRX3979823: SRR7048312–319  (3uM_24hr_rep1)
# SRX3979824: SRR7048320–327  (3uM_24hr_rep2)
# SRX3979825: SRR7048328–335  (3uM_24hr_rep3)
# ---------------------------------------------------------------------------
echo "[8/16] CC671_3uM_24hr_rep1 (SRX3979823: SRR7048312–319)"
merge_sample "CC671_3uM_24hr_rep1" \
  SRR7048312 SRR7048313 SRR7048314 SRR7048315 \
  SRR7048316 SRR7048317 SRR7048318 SRR7048319

echo "[9/16] CC671_3uM_24hr_rep2 (SRX3979824: SRR7048320–327)"
merge_sample "CC671_3uM_24hr_rep2" \
  SRR7048320 SRR7048321 SRR7048322 SRR7048323 \
  SRR7048324 SRR7048325 SRR7048326 SRR7048327

echo "[10/16] CC671_3uM_24hr_rep3 (SRX3979825: SRR7048328–335)"
merge_sample "CC671_3uM_24hr_rep3" \
  SRR7048328 SRR7048329 SRR7048330 SRR7048331 \
  SRR7048332 SRR7048333 SRR7048334 SRR7048335

# ---------------------------------------------------------------------------
# CC-671 10uM — 6hr (3 biological replicates)
# SRX3979826: SRR7048336–343  (10uM_6hr_rep1)
# SRX3979827: SRR7048344–351  (10uM_6hr_rep2)
# SRX3979828: SRR7048352–359  (10uM_6hr_rep3)
# ---------------------------------------------------------------------------
echo "[11/16] CC671_10uM_6hr_rep1 (SRX3979826: SRR7048336–343)"
merge_sample "CC671_10uM_6hr_rep1" \
  SRR7048336 SRR7048337 SRR7048338 SRR7048339 \
  SRR7048340 SRR7048341 SRR7048342 SRR7048343

echo "[12/16] CC671_10uM_6hr_rep2 (SRX3979827: SRR7048344–351)"
merge_sample "CC671_10uM_6hr_rep2" \
  SRR7048344 SRR7048345 SRR7048346 SRR7048347 \
  SRR7048348 SRR7048349 SRR7048350 SRR7048351

echo "[13/16] CC671_10uM_6hr_rep3 (SRX3979828: SRR7048352–359)"
merge_sample "CC671_10uM_6hr_rep3" \
  SRR7048352 SRR7048353 SRR7048354 SRR7048355 \
  SRR7048356 SRR7048357 SRR7048358 SRR7048359

# ---------------------------------------------------------------------------
# CC-671 10uM — 24hr (3 biological replicates)
# SRX3979829: SRR7048360–367  (10uM_24hr_rep1)
# SRX3979830: SRR7048368–375  (10uM_24hr_rep2)
# SRX3979831: SRR7048376–383  (10uM_24hr_rep3)
# ---------------------------------------------------------------------------
echo "[14/16] CC671_10uM_24hr_rep1 (SRX3979829: SRR7048360–367)"
merge_sample "CC671_10uM_24hr_rep1" \
  SRR7048360 SRR7048361 SRR7048362 SRR7048363 \
  SRR7048364 SRR7048365 SRR7048366 SRR7048367

echo "[15/16] CC671_10uM_24hr_rep2 (SRX3979830: SRR7048368–375)"
merge_sample "CC671_10uM_24hr_rep2" \
  SRR7048368 SRR7048369 SRR7048370 SRR7048371 \
  SRR7048372 SRR7048373 SRR7048374 SRR7048375

echo "[16/16] CC671_10uM_24hr_rep3 (SRX3979831: SRR7048376–383)"
merge_sample "CC671_10uM_24hr_rep3" \
  SRR7048376 SRR7048377 SRR7048378 SRR7048379 \
  SRR7048380 SRR7048381 SRR7048382 SRR7048383

echo ""
echo "All done — $(date)"
echo "Merged files written to: $OUT"
echo ""
echo "Summary of output files:"
ls -lh "$OUT"/*.fastq.gz | awk '{print $5, $9}'
```

## fastqc of cleaned data
```{bash eval=FALSE, include=FALSE}

```

# 4. IRFinder

## 4.1. Build reference genome

Using 150 bp mapability files for Liu et al. 2025 (Velde 2026)
```{r build reference genome with 150bp mapability files}

```


Using 100 bp mapability file for Zhu et al. 2018 and ...
```{r build reference genome with 100bp mapability files}

```

## 4.2. Run IRFinder

### Velde et al. 2026

```{r run IRFinder for ERP171667}

```

### Zhu et al. 2018
```{r run IRFinder for SRP141379}

```

### Iwai et al. 2018
```{r run IRFinder for SRP112607}

```


# 5. Shortlisting introns / filtering

## 5.1. 
First identify introns that have a mean IR ratio > 0.05 in at least one group.
### Velde et al: either DMSO or T025

### Zhu et al: either CC671 3uM @ 6 hours, CC671 10uM @ 6 hours, CC671 3uM @ 24 hours, or CC671 10uM @ 24 hours.

## 5.2. Within sample filtering
Met the criteria outlined in Shah et al. 2022.
1.
2.
3.
4.

```{bash eval=FALSE, include=FALSE}

```

# Quick summary statistics

## Intron retention events

```{r}
metadata_sc5903_mw |> summarise(
  mean_intron = mean(sample_introns),
  se_introns = sd(sample_introns) / sqrt(n()),
  n_samples = n()
)

metadata_sc5903_mw |> 
  group_by(Hours) |> 
  summarise(
  mean_intron = mean(sample_introns),
  se_introns = sd(sample_introns) / sqrt(n()),
  n_samples = n()
)

metadata_sc5903_mw |> 
  group_by(Hours, exp_design) |> 
  summarise(
  mean_intron = mean(sample_introns),
  se_introns = sd(sample_introns) / sqrt(n()),
  n_samples = n()
)
```
```{r}
library(glmmTMB)
library(emmeans)

table(metadata_sc5903_mw$Hours)

model <- glmmTMB(
  sample_introns ~ Hours,
  family = nbinom2,
  data = metadata_sc5903_mw)

summary(model)
drop1(model, test = "Chisq")

emmeans(model, pairwise ~ Hours, type = "response")
```


## Sequencing depth

```{r}
metadata_sc5903_mw |> summarise(
  mean_depth = mean(`Sequencing depth (FASTQC)`),
  se_depth = sd(`Sequencing depth (FASTQC)`) / sqrt(n()),
  n_samples = n()
)

metadata_sc5903_mw |> 
  group_by(Hours) |> 
  summarise(
  mean_depth = mean(`Sequencing depth (FASTQC)`),
  se_depth = sd(`Sequencing depth (FASTQC)`) / sqrt(n()),
  n_samples = n()
)

metadata_sc5903_mw |> 
  group_by(Hours, exp_design) |> 
  summarise(
  mean_depth = mean(`Sequencing depth (FASTQC)`),
  se_depth = sd(`Sequencing depth (FASTQC)`) / sqrt(n()),
  n_samples = n()
)
```

Mean IR ratio across introns within a sample

```{r}
metadata_sc5903_mw |> summarise(
  mean_ir_ratio_all_introns = mean(mean_ir_ratio),
  se_ir_ratio_all_introns = sd(mean_ir_ratio) / sqrt(n()),
  n_samples = n()
)

metadata_sc5903_mw |> 
  group_by(Hours) |> 
  summarise(
  mean_ir_ratio_all_introns = mean(mean_ir_ratio),
  se_ir_ratio_all_introns = sd(mean_ir_ratio) / sqrt(n()),
  n_samples = n()
)

metadata_sc5903_mw |> 
  group_by(Hours, exp_design) |> 
  summarise(
  mean_ir_ratio_all_introns = mean(mean_ir_ratio),
  se_ir_ratio_all_introns = sd(mean_ir_ratio) / sqrt(n()),
  n_samples = n()
)
```


# 6. Differential intron retention events (dIR)

```{r}
# Split into the two studies
Zhu <- metadata_sc5903_mw %>% filter(Citation == "Zhu et al. 2018")
Zhu <- Zhu |>  mutate(Concentration = factor(Concentration, levels = c("0uM", "3uM", "10uM"), ordered = TRUE))

Velde <- metadata_sc5903_mw %>% filter(Citation == "Velde et al. 2026") 

Iwai <- metadata_sc5903_mw |> filter(Citation == "Iwai et al. 2018") 
Iwai <- Iwai |> mutate(Concentration = factor(Concentration, levels = c("0nM", "30nM", "100nM", "300nM"), ordered = TRUE))
```


## 6.1. Velde et al. 2026 (DMSO vs T025)

```{r import dIR data for Velde et al 2026}
Velde_T025_vs_DMSO <- read_csv("results/ir/ERP171667/T025_vs_DMSO.csv")
Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |>
  mutate("Hours" = 6,
    "Concentration" = "1uM",
    "Study" = "Velde")

Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |> 
  mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)
# Analysing introns where the delta IR ratio was > 0.05
erp_combined <- read_tsv(
  "results/ir/ERP171667_shah_filtered_combined.tsv",
  show_col_types = FALSE)
#36626 observations, because it is all introns in all samples

erp_combined <- erp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )
# #sanity check
 dim(erp_combined)
 head(erp_combined)

# Delta IR
delta_erp <- read_tsv(
  "results/ir/ERP171667_delta_ir.tsv",
  show_col_types = FALSE
) #6106 rows

delta_erp <- delta_erp |> mutate(
    EnsemblID = str_extract(IntronID, "ENSG[0-9]+"))

delta_erp_filter <- delta_erp |> filter(delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                                        delta_mean_ir_ratio_T025_vs_DMSO <= -0.1)  
#2559 introns that have a delta IR ratio of greater than absolute 0.1.
```

Summary (ERP171667):
2559 introns that have a delta IR ratio of greater than or equal to |0.1|.

### Scatterplot
Use the introns that have delta IR > absolute value of 0.1. Then compare the IR ratio in MB231 cells and those treated with T-025.

```{r}
erp_mean <- erp_combined |> 
  filter(IntronID %in% delta_erp_filter$IntronID) |> #only including introns that passed the structural filtering,
  #and have a delta IR ratio greater than or equal to |0.1|
  group_by(IntronID, condition) %>%
  summarise(
    #mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

erp_wide <- erp_mean %>%
  pivot_wider(id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio)
# 2559 Introns.

sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_erp_filter$IntronID)
#1618 dIR have a base expression of 10, FDR < 0.01, and a delta IR ratio greater than or equal to |0.1|
erp_sig_wide <- erp_wide |> filter(IntronID %in% sig_dIR_erp$intron_id)

ERP171667_ir_ratio_delta_ir <- ggplot() +
  geom_point(data = erp_wide, aes(x = T025, y = DMSO), alpha = 0.4, size = 0.5) +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  #geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "blue", alpha = 0.4, size = 1)+
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = "mean IR ratio - T-025",
    y = "mean IR ratio - DMSO")
ERP171667_ir_ratio_delta_ir_deseq_sig <- ggplot() +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "#701705", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(x = "mean IR ratio - T-025",
    y = "mean IR ratio - DMSO")
```

Figure X. Scatterplot showing intron retention events that had a mean IR ratio of greater than 0.05 in at xx cells treated with DMSO or T025 after 24 hours. Intron retention events associated with significant dIR (p-adjust < 0.01) and delta IR ratio of > 0.1 or < -0.1 are shown in blue. dIR was tested using DESeq2.

```{r}
library(dplyr)
# For the full dataset
erp_wide <- erp_wide %>%
  mutate(
    position = case_when(
      DMSO > T025 + 0.1 ~ "above",
      DMSO < T025 - 0.1 ~ "below",
      TRUE ~ "within"))

table(erp_wide$position)
1653/906
# or
erp_wide %>% count(position)
# For the significant subset
erp_sig_wide <- erp_sig_wide %>%
  mutate(
    position = case_when(
      DMSO > T025 + 0.1 ~ "above",
      DMSO < T025 - 0.1 ~ "below",
      TRUE ~ "within"))

erp_sig_wide %>% count(position)

1033/585 # fold difference?
```

1618 introns that have a base expression of 10, FDR < 0.01, and a delta IR ratio greater than or equal to |0.1|.

### Intron retention events

```{r}
mean_Velde <- Velde |>
  group_by(conc_uM, exp_design) |>
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()), #sd/sqrt(n)
    .groups  = "drop")

ERP171667_IR_burden <- ggplot() +
  geom_col(data = mean_Velde, 
           aes(x = conc_uM, 
               y = mean_dir),
           fill = "#9E9E9E",
           colour = "black") +
  geom_jitter(data = Velde, aes(x = conc_uM, y = sample_introns, shape = exp_design), width = 0.2) +
  geom_errorbar(data = mean_Velde,
    aes(x = conc_uM, 
        ymin = mean_dir - se_dir, 
        ymax = mean_dir + se_dir),
    width = 0.1) +
  theme_bw()+
  labs(y = "Intron retention events",
       x = expression("T-025 Concentration (" * mu * "M)")) +
  scale_y_continuous(limits = c(0, 5000))+
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = c(0.95, 0.99),
        legend.justification = c("right", "top"))

2959.67/2031
# ggsave(filename = "ERP171667_IR_burden.PNG",
#        plot = ERP171667_IR_burden,
#        height = 180,
#        width = 160,
#        unit = "mm")
```

Figure X. Number of intron retention events passing filtering criteria (structural criteria within samples, and  IR ratio > 0.05 in at least one group after intra-sample filtering). Error bars represent standard error of the mean. Coloured dot is the mean of the samples.

### Determine whether time or concentration drive difference in intron retention between control and treatment
```{r}
library(glmmTMB)
# Treat Hours and Concentration as factors if you only have a few discrete levels 
Velde$Concentration <- factor(Velde$Concentration)

table(Velde$Concentration)

model <- glmmTMB(
  sample_introns ~ Concentration,
  family = nbinom2,
  data = Velde)

summary(model)
drop1(model, test = "Chisq")

library(emmeans)
emmeans(model, pairwise ~ Concentration, type = "response")
```

### Volcano plot
```{r}
# merge delta IR and the dIR data
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))

volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> 
  filter(baseMean > 10)

volcano_velde |> group_by(significance) |> summarise(n_introns = n())

volc_data <- volcano_velde |>
  filter(baseMean > 10) |> #ensures that only introns with enough expression are included
  as.data.frame() |>
  na.omit()
# reduces down to 5447 introns 

ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank())+
  labs(title = "24 hours, T-025 1uM vs DMSO",
       x = expression(Delta ~ "IR ratio"))

ir_volcano_plot_ERP171667 <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  labs(title = "24 hours, T-025 1uM vs DMSO",
       x = expression(Delta ~ "IR ratio (T-025/DMSO)"))

ir_volcano_plot_ERP171667

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_T025_vs_DMSO > 0.5) |> arrange(desc(delta_mean_ir_ratio_T025_vs_DMSO)) |> dplyr::select(...1, intron_id, delta_mean_ir_ratio_T025_vs_DMSO, padj)

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_T025_vs_DMSO < -0.5) |> arrange(desc(delta_mean_ir_ratio_T025_vs_DMSO)) |> dplyr::select(...1, intron_id, delta_mean_ir_ratio_T025_vs_DMSO, padj)
```

Figure X. Volcano plot of delta mean ir ratio in cells treated with T025 (n=3) and xx cells treated with DMSO (n =3) after 24 hours. Significant intron retention events are shown in red, where significance indicates the intron had a basemean > 10, the delta IR ratio was greater than 0.1, or less than -0.1, and the p-adjust (calculated in DESeq2 when calculating Log fold change) was < 0.001.
```{r}
volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_T025_vs_DMSO > 0.5 | delta_mean_ir_ratio_T025_vs_DMSO < -0.75) |> arrange(delta_mean_ir_ratio_T025_vs_DMSO) |> dplyr::select(...1, intron_id, padj)
```


```{r}
ir_volcano_plot_ERP171667 <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  labs(x = expression(Delta ~ "IR ratio (T-025 1" * mu * "M/DMSO)"))
```


## 6.2. Zhu et al. 2018
```{r import dIR data for Zhu et al. 2018}
CC671_3uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")
CC671_3uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")

CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr|>
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM",
    "Study" = "Zhu",
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr |> mutate(
    "Hours" = 24,
    "Concentration" = "3uM",
    "Study" = "Zhu",
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr |> mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu", 
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr|> mutate(
    "Hours" = 24,
    "Concentration" = "10uM",
    "Study" = "Zhu",
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

# Filtering of delta IR
srp_combined <- read_tsv(
  "results/ir/SRP141379_shah_filtered_combined.tsv",
  show_col_types = FALSE)

srp_combined <- srp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )

# ---- distinct introns per gene ---------------------------------------------
srp_combined |>
  distinct(EnsemblID, IntronID) |>
  count(EnsemblID, name = "n_distinct_introns") |>
  arrange(desc(n_distinct_introns))

#Quality control data
srp_combined <- srp_combined |> 
  mutate(EnsemblID = str_extract(`Name`, "ENSG[0-9]+"))
#sanity check
dim(srp_combined)
head(srp_combined)

delta_ir_srp_zhu <- read_tsv(
  "results/ir/SRP141379_delta_ir_by_timepoint.tsv",
  show_col_types = FALSE)
```

### Intron retention burden
Aim: Number of intron retention events observed per sample.

```{r}
mean_Zhu <- Zhu |> group_by(exp_design, Hours, conc_uM) |> 
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop")

# Define labels for the facet
hour_labels <- c(
  "6"  = "6 hours",
  "24" = "24 hours")

IR_burden_SRP1141379 <- ggplot() +
  geom_col(data = mean_Zhu, aes(x = conc_uM, y = mean_dir),
           fill = "#9E9E9E", colour = "black") +
  geom_jitter(data = Zhu, aes(x = conc_uM, y = sample_introns, shape = exp_design), width = 0.2) +
  geom_errorbar(
    data = mean_Zhu,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), 
    width = 0.1) +
  facet_wrap(
    ~ Hours, 
    labeller = labeller(Hours = hour_labels)
  ) +
  theme_bw() +
  labs(
    y = "Intron retention events", 
    x = expression("CC-671 Concentration (" * mu * "M)")) +
  scale_y_continuous(limits = c(0, 5000))+
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = c(0.98, 0.99),
legend.justification = c("right", "top"))
  
# ggsave(filename = "SRP1141379_IR_burden.PNG",
#          plot = IR_burden_SRP1141379,
#     units = "mm",width = 200,
#          height = 180)
```

### Determine whether time or concentration drive difference in intron retention between control and treatment
```{r}
library(glmmTMB)
# Treat Hours and Concentration as factors if you only have a few discrete levels 
# (e.g. 0h/6h/24h, or Low/Med/High). Keep as numeric only if truly continuous/dose-response.
Zhu$Hours <- factor(Zhu$Hours)
Zhu$Concentration <- factor(Zhu$Concentration)
Zhu$group <- interaction(Zhu$exp_design, Zhu$Concentration, drop = TRUE)
# drop=TRUE removes empty combinations like Treatment.0uM

table(Zhu$group, Zhu$Hours)

model <- glmmTMB(
  sample_introns ~ group * Hours,
  family = nbinom2,
  data = Zhu)

summary(model)
drop1(model, test = "Chisq")

library(emmeans)

# --- Estimated marginal means (the "estimate" values) ---
emm_group_by_hours <- emmeans(model, ~ group | Hours, type = "response")
emm_hours_by_group <- emmeans(model, ~ Hours | group, type = "response")

# Pull them out as data frames so you can save/export/plot
emm_group_by_hours_df <- as.data.frame(emm_group_by_hours)
emm_hours_by_group_df <- as.data.frame(emm_hours_by_group)

emm_group_by_hours_df
emm_hours_by_group_df

# --- Reverse pairwise contrasts ---
# Use revpairwise instead of pairwise, or set reverse = TRUE in contrast()
rev_group_by_hours <- emmeans(model, revpairwise ~ group | Hours, type = "response")
rev_hours_by_group <- emmeans(model, revpairwise ~ Hours | group, type = "response")

rev_group_by_hours
rev_hours_by_group

# As data frames
rev_group_by_hours_df <- as.data.frame(rev_group_by_hours$contrasts)
rev_hours_by_group_df <- as.data.frame(rev_hours_by_group$contrasts)

```

### Scatterplots of IR ratio

#### CC671 3uM @ 6 hours
```{r}
dIR_6hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)
# 6 hours, 3uM
delta_zhu_6hr_3uM_filter <- delta_ir_srp_zhu |> 
  filter(hours == 6,
         delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #8786 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #1648 introns

zhu_mean_6hr_3uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_6hr_3uM_filter$IntronID, hours == 6) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

zhu_wide_6hr_3uM <- zhu_mean_6hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio)

sig_dIR_zhu_6hr_3um <- dIR_6hrs_CC671_3uM_vs_DMSO |> 
  filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_zhu_6hr_3uM_filter$IntronID)

zhu_wide_6hr_3uM_sig <- zhu_wide_6hr_3uM |> filter(IntronID %in% sig_dIR_zhu_6hr_3um$intron_id)

ir_ratio_zhu.1 <- ggplot() +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = zhu_wide_6hr_3uM, aes(x = CC671_3uM, y = DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("mean IR ratio (CC-671 3" * mu * "M)"),
    y = "mean IR ratio (DMSO)")

ir_ratio_zhu.1

ir_ratio_zhu.1_sig <- ggplot() +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = zhu_wide_6hr_3uM_sig, aes(x = CC671_3uM, y = DMSO), colour = "#701705", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("mean IR ratio (CC-671 3" * mu * "M)"),
    y = "mean IR ratio (DMSO)")

ir_ratio_zhu.1_sig
```

```{r}
library(dplyr)

# All points (zhu_wide_6hr_3uM)
zhu_wide_6hr_3uM <- zhu_wide_6hr_3uM %>%
  mutate(
    position = case_when(
      DMSO > CC671_3uM + 0.1 ~ "above",
      DMSO < CC671_3uM - 0.1 ~ "below",
      TRUE ~ "within"    ))

zhu_wide_6hr_3uM %>% count(position)
1127/521

# Significant subset (zhu_wide_6hr_3uM_sig)
zhu_wide_6hr_3uM_sig <- zhu_wide_6hr_3uM_sig %>%
  mutate(
    position = case_when(
      DMSO > CC671_3uM + 0.1 ~ "above",
      DMSO < CC671_3uM - 0.1 ~ "below",
      TRUE ~ "within"))

zhu_wide_6hr_3uM_sig %>% count(position)
749/284
```


#### CC671 10uM @ 6 hours
```{r}
dIR_6hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)
# 6 hours, 10uM
delta_zhu_6hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 6, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 8548 introns
                                                          delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) #2215 introns

zhu_mean_6hr_10uM <- zhu_mean_6hr_3uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_6hr_10uM_filter$IntronID, hours == 6) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

zhu_wide_6hr_10uM <- zhu_mean_6hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio  )

sig_dIR_zhu_6hr_10um <- dIR_6hrs_CC671_10uM_vs_DMSO |> 
  filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_zhu_6hr_10uM_filter$IntronID)

zhu_wide_6hr_10uM_sig <- zhu_wide_6hr_10uM |> filter(IntronID %in% sig_dIR_zhu_6hr_10um$intron_id)


zhu_mean_6hr_10uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_6hr_10uM_filter$IntronID, hours == 6) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

ir_ratio_zhu.2 <- ggplot() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank()) +
  geom_point(data = zhu_wide_6hr_10uM, aes(x = CC671_10uM, y = DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1) +
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1) +
  labs(
    x = expression("mean IR ratio (CC-671 10" * mu * "M)"),
    y = "mean IR ratio (DMSO)")

ir_ratio_zhu.2


ir_ratio_zhu.2_sig <- ggplot() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank()) +
  geom_point(data = zhu_wide_6hr_10uM_sig, aes(x = CC671_10uM, y = DMSO), colour = "#701705", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1) +
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1) +
  labs(
    x = expression("mean IR ratio (CC-671 10" * mu * "M)"),
    y = "mean IR ratio (DMSO)")

ir_ratio_zhu.2_sig
```

```{r}
library(dplyr)

# All points (zhu_wide_6hr_3uM)
zhu_wide_6hr_10uM <- zhu_wide_6hr_10uM %>%
  mutate(
    position = case_when(
      DMSO > CC671_10uM + 0.1 ~ "above",
      DMSO < CC671_10uM - 0.1 ~ "below",
      TRUE ~ "within"))

zhu_wide_6hr_10uM %>% count(position)
1706/509

# Significant subset (zhu_wide_6hr_3uM_sig)
zhu_wide_6hr_10uM_sig <- zhu_wide_6hr_10uM_sig %>%
  mutate(
    position = case_when(
      DMSO > CC671_10uM + 0.1 ~ "above",
      DMSO < CC671_10uM - 0.1 ~ "below",
      TRUE ~ "within"))

zhu_wide_6hr_10uM_sig %>% count(position)
1198/279
```

#### CC671 3uM @ 24 hours

```{r}
dIR_24hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)
# 24 hours, 3uM
delta_zhu_24hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 24,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #9302 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #948 introns 

zhu_mean_24hr_3uM <- zhu_mean_24hr_3uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_24hr_3uM_filter$IntronID, hours == 24) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

zhu_wide_24hr_3uM <- zhu_mean_24hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

sig_dIR_zhu_24hr_3um <- dIR_24hrs_CC671_3uM_vs_DMSO |> 
  filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_zhu_24hr_3uM_filter$IntronID)

zhu_wide_24hr_3uM_sig <- zhu_wide_24hr_3uM |> filter(IntronID %in% sig_dIR_zhu_24hr_3um$intron_id)

ir_ratio_zhu.3 <- ggplot() +
  geom_point(data = zhu_wide_24hr_3uM, aes(x = CC671_3uM, y = DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("mean IR ratio (CC-671 3" * mu * "M)"),
    y = "mean IR ratio (DMSO)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())

ir_ratio_zhu.3

ir_ratio_zhu.3_sig <- ggplot() +
  geom_point(data = zhu_wide_24hr_3uM_sig, aes(x = CC671_3uM, y = DMSO), colour = "#701705", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("mean IR ratio (CC-671 3" * mu * "M)"),
    y = "mean IR ratio (DMSO)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())

ir_ratio_zhu.3_sig
```

```{r}
library(dplyr)

# All points (zhu_wide_6hr_3uM)
zhu_wide_24hr_3uM <- zhu_wide_24hr_3uM %>%
  mutate(
    position = case_when(
      DMSO > CC671_3uM + 0.1 ~ "above",
      DMSO < CC671_3uM - 0.1 ~ "below",
      TRUE ~ "within"
    )
  )

zhu_wide_24hr_3uM %>% count(position)
499/449

# Significant subset (zhu_wide_6hr_3uM_sig)
zhu_wide_24hr_3uM_sig <- zhu_wide_24hr_3uM_sig %>%
  mutate(
    position = case_when(
      DMSO > CC671_3uM + 0.1 ~ "above",
      DMSO < CC671_3uM - 0.1 ~ "below",
      TRUE ~ "within"
    )
  )

zhu_wide_24hr_3uM_sig %>% count(position)
331/234
```

#### CC671 10uM @ 24 hours

```{r}
dIR_24hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)
# 24 hours, 10um
delta_zhu_24hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 24, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 9288 differentially expressed
                                                 delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) # 1069 introns         

zhu_mean_24hr_10uM <- zhu_mean_24hr_10uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_24hr_10uM_filter$IntronID, hours == 24) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

# ---- 3. Now pivot_wider works, because each cell has one value ------------
zhu_wide_24hr_10uM <- zhu_mean_24hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio)

sig_dIR_zhu_24hr_10um <- dIR_24hrs_CC671_10uM_vs_DMSO |> 
  filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_zhu_24hr_10uM_filter$IntronID)

zhu_wide_24hr_10uM_sig <- zhu_wide_24hr_10uM |> filter(IntronID %in% sig_dIR_zhu_24hr_10um$intron_id)

ir_ratio_zhu.4 <- ggplot() +
  geom_point(data = zhu_wide_24hr_10uM, aes(x = CC671_10uM, y = DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("mean IR ratio (CC-671 10" * mu * "M)"),
    y = "mean IR ratio (DMSO)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())

ir_ratio_zhu.4

ir_ratio_zhu.4_sig <- ggplot() +
  geom_point(data = zhu_wide_24hr_10uM_sig, aes(x = CC671_10uM, y = DMSO), colour = "#701705", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("mean IR ratio (CC-671 10" * mu * "M)"),
    y = "mean IR ratio (DMSO)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank()) 

ir_ratio_zhu.4_sig
```

```{r}
library(dplyr)

# All points (zhu_wide_6hr_3uM)
zhu_wide_24hr_10uM <- zhu_wide_24hr_10uM %>%
  mutate(
    position = case_when(
      DMSO > CC671_10uM + 0.1 ~ "above",
      DMSO < CC671_10uM - 0.1 ~ "below",
      TRUE ~ "within"))

zhu_wide_24hr_10uM %>% count(position)
637/432

# Significant subset (zhu_wide_6hr_3uM_sig)
zhu_wide_24hr_10uM_sig <- zhu_wide_24hr_10uM_sig %>%
  mutate(
    position = case_when(
      DMSO > CC671_10uM + 0.1 ~ "above",
      DMSO < CC671_10uM - 0.1 ~ "below",
      TRUE ~ "within"))

zhu_wide_24hr_10uM_sig %>% count(position)
442/240
```



### Volcano plot
#### CC671 3uM @ 6 hours
```{r}
dIR_6hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_6hr_3uM <- delta_ir_srp_zhu |> filter(hours == 6)

volcano_zhu_6hr_3uM <- left_join(
  dIR_6hrs_CC671_3uM_vs_DMSO,
  delta_ir_6hr_3uM,
  join_by(intron_id == IntronID)) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> 
  filter(baseMean > 10)

volc_data <- volcano_zhu_6hr_3uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

volc_data |> group_by(significance) |> summarise(n_introns = n())

ir_volcano_plot_SRP141379_3uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO,
                                                           y = -log10(padj),
                                                           colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank())+
  labs( x = expression(Delta ~ "IR ratio (CC-671 3" * mu * "M/DMSO)"))
ir_volcano_plot_SRP141379_3uM_6hr

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO < -0.5) |> arrange(desc(delta_mean_ir_ratio_CC671_3uM_vs_DMSO)) |> dplyr::select(intron_id, delta_mean_ir_ratio_CC671_3uM_vs_DMSO,padj)
```

```{r}
volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO < -0.5) |> arrange((delta_mean_ir_ratio_CC671_3uM_vs_DMSO)) |> dplyr::select(...1, intron_id, padj)
```


```{r}
ir_l2fc_plot_SRP141379_3uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = log2FoldChange, colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank())+
   geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  labs(title = "6 hours",
       x = expression(Delta ~ "IR ratio (CC-671 3" * mu * "M/DMSO)")) +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())
```


#### CC671 10uM @ 6 hours
```{r}
dIR_6hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_6hr_10uM <- delta_ir_srp_zhu |> filter(hours == 6)

volcano_zhu_6hr_10uM <- left_join(
  dIR_6hrs_CC671_10uM_vs_DMSO,
  delta_ir_6hr_10uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> 
  filter(baseMean > 10)

volc_data <- volcano_zhu_6hr_10uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

ir_volcano_plot_SRP141379_10uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO,
                                                            y = -log10(padj), 
                                                            colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank())+
  labs(
    #title = "6 hours",
       x = expression(Delta ~ "IR ratio (CC671 10" * mu * "M/DMSO)"))

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_10uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO < -0.5) |> arrange(desc(delta_mean_ir_ratio_CC671_10uM_vs_DMSO)) |> dplyr::select(delta_mean_ir_ratio_CC671_10uM_vs_DMSO, intron_id, padj)
```

```{r}
volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_10uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO < -0.5) |> arrange(delta_mean_ir_ratio_CC671_10uM_vs_DMSO) |> dplyr::select(...1, intron_id, padj)
```

#### CC671 3uM @ 24 hours

```{r}
dIR_24hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)
delta_ir_24hr_3uM <- delta_ir_srp_zhu |> filter(hours == 24)

volcano_zhu_24hr_3uM <- left_join(
  dIR_24hrs_CC671_3uM_vs_DMSO,
  delta_ir_24hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> 
  filter(baseMean > 10)

volc_data <- volcano_zhu_24hr_3uM |>
  filter(baseMean > 10) |> #9332 introns
  as.data.frame() |>
  na.omit()

volc_data |> group_by(significance) |> summarise(n_introns = n())


ir_volcano_plot_SRP141379_3uM_24hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO,
                                                            y = -log10(padj),
                                                            colour = significance)) +
  geom_point(alpha = 0.6, size = 0.7) +
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank())+
  labs(x= expression(Delta ~ "IR ratio (CC671 3" * mu * "M/DMSO)"))

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO < -0.5) |> arrange(desc(delta_mean_ir_ratio_CC671_3uM_vs_DMSO)) |> dplyr::select(...1, intron_id, padj)

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO < -0.5) |> arrange(delta_mean_ir_ratio_CC671_3uM_vs_DMSO) |> dplyr::select(delta_mean_ir_ratio_CC671_3uM_vs_DMSO, intron_id, padj)

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0.3 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO < -0.5) |> arrange(delta_mean_ir_ratio_CC671_3uM_vs_DMSO) |> dplyr::select(delta_mean_ir_ratio_CC671_3uM_vs_DMSO, intron_id, padj)
```

9332 introns have base mean > 10


#### CC671 10uM @ 24 hours

```{r}
dIR_24hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_24hr_10uM <- delta_ir_srp_zhu |> filter(hours == 24, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA")

volcano_zhu_24hr_10uM <- left_join(
  dIR_24hrs_CC671_10uM_vs_DMSO,
  delta_ir_24hr_10uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> 
  filter(baseMean > 10)

volc_data <- volcano_zhu_24hr_10uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

ir_volcano_plot_SRP141379_10uM_24hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO,
                                                             y = -log10(padj),
                                                             colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  theme(panel.grid = element_blank())+
  labs(x = expression(Delta ~ "IR ratio (CC-671 10" * mu * "M/DMSO)"))

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_10uM_vs_DMSO > 0.5 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO < -0.5) |> arrange(desc(delta_mean_ir_ratio_CC671_10uM_vs_DMSO)) |> dplyr::select(...1, intron_id, padj)

volc_data |> filter(significance == "#701705") |> filter(
  delta_mean_ir_ratio_CC671_10uM_vs_DMSO > 0.4 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO < -0.5) |> arrange(delta_mean_ir_ratio_CC671_10uM_vs_DMSO)|> dplyr::select(delta_mean_ir_ratio_CC671_10uM_vs_DMSO, intron_id, padj)
```

9332 introns

```{r}
ir_volcano_plot_full_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_6hr, ir_volcano_plot_SRP141379_10uM_6hr, ir_volcano_plot_SRP141379_3uM_24hr, ir_volcano_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("A", "B", "C", "D"))

# ggsave(ir_volcano_plot_full_SRP141379,
#        filename = "ir_volc_SRP141379.pdf",
#        height = 180,
#        width = 180,
#        units = "mm")
```

```{r}
ir_volcano_plot_full_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_6hr, ir_volcano_plot_SRP141379_3uM_24hr, ir_volcano_plot_SRP141379_10uM_6hr, ir_volcano_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("B", "C", "", ""))

# ir_l2fc_plot_full_SRP141379 <- ggpubr::ggarrange(ir_l2fc_plot_SRP141379_3uM_6hr, ir_l2fc_plot_SRP141379_3uM_24hr, ir_l2fc_plot_SRP141379_10uM_6hr, ir_l2fc_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("D", "E", "", ""))

# SRP141379_ir_burden_volcano <- ggarrange(IR_burden_SRP1141379, ir_volcano_plot_full_SRP141379, ir_l2fc_plot_full_SRP141379, ncol = 1, heights = c(1, 1.2, 1.2), labels = c("A", "", ""))

# ggsave(SRP141379_ir_burden_volcano,
#        filename = "SRP141379_ir_burden_volcano.png",
#        height = 280,
#        width = 180,
#        units = "mm")
```

## 6.3. Iwai et al. 2018

### Intron retention burden

```{r}
mean_iwai <- Iwai |>
  group_by(exp_design) |>
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop")

ggplot() +
  geom_point(data = mean_iwai, aes(x = exp_design, y = mean_dir), size = 2) +
  geom_errorbar(
    data = mean_iwai,
    aes(x = exp_design, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), width = 0.2)+
  geom_jitter(data = Iwai, aes(x = exp_design, y = sample_introns, colour = Concentration), width = 0.1, height = 0) +
  theme_bw()+
  labs(y = "Number of intron retention events", colour = "")+
  theme(panel.grid.major.x = element_blank())


iwai_ir_burden <- ggplot() +
  geom_col(data = Iwai, aes(x = conc_uM, y = sample_introns),
           fill = "#9E9E9E", colour = "black") +
  theme_bw() +
  labs(
    y = "Intron retention events", 
    x = expression("T-025 Concentration (" * mu * "M)")
  ) +
  scale_y_continuous(limits = c(0, 8250))+
  theme(panel.grid.major.x = element_blank(),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 11))

```
### IR ratio
```{r}
delta_ir_srp_iwai <- read_tsv(
  "results/ir/SRP112607_delta_ir.tsv",
  show_col_types = FALSE)

delta_ir_srp_iwai <- delta_ir_srp_iwai |> mutate(
    EnsemblID = str_extract(IntronID, "ENSG[0-9]+"))

# 30nM
delta_iwai_30nM_filter <- delta_ir_srp_iwai |> filter(                                                    delta_mean_ir_ratio_30nM_vs_DMSO != "NA" &                                                       delta_mean_ir_ratio_30nM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_30nM_vs_DMSO <= -0.1) #1173 introns


ir_ratio_iwai_30 <- ggplot() +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = delta_iwai_30nM_filter, aes(x = mean_ir_ratio_30nM, y = mean_ir_ratio_DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("IR ratio (T-025 0.03" * mu * "M)"),
    y = "IR ratio (DMSO)")

ir_ratio_iwai_30

delta_iwai_30nM_count <- delta_iwai_30nM_filter %>%
  mutate(
    position = case_when(
      mean_ir_ratio_DMSO > mean_ir_ratio_30nM + 0.1 ~ "above",
      mean_ir_ratio_DMSO < mean_ir_ratio_30nM - 0.1 ~ "below",
      TRUE ~ "within"    ))
delta_iwai_30nM_count %>% count(position)
800/366 # 2.19

# 100nM
delta_iwai_100nM_filter <- delta_ir_srp_iwai |> filter(                                                    delta_mean_ir_ratio_100nM_vs_DMSO != "NA" &                                                       delta_mean_ir_ratio_100nM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_100nM_vs_DMSO <= -0.1) #3439 introns


ir_ratio_iwai_100 <- ggplot() +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = delta_iwai_100nM_filter, aes(x = mean_ir_ratio_100nM, y = mean_ir_ratio_DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("IR ratio (T-025 0.1" * mu * "M)"),
    y = "IR ratio (DMSO)")

ir_ratio_iwai_100 

delta_iwai_100nM_count <- delta_iwai_100nM_filter %>%
  mutate(
    position = case_when(
      mean_ir_ratio_DMSO > mean_ir_ratio_100nM + 0.1 ~ "above",
      mean_ir_ratio_DMSO < mean_ir_ratio_100nM - 0.1 ~ "below",
      TRUE ~ "within"    ))
delta_iwai_100nM_count %>% count(position)
2976/456 #6.53

# 300nM
delta_iwai_300nM_filter <- delta_ir_srp_iwai |> filter(                                                    delta_mean_ir_ratio_300nM_vs_DMSO != "NA" &                                                       delta_mean_ir_ratio_300nM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_300nM_vs_DMSO <= -0.1) #6423 introns


ir_ratio_iwai_300 <- ggplot() +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = delta_iwai_300nM_filter, aes(x = mean_ir_ratio_300nM, y = mean_ir_ratio_DMSO), colour = "black", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = expression("IR ratio (T-025 0.3" * mu * "M)"),
    y = "IR ratio (DMSO)")

ir_ratio_iwai_300 

delta_iwai_300nM_count <- delta_iwai_300nM_filter %>%
  mutate(
    position = case_when(
      mean_ir_ratio_DMSO > mean_ir_ratio_300nM + 0.1 ~ "above",
      mean_ir_ratio_DMSO < mean_ir_ratio_300nM - 0.1 ~ "below",
      TRUE ~ "within"    ))
delta_iwai_300nM_count %>% count(position)
6012/402 # 14.96


iwai_ir_ratio <- ggarrange(ir_ratio_iwai_30, ir_ratio_iwai_100 ,ir_ratio_iwai_300, ncol = 1, labels = c("B", "C", "D"))

iwai_full_plot <- ggarrange(iwai_ir_burden, iwai_ir_ratio, ncol = 2, widths = c(1, 0.8), labels = c("A", ""))
iwai_full_plot

ggsave(filename = "SRP112607_ir_plot.PDF",
       plot = iwai_full_plot,
       width = 180,
       height = 180,
       unit = "mm")
```



# 7. DEG

## 7.1. Velde et al. 2026 (DMSO vs T025)
```{r}
dEG_T025_vs_DMSO <- read_csv("results/genes/ERP171667/T025_vs_DMSO.csv") 
dEG_T025_vs_DMSO <- dEG_T025_vs_DMSO |> mutate("EnsemblID" = ...1)

library(AnnotationDbi)
library(org.Hs.eg.db)

# Strip version suffixes if present (e.g. ENSG00000141510.15 -> ENSG00000141510)
dEG_T025_vs_DMSO <- dEG_T025_vs_DMSO |>
  mutate(EnsemblID_clean = sub("\\..*", "", EnsemblID))

dEG_T025_vs_DMSO$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = dEG_T025_vs_DMSO$EnsemblID_clean,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

dEG_T025_vs_DMSO <- dEG_T025_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      (log2FoldChange >= 2 | log2FoldChange <= -2)  ~ "#701705",
      TRUE                       ~ "#a3a8ac"
    ))

dEG_T025_vs_DMSO |> group_by(significance) |> summarise(n_genes = n())

volc_data <- dEG_T025_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

deg_ERP171667_plot <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  geom_vline(xintercept = -2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  theme_bw()+
  labs(x = expression("log2FoldChange (T-025 1" * mu * "M/DMSO)"))+
  theme(panel.grid = element_blank())
```

### Upregulated?

```{r}
dEG_T025_vs_DMSO |> 
  filter(log2FoldChange >= 2,
         padj < 0.01) |> 
  arrange(desc(log2FoldChange)) |> 
  dplyr::select(log2FoldChange, SYMBOL)

#3130 genes of 17 724 are have increased expression in T025 compared to DMSO
```

2151 genes of 17 724 are have increased expression in T025 compared to DMSO


### Upregualted genes the same genes with increased IR?

Are the genes upregulated in expression the same genes from which the isoforms have increasing IR ratio
```{r}
dEG_ERP171667_upreg_filter <- dEG_T025_vs_DMSO |> 
  filter(log2FoldChange >= 2,
         padj <= 0.01) |> 
  pull(EnsemblID)

volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac"
    ))

volcano_velde |> filter(significance != "#a3a8ac") |>
  #1988 introns
  filter(EnsemblID.x %in% dEG_ERP171667_upreg_filter) # 64 genes that have upregulated introns have the genes differentially expressed

ERP171667_upreg_genes_and_increased_ir <- volcano_velde |> filter(significance != "#a3a8ac") |>
  #1988 introns
  filter(EnsemblID.x %in% dEG_ERP171667_upreg_filter) 
```

### Intron retention events that are significant but the expression of the gene is not upregulated (dEG) (where upregulated is defined as a log2FoldChange of >= 3)
```{r}
volcano_velde |> filter(significance != "#a3a8ac") |>
  #1988 introns
  filter(!EnsemblID.x %in% dEG_ERP171667_upreg_filter) #1924 introns 
```


## 7.2. Zhu et al. 2018

### CC671 3uM @ 6 hours
```{r}
dEG_6hr_CC671_3uM_vs_DMSO <- read_csv("results/genes/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")

dEG_6hr_CC671_3uM_vs_DMSO <- dEG_6hr_CC671_3uM_vs_DMSO |> mutate("EnsemblID" = ...1)
dEG_6hr_CC671_3uM_vs_DMSO <- dEG_6hr_CC671_3uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.001 ~ "red",
      (log2FoldChange >= 2 | log2FoldChange <= -2)  ~ "#701705",
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    ))
library(AnnotationDbi)
library(org.Hs.eg.db)

# Strip version suffixes if present (e.g. ENSG00000141510.15 -> ENSG00000141510)
dEG_6hr_CC671_3uM_vs_DMSO <- dEG_6hr_CC671_3uM_vs_DMSO |>
  mutate(EnsemblID_clean = sub("\\..*", "", EnsemblID))

dEG_6hr_CC671_3uM_vs_DMSO$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = dEG_6hr_CC671_3uM_vs_DMSO$EnsemblID_clean,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

volc_data <- dEG_6hr_CC671_3uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(x = "log2FoldChange (CC671 3uM/DMSO)",
       title = "6 hours")

volcano_zhu_6hr_3uM_gene <-ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() + 
  geom_vline(xintercept = -2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  theme_bw()+
  labs(
    x = expression("log2FoldChange (CC-671 3" * mu * "M/DMSO)"),
       title = "6 hours") +
  theme(panel.grid= element_blank())+
  scale_x_continuous(limits = c(-5, 11))
```

#### Upregulated genes
```{r}
dEG_6hr_CC671_3uM_vs_DMSO |> 
  filter(log2FoldChange >= 2,
         padj <= 0.05) |> 
  arrange(desc(log2FoldChange))

#351 genes of 18114 are have increased expression in cells treated with CC671 compared to DMSO
```

#### Increased gene expression == increased intron retention?
```{r}
# SRP141379_3uM_6hr_upreg_filter <- dEG_6hr_CC671_3uM_vs_DMSO |> 
#   filter(log2FoldChange >= 3,
#          padj <= 0.05)|> 
#   pull(EnsemblID)
# # 351 introns
# 
# SRP141379_3uM_6hr_upreg_genes_and_increased_ir <-  volcano_zhu_6hr_3uM |>
#   filter(baseMean > 10) |> filter(significance != "#a3a8ac") |>
#   #1079 introns
# filter(EnsemblID %in% SRP141379_3uM_6hr_upreg_filter) 
# 19 introns with significant delta IR, and log2Fold change were derived from genes that had upregulated gene expression in treated cells
```


### CC671 10uM @ 6 hours
```{r}
dEG_6hr_CC671_10uM_vs_DMSO <- read_csv("results/genes/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
dEG_6hr_CC671_10uM_vs_DMSO <- dEG_6hr_CC671_10uM_vs_DMSO |> mutate("EnsemblID" = ...1)
dEG_6hr_CC671_10uM_vs_DMSO <- dEG_6hr_CC671_10uM_vs_DMSO |>
  mutate(EnsemblID_clean = sub("\\..*", "", EnsemblID))

dEG_6hr_CC671_10uM_vs_DMSO$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = dEG_6hr_CC671_10uM_vs_DMSO$EnsemblID_clean,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

dEG_6hr_CC671_10uM_vs_DMSO <- dEG_6hr_CC671_10uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      (log2FoldChange >= 2 | log2FoldChange <= -2)  ~ "#701705",
      TRUE                       ~ "#a3a8ac"
    ))

volc_data <- dEG_6hr_CC671_10uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

volcano_zhu_6hr_10uM_gene <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  geom_vline(xintercept = -2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  theme_bw()+
  labs(
     x = expression("log2FoldChange (CC-671 10" * mu * "M/DMSO)"))+
  theme(panel.grid = element_blank())+
  scale_x_continuous(limits = c(-5, 11))
```

#### Upregulated genes
```{r}
dEG_6hr_CC671_10uM_vs_DMSO |> 
  filter(log2FoldChange >= 2,
         padj <= 0.01) |> 
  arrange(desc(log2FoldChange))
```


### CC671 3uM @ 24 hours
```{r}
dEG_24hr_CC671_3uM_vs_DMSO <- read_csv("results/genes/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
dEG_24hr_CC671_3uM_vs_DMSO <- dEG_24hr_CC671_3uM_vs_DMSO |> mutate("EnsemblID" = ...1)
dEG_24hr_CC671_3uM_vs_DMSO <- dEG_24hr_CC671_3uM_vs_DMSO |>
  mutate(EnsemblID_clean = sub("\\..*", "", EnsemblID))

dEG_24hr_CC671_3uM_vs_DMSO$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = dEG_24hr_CC671_3uM_vs_DMSO$EnsemblID_clean,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
# Volcano Plot
dEG_24hr_CC671_3uM_vs_DMSO <- dEG_24hr_CC671_3uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      (log2FoldChange >= 2 | log2FoldChange <= -2)  ~ "#701705",
      TRUE                       ~ "#a3a8ac"
    ))

volc_data <- dEG_24hr_CC671_3uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

volcano_zhu_24hr_3uM_gene <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  geom_vline(xintercept = -2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  scale_colour_identity() +
  theme_bw()+
  labs(x = expression("log2FoldChange (CC-671 3" * mu * "M/DMSO)"),
       title = "24 hours")+
  theme(panel.grid = element_blank())+
  scale_x_continuous(limits = c(-5, 7.5))
```

#### Upregulated genes
```{r}
dEG_24hr_CC671_3uM_vs_DMSO |> 
  filter(log2FoldChange >= 2,
         padj <= 0.01) |> 
  arrange(desc(log2FoldChange))
#88 genes of 17923 are have increased expression in cells treated with CC671 compared to DMSO
```

### CC671 10uM @ 24 hours

```{r}
dEG_24hr_CC671_10uM_vs_DMSO <- read_csv("results/genes/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")
dEG_24hr_CC671_10uM_vs_DMSO <- dEG_24hr_CC671_10uM_vs_DMSO |> mutate("EnsemblID" = ...1)

dEG_24hr_CC671_10uM_vs_DMSO <- dEG_24hr_CC671_10uM_vs_DMSO |>
  mutate(EnsemblID_clean = sub("\\..*", "", EnsemblID))

dEG_24hr_CC671_10uM_vs_DMSO$SYMBOL <- mapIds(
  org.Hs.eg.db,
  keys = dEG_24hr_CC671_10uM_vs_DMSO$EnsemblID_clean,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)
dEG_24hr_CC671_10uM_vs_DMSO <- dEG_24hr_CC671_10uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      (log2FoldChange >= 2 | log2FoldChange <= -2)  ~ "#701705",
      TRUE                       ~ "#a3a8ac"))

volc_data <- dEG_24hr_CC671_10uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

volcano_zhu_24hr_10uM_gene <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  geom_vline(xintercept = -2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 2, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  theme_bw()+
  labs( x = expression("log2FoldChange (CC-671 10" * mu * "M/DMSO)"))+
  theme(panel.grid = element_blank())+
  scale_x_continuous(limits = c(-5, 7.5))
```


```{r}
dEG_24hr_CC671_10uM_vs_DMSO |> 
  filter(log2FoldChange >= 2,
         padj <= 0.01) |> 
  arrange(desc(log2FoldChange)) |> 
  dplyr::select(SYMBOL, log2FoldChange)
```


```{r}
dEG_SRP141379_volcano <- ggarrange(volcano_zhu_6hr_3uM_gene, 
                                   volcano_zhu_24hr_3uM_gene, 
                                   volcano_zhu_6hr_10uM_gene,  
                                   volcano_zhu_24hr_10uM_gene, 
                                   ncol =2, nrow = 2, 
                                   labels = c("A", "B", "", ""))
```


# 9. Cross study comparisons
## 9.1. Are intron events consistent between T-025 and CC-671 after 24 hours?
```{r}
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))
volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> filter(baseMean > 10)

sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_erp_filter$IntronID)

#1618 dIR have a base expression of 10, FDR < 0.01, and a delta IR ratio greater than or equal to |0.1|
```


```{r}
# name each Zhu dataset with its delta column, so the loop knows which column to use
zhu_datasets <- list(
  "6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  "24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)

# only increases in IR ratio
filter_hits <- function(df, delta_col, delta_thresh = 0.1) {
  df |>
    filter(
      .data[[delta_col]] >= delta_thresh,   # only increases in IR ratio
      significance != "#a3a8ac", # FDR < 0.01
      baseMean > 10 #expression sufficient for differential analysis
    ) |>
    pull(intron_id)
}


# ERP171667 hits (baseline, computed once)

velde_hits <- filter_hits(volcano_velde, "delta_mean_ir_ratio_T025_vs_DMSO")

# Loop over the four Zhu comparisons
comparison_results <- map_dfr(names(zhu_datasets), function(nm) {
  entry    <- zhu_datasets[[nm]]
  zhu_hits <- filter_hits(entry$df, entry$delta_col)
  shared   <- intersect(velde_hits, zhu_hits)

  tibble(
    comparison   = nm,
    n_velde_hits = length(velde_hits),
    n_zhu_hits   = length(zhu_hits),
    n_shared     = length(shared),
    shared_ids   = list(shared)   # list-column so you can inspect IDs later
  )
})

comparison_results
```

#### increased

```{r}
# only increases in IR ratio
filter_hits <- function(df, delta_col, delta_thresh = 0.1) {
  df |>
    filter(
      .data[[delta_col]] >= delta_thresh,   # only increases in IR ratio
      significance != "#a3a8ac", # FDR < 0.01
      baseMean > 10 #expression sufficient for differential analysis
    ) |>
    pull(intron_id)
}
zhu_datasets <- list(
"6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
"6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
"24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
"24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"))

 # Apply filter_hits across each dataset in the list
 sig_list_introns <- lapply(zhu_datasets, function(x) {
filter_hits(x$df, x$delta_col) })
 
 # Add the Velde dataset in the same way
 sig_list_introns[["T025_1uM (Velde)"]] <- filter_hits(
volcano_velde,
delta_col = "delta_mean_ir_ratio_T025_vs_DMSO")
 
 # quick sanity check
sapply(sig_list_introns, length)
```

#### both directions
```{r}
# If want both directions
filter_hits <- function(df, delta_col, delta_thresh = 0.1) {
  df |>
    filter(
      abs(.data[[delta_col]]) >= delta_thresh,
      significance != "#a3a8ac",
      baseMean > 10
    ) |>
    pull(intron_id)
}
 
zhu_datasets <- list(
"6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
"6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
"24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
"24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"))

 # Apply filter_hits across each dataset in the list
 sig_list_introns <- lapply(zhu_datasets, function(x) {
filter_hits(x$df, x$delta_col) })
 
 # Add the Velde dataset in the same way
 sig_list_introns[["T025_1uM (Velde)"]] <- filter_hits(
volcano_velde,
delta_col = "delta_mean_ir_ratio_T025_vs_DMSO")
 
 # quick sanity check
sapply(sig_list_introns, length)

```

```{r}
library(UpSetR)

all_hits <- list(
  `T-025 1uM 24hr`    = velde_hits,
  `CC-671 3uM 6hr`   = filter_hits(volcano_zhu_6hr_3uM,   "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  `CC-671 10uM 6hr`  = filter_hits(volcano_zhu_6hr_10uM,  "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  `CC-671 3uM 24hr`  = filter_hits(volcano_zhu_24hr_3uM,  "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  `CC-671 10uM 24hr` = filter_hits(volcano_zhu_24hr_10uM, "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)

# upset(fromList(all_hits), order.by = "freq")

set_sizes <- sapply(all_hits, length)
sorted_names <- names(sort(set_sizes, decreasing = TRUE))

# rebuild bar_colors in the same sorted order
color_lookup <- c(
  `T-025 1uM 24hr`    = "#005b96",
  `CC-671 3uM 6hr`    = "#ee5d6c",
  `CC-671 10uM 6hr`   = "#ee5d6c",
  `CC-671 3uM 24hr`   = "#ce4993",
  `CC-671 10uM 24hr`  = "#ce4993"
)
sorted_colors <- color_lookup[sorted_names]

dIR_significant_shared_plot <- upset(fromList(all_hits),
      order.by = "freq",
      sets = sorted_names,
      keep.order = TRUE,
      sets.bar.color = sorted_colors,
      text.scale = 1.2)
```

```{r}
library(UpSetR)

all_hits <- list(
  `T-025 1uM 24hr`    = velde_hits,
  `CC-671 3uM 6hr`   = filter_hits(volcano_zhu_6hr_3uM,   "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  `CC-671 10uM 6hr`  = filter_hits(volcano_zhu_6hr_10uM,  "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  `CC-671 3uM 24hr`  = filter_hits(volcano_zhu_24hr_3uM,  "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  `CC-671 10uM 24hr` = filter_hits(volcano_zhu_24hr_10uM, "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"))

set_sizes <- sapply(all_hits, length)
sorted_names <- names(sort(set_sizes, decreasing = TRUE))

color_lookup <- c(
  `T-025 1uM 24hr`    = "#005b96",
  `CC-671 3uM 6hr`    = "#ee5d6c",
  `CC-671 10uM 6hr`   = "#ee5d6c",
  `CC-671 3uM 24hr`   = "#ce4993",
  `CC-671 10uM 24hr`  = "#ce4993")
sorted_colors <- color_lookup[sorted_names]

# metadata for row shading — must match sorted_names order
metadata <- data.frame(
  sets  = sorted_names,
  study = ifelse(sorted_names == "T-025 1uM 24hr", "T-025", "CC-671"))

dIR_significant_shared_plot <- upset(
  fromList(all_hits),
  order.by = "freq",
  sets = sorted_names,
  keep.order = TRUE,
  sets.bar.color = sorted_colors,
  text.scale = 1.2,
  set.metadata = list(
    data = metadata,
    plots = list(
      list(type = "matrix_rows",
           column = "study",
           colors = c("T-025" = "#cccccc",
                      "CC-671" = "#a3a3a3"),
           alpha = 0.5))))

dIR_significant_shared_plot
```


```{r}
library(dplyr)
library(purrr)
library(tibble)

zhu_datasets <- list(
  "6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  "24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)

# direction = "up", "down", or "both" (both = union of up & down, not requiring same sign in comparison)
filter_hits <- function(df, delta_col, direction = "both", delta_thresh = 0.1) {
  d <- .data <- NULL  # avoid lint noise
  vals <- df[[delta_col]]

  keep <- switch(direction,
    up   = vals >= delta_thresh,
    down = vals <= -delta_thresh,
    both = vals >= delta_thresh | vals <= -delta_thresh
  )

  df |>
    filter(keep,
           significance != "#a3a8ac",
           baseMean > 10) |>
    pull(intron_id)
}

# Velde hits, split by direction
velde_up   <- filter_hits(volcano_velde, "delta_mean_ir_ratio_T025_vs_DMSO", "up")
velde_down <- filter_hits(volcano_velde, "delta_mean_ir_ratio_T025_vs_DMSO", "down")

comparison_results <- map_dfr(names(zhu_datasets), function(nm) {
  entry <- zhu_datasets[[nm]]

  zhu_up   <- filter_hits(entry$df, entry$delta_col, "up")
  zhu_down <- filter_hits(entry$df, entry$delta_col, "down")

  shared_up      <- intersect(velde_up, zhu_up)          # both increase
  shared_down    <- intersect(velde_down, zhu_down)      # both decrease
  shared_opposite_ud <- intersect(velde_up, zhu_down)     # Velde up, Zhu down
  shared_opposite_du <- intersect(velde_down, zhu_up)     # Velde down, Zhu up

  tibble(
    comparison        = nm,
    n_velde_up        = length(velde_up),
    n_velde_down      = length(velde_down),
    n_zhu_up          = length(zhu_up),
    n_zhu_down        = length(zhu_down),
    n_shared_up       = length(shared_up),
    n_shared_down     = length(shared_down),
    n_shared_opposite = length(shared_opposite_ud) + length(shared_opposite_du),
    n_shared_same_dir = length(shared_up) + length(shared_down),
    shared_up_ids      = list(shared_up),
    shared_down_ids     = list(shared_down))
})

comparison_results |>
  dplyr::select(comparison, n_velde_up, n_velde_down, n_zhu_up, n_zhu_down,
         n_shared_up, n_shared_down, n_shared_same_dir, n_shared_opposite)
```

```{r}
up_only <- list(
  `T-025 24hr 1uM` = velde_up,
  `CC-671 6hr 3uM`    = filter_hits(volcano_zhu_6hr_3uM,   "delta_mean_ir_ratio_CC671_3uM_vs_DMSO",  "up"),
  `CC-671 6hr 10uM`   = filter_hits(volcano_zhu_6hr_10uM,  "delta_mean_ir_ratio_CC671_10uM_vs_DMSO", "up"),
  `CC-671 24hr 3uM`   = filter_hits(volcano_zhu_24hr_3uM,  "delta_mean_ir_ratio_CC671_3uM_vs_DMSO",  "up"),
  `CC-671 24hr 10uM`  = filter_hits(volcano_zhu_24hr_10uM, "delta_mean_ir_ratio_CC671_10uM_vs_DMSO", "up"))

upset(fromList(up_only), order.by = "freq", nsets = 5)
```

22 introns were shared in all 5 samples.

```{r}
shared_all <- Reduce(intersect, up_only)
length(shared_all)   # should be 22
shared_all           # the intron IDs themselves

shared_all_annotated <- purrr::reduce(
  list(
    volcano_velde |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_velde   = delta_mean_ir_ratio_T025_vs_DMSO,
             padj_velde    = padj,
             baseMean_velde = baseMean,
             log2FC_velde   = log2FoldChange),

    volcano_zhu_6hr_3uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_6hr_3uM   = delta_mean_ir_ratio_CC671_3uM_vs_DMSO,
             padj_6hr_3uM    = padj,
             baseMean_6hr_3uM = baseMean,
             log2FC_6hr_3uM   = log2FoldChange),

    volcano_zhu_6hr_10uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_6hr_10uM   = delta_mean_ir_ratio_CC671_10uM_vs_DMSO,
             padj_6hr_10uM    = padj,
             baseMean_6hr_10uM = baseMean,
             log2FC_6hr_10uM   = log2FoldChange),

    volcano_zhu_24hr_3uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_24hr_3uM   = delta_mean_ir_ratio_CC671_3uM_vs_DMSO,
             padj_24hr_3uM    = padj,
             baseMean_24hr_3uM = baseMean,
             log2FC_24hr_3uM   = log2FoldChange),

    volcano_zhu_24hr_10uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_24hr_10uM   = delta_mean_ir_ratio_CC671_10uM_vs_DMSO,
             padj_24hr_10uM    = padj,
             baseMean_24hr_10uM = baseMean,
             log2FC_24hr_10uM   = log2FoldChange)
  ),
  left_join, by = "intron_id"
)

shared_all_annotated

library(org.Hs.eg.db)
library(AnnotationDbi)
library(stringr)

# Extract the Ensembl gene ID from intron_id using regex
shared_all_annotated <- shared_all_annotated |>
  mutate(ensembl_id = str_extract(intron_id, "ENSG[0-9]+"))

# Check extraction worked before mapping
sum(is.na(shared_all_annotated$ensembl_id))  # should be 0 - flag if not

# Map Ensembl ID -> gene symbol
shared_all_annotated <- shared_all_annotated |>
  mutate(gene_symbol = mapIds(
    org.Hs.eg.db,
    keys = ensembl_id,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  ))

# Move the new columns to the front for readability
shared_all_annotated <- shared_all_annotated |>
  dplyr::select(intron_id, ensembl_id, gene_symbol, everything())

shared_all_annotated

write.csv(shared_all_annotated, file = "shared_all_annotated.csv")
```

```{r}
library(tidyr)
library(dplyr)
library(stringr)

shared_all_long <- shared_all_annotated |>
  pivot_longer(
    cols = -intron_id,
    names_to = c(".value", "condition"),
    names_pattern = "^(delta|padj|baseMean|log2FC)_(.*)$")

shared_all_long
shared_all_long <- shared_all_long |>
  mutate(condition = factor(condition,
                             levels = c("velde", "6hr_3uM", "6hr_10uM", "24hr_3uM", "24hr_10uM")))
shared_all_long <- shared_all_long |>
  mutate(
    condition = recode(condition,
      "velde"      = "T-025 1uM 24hrs",
      "6hr_3uM"    = "CC-671 3uM 6hrs",
      "6hr_10uM"   = "CC-671 10uM 6hrs",
      "24hr_3uM"   = "CC-671 3uM 24hrs",
      "24hr_10uM"  = "CC-671 10uM 24hrs"),
    condition = factor(condition,
      levels = c("T-025 1uM 24hrs",
                 "CC-671 3uM 6hrs",
                 "CC-671 10uM 6hrs",
                 "CC-671 3uM 24hrs",
                 "CC-671 10uM 24hrs")))

ggplot(shared_all_long, aes(x = condition, y = intron_id, fill = delta)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
filter_hits <- function(df, delta_col, delta_thresh = 0.1) {
  df |>
    filter(
      .data[[delta_col]] >= delta_thresh,   # only increases in IR ratio
      significance != "#a3a8ac",            # FDR < 0.01
      baseMean > 10                         # expression sufficient for differential analysis
    ) |>
    pull(intron_id)
}

zhu_datasets <- list(
  "6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  "24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)

# Apply filter_hits across each dataset in the list
sig_list_introns <- lapply(zhu_datasets, function(x) {
  filter_hits(x$df, x$delta_col)
})

# Add the Velde dataset in the same way
sig_list_introns[["T025_1uM (Velde)"]] <- filter_hits(
  volcano_velde,
  delta_col = "delta_mean_ir_ratio_T025_vs_DMSO"
)

# quick sanity check
sapply(sig_list_introns, length)
```


## 9.2. Are differentially expressed genes consistent among treatments?
```{r}
library(UpSetR)
# or: library(ComplexUpset)  # if you want ggplot-styled output matching your theme_bw() plots

# Helper to build a "significant gene" vector from a DESeq2-style results table
get_sig_genes <- function(df, lfc_thresh = 2, padj_thresh = 0.01, basemean_thresh = 10) {
  df |>
    filter(baseMean > basemean_thresh) |>
    filter((log2FoldChange >= lfc_thresh | log2FoldChange <= -lfc_thresh) & padj < padj_thresh) |>
    pull(EnsemblID)
}

sig_list <- list(
  "T025_1uM (ERP171667)"    = get_sig_genes(dEG_T025_vs_DMSO),
  "CC671_3uM_6hr (SRP141379)"  = get_sig_genes(dEG_6hr_CC671_3uM_vs_DMSO),
  "CC671_10uM_6hr (SRP141379)" = get_sig_genes(dEG_6hr_CC671_10uM_vs_DMSO),
  "CC671_3uM_24hr (SRP141379)"  = get_sig_genes(dEG_24hr_CC671_3uM_vs_DMSO),
  "CC671_10uM_24hr (SRP141379)" = get_sig_genes(dEG_24hr_CC671_10uM_vs_DMSO)
)

# quick sanity check
sapply(sig_list, length)

upset(
  fromList(sig_list),
  order.by = "freq",
  nsets = length(sig_list),
  mainbar.y.label = "Intersection size (DEGs)",
  sets.x.label = "DEGs per comparison",
  text.scale = 1.3
)

all_hits <- list(
  `T-025 1uM 24hr`    = get_sig_genes(dEG_T025_vs_DMSO),
  `CC-671 3uM 6hr`   = get_sig_genes(dEG_6hr_CC671_3uM_vs_DMSO),
  `CC-671 10uM 6hr`  = get_sig_genes(dEG_6hr_CC671_10uM_vs_DMSO),
  `CC-671 3uM 24hr`  = get_sig_genes(dEG_24hr_CC671_3uM_vs_DMSO),
  `CC-671 10uM 24hr` = get_sig_genes(dEG_24hr_CC671_10uM_vs_DMSO))

set_sizes <- sapply(all_hits, length)
sorted_names <- names(sort(set_sizes, decreasing = TRUE))

color_lookup <- c(
  `T-025 1uM 24hr`    = "#005b96",
  `CC-671 3uM 6hr`    = "#ee5d6c",
  `CC-671 10uM 6hr`   = "#ee5d6c",
  `CC-671 3uM 24hr`   = "#ce4993",
  `CC-671 10uM 24hr`  = "#ce4993")
sorted_colors <- color_lookup[sorted_names]

# metadata for row shading — must match sorted_names order
metadata <- data.frame(
  sets  = sorted_names,
  study = ifelse(sorted_names == "T-025 1uM 24hr", "T-025", "CC-671"))

deg_significant_shared_plot <- upset(
  fromList(all_hits),
  order.by = "freq",
  sets = sorted_names,
  keep.order = TRUE,
  sets.bar.color = sorted_colors,
  text.scale = 1.2,
  set.metadata = list(
    data = metadata,
    plots = list(
      list(type = "matrix_rows",
           column = "study",
           colors = c("T-025" = "#cccccc",
                      "CC-671" = "#a3a3a3"),
           alpha = 0.5))))

deg_significant_shared_plot
```

```{r}
# Genes shared across ALL 5 comparisons (the "39" bar)
shared_all_deg <- Reduce(intersect, sig_list)
length(shared_all_deg)
shared_all_deg

library(org.Hs.eg.db)
library(AnnotationDbi)

# Ensembl IDs sometimes have version suffixes (e.g. ENSG00000141510.15) — 
# org.Hs.eg.db expects unversioned IDs, so strip the suffix if present
shared_all_clean_deg <- sub("\\..*$", "", shared_all_deg)

shared_all_df <- tibble(
  EnsemblID = shared_all_deg,
  EnsemblID_clean = shared_all_clean_deg,
  symbol = mapIds(org.Hs.eg.db,
                   keys = shared_all_clean_deg,
                   column = "SYMBOL",
                   keytype = "ENSEMBL",
                   multiVals = "first"),
  entrez = mapIds(org.Hs.eg.db,
                   keys = shared_all_clean_deg,
                   column = "ENTREZID",
                   keytype = "ENSEMBL",
                   multiVals = "first"),
  gene_name = mapIds(org.Hs.eg.db,
                      keys = shared_all_clean_deg,
                      column = "GENENAME",
                      keytype = "ENSEMBL",
                      multiVals = "first")
)

shared_all_df

write.csv(shared_all_df, file = "shared_all_genes.csv")

sum(is.na(shared_all_df$symbol))

shared_all_15 <- shared_all_df |> filter(symbol != "NA")
shared_all_df |> filter(is.na(symbol)) |> dplyr::select(EnsemblID)
```


# 10. Plots for report

## Liu et al. 2025 Introns

```{r}
mean_Velde <- Velde |>
  group_by(conc_uM, exp_design) |>
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()),
    .groups  = "drop"
  ) |>
  mutate(conc_uM = factor(conc_uM))

Velde <- Velde |> mutate(conc_uM = factor(conc_uM))

sig_df_Velde <- data.frame(
  xmin = "0",
  xmax = "1",
  y_position = 3700,
  annotations = "*"   # p = 0.0158
)

ERP171667_IR_burden <- ggplot() +
  geom_col(data = mean_Velde, aes(x = conc_uM, y = mean_dir), fill = "#9E9E9E", colour = "black") +
  geom_jitter(data = Velde, aes(x = conc_uM, y = sample_introns, shape = exp_design), width = 0.2) +
  geom_errorbar(
    data = mean_Velde,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), width = 0.1) +
  geom_signif(
    data = sig_df_Velde,
    aes(xmin = xmin, xmax = xmax, y_position = y_position, annotations = annotations),
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4
  ) +
  theme_bw() +
  labs(y = "Intron retention events",
       x = expression("T-025 Concentration (" * mu * "M)")#, title = "ERP171667"
       ) +
  scale_y_continuous(limits = c(0, 5250)) +
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 11),
        legend.title = element_blank(),
        legend.position = "right",
        legend.margin = margin(t = 0, b = 0),       # trims padding above/below legend
        legend.box.spacing = unit(0.2, "cm"))

ERP171667_IR_burden

ERP171667_ir_ratio_delta_ir <- ggplot() +
  geom_point(data = erp_wide, aes(x = T025, y = DMSO), alpha = 0.4, size = 0.5) +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  #geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "blue", alpha = 0.4, size = 1)+
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = "mean IR ratio - T-025",
    y = "mean IR ratio - DMSO"
  )
ERP171667_ir_ratio_delta_ir_deseq_sig <- ggplot() +
  #geom_point(data = erp_wide, aes(x = T025, y = DMSO), alpha = 0.4, size = 0.5) +
  theme_bw()+
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "#701705", alpha = 0.4, size = 0.5) +
  geom_abline(slope = 1, intercept = 0.1, linetype = "dashed", color = "black", size = 0.1)+
  geom_abline(slope = 1, intercept = -0.1, linetype = "dashed", color = "black", size = 0.1)+
  labs(
    x = "mean IR ratio - T-025",
    y = "mean IR ratio - DMSO")

# merge delta IR and the dIR data
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))

volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01  ~ "#701705",
      TRUE                       ~ "#a3a8ac")) |> 
  filter(baseMean > 10)

volcano_velde |> group_by(significance) |> summarise(n_introns = n())

volc_data <- volcano_velde |>
  filter(baseMean > 10) |> #ensures that only introns with enough expression are included
  as.data.frame() |>
  na.omit()

ir_volcano_plot_ERP171667 <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  geom_vline(xintercept = -0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_vline(xintercept = 0.1, linetype = "dashed", alpha = 0.7, colour = "black", size =0.5)+
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", alpha = 0.7, colour = "black",size = 0.5)+
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid = element_blank(),
        panel.grid.major = element_blank())+
  labs(x = expression(Delta ~ "IR ratio (T-025 1" * mu * "M/DMSO)"))

ERP171667_ir_ratio <- ggarrange(ERP171667_ir_ratio_delta_ir, ERP171667_ir_ratio_delta_ir_deseq_sig, ncol = 1, labels = c("B", "C"))

ERP_ir_plot1 <- ggarrange(ERP171667_IR_burden, ERP171667_ir_ratio, labels = c("A", ""), ncol = 2, widths = c(1, 0.8))

ERP_ir_plot <- ggarrange(ERP_ir_plot1, ir_volcano_plot_ERP171667, ncol =1, heights = c(1.5, 0.7), labels = c("", "D"))

ggsave(filename = "ERP171667_ir_plot.PNG",
       plot = ERP_ir_plot,
       width = 180,
       height = 200,
       unit = "mm")

ggsave(filename = "ERP171667_ir_plot.PDF",
       plot = ERP_ir_plot,
       width = 180,
       height = 200,
       unit = "mm")

```

## Zhu et al Introns

### 6 hours

```{r}
sig_df_6hr <- data.frame(
  xmin = c("0", "0", "3"),
  xmax = c("3", "10", "10"),
  y_position = c(4875, 5200, 4700),
  annotations = c("****", "****", "**")   # Hours=6: p<0.0001, p<0.0001, p=0.0029
)

mean_Zhu_6hr <- mean_Zhu |> filter(Hours == 6) |> mutate(conc_uM = factor(conc_uM))
Zhu_6hr       <- Zhu |> filter(Hours == 6) |> mutate(conc_uM = factor(conc_uM))

IR_burden_SRP1141379_6hr <- ggplot() +
  geom_col(data = mean_Zhu_6hr, aes(x = conc_uM, y = mean_dir),
           fill = "#9E9E9E", colour = "black") +
  geom_jitter(data = Zhu_6hr, aes(x = conc_uM, y = sample_introns, shape = exp_design), width = 0.2) +
  geom_errorbar(
    data = mean_Zhu_6hr,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), 
    width = 0.1) +
  geom_signif(
    data = sig_df_6hr,
    aes(xmin = xmin, xmax = xmax, y_position = y_position, annotations = annotations),
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4
  ) +
  theme_bw() +
  labs(
    title = "6 hours",
    y = "Intron retention events", 
    x = expression("CC-671 Concentration (" * mu * "M)")) +
  scale_y_continuous(limits = c(0, 5250)) +
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 11),
        legend.title = element_blank(),
        legend.position = "right",
        legend.margin = margin(t = 0, b = 0),       # trims padding above/below legend
        legend.box.spacing = unit(0.2, "cm"))       # gap between plot panel and legend)

IR_burden_SRP1141379_6hr
```

```{r}
ir_ratio_zhu_6hours <- ggarrange(ir_ratio_zhu.1, ir_ratio_zhu.2, ir_ratio_zhu.1_sig,  ir_ratio_zhu.2_sig, ncol = 2, nrow = 2, labels =c("B", "E", "C", "F"))

ir_ratio_zhu_6hours

ir_volcano_plot_6hr_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_6hr, ir_volcano_plot_SRP141379_10uM_6hr, ncol = 2, labels = c("D", "G"))

Zhu_6hours_ir <- ggarrange(IR_burden_SRP1141379_6hr, ir_ratio_zhu_6hours, ir_volcano_plot_6hr_SRP141379,  ncol = 1, heights = c(1.25, 1.5, 0.7), labels = c("A", ""))
Zhu_6hours_ir

# ggsave(Zhu_6hours_ir,
#        filename = "Zhu_6hours_ir.PDF",
#        height = 280,
#        width = 180,
#        units = "mm")
```

```{r}
ir_volcano_plot_full_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_6hr, ir_volcano_plot_SRP141379_3uM_24hr, ir_volcano_plot_SRP141379_10uM_6hr, ir_volcano_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("B", "C", "", ""))

# SRP141379_ir_burden_volcano <- ggarrange(IR_burden_SRP1141379, ir_volcano_plot_full_SRP141379, ir_l2fc_plot_full_SRP141379, ncol = 1, heights = c(1, 1.2, 1.2), labels = c("A", "", ""))
# 
# ggsave(SRP141379_ir_burden_volcano,
#        filename = "SRP141379_ir_burden_volcano.png",
#        height = 280,
#        width = 180,
#        units = "mm")
```

### 24 hours
```{r}
hour_labels <- c(
  "6"  = "6 hours",
  "24" = "24 hours")

mean_Zhu_24hr <- mean_Zhu |> filter(Hours == 24) |> mutate(conc_uM = factor(conc_uM))
Zhu_24hr       <- Zhu |> filter(Hours == 24) |> mutate(conc_uM = factor(conc_uM))

sig_df_24hr <- data.frame(
  xmin = c("0", "0", "3"),
  xmax = c("3", "10", "10"),
  y_position = c(4200, 4600, 3900),
  annotations = c("**", "***", "ns"))

IR_burden_SRP1141379_24hr <- ggplot() +
  geom_col(data = mean_Zhu_24hr, aes(x = conc_uM, y = mean_dir),
           fill = "#9E9E9E", colour = "black") +
  geom_jitter(data = Zhu_24hr, aes(x = conc_uM, y = sample_introns, shape = exp_design), width = 0.2) +
  geom_errorbar(
    data = mean_Zhu_24hr,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), 
    width = 0.1) +
  geom_signif(data = sig_df_24hr,
    aes(xmin = xmin, xmax = xmax, y_position = y_position, annotations = annotations),
    manual = TRUE,
    tip_length = 0.01,
    textsize = 4) +
  theme_bw() +
  labs(
    title = "24 hours",
    y = "Intron retention events", 
    x = expression("CC-671 Concentration (" * mu * "M)")) +
  scale_y_continuous(limits = c(0, 5250)) +
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 11),
        legend.title = element_blank(),
        legend.position = "right",
        legend.margin = margin(t = 0, b = 0),       # trims padding above/below legend
        legend.box.spacing = unit(0.2, "cm"))       # gap between plot panel and legend)

IR_burden_SRP1141379_24hr
```

```{r}
ir_ratio_zhu_24hours <- ggarrange(ir_ratio_zhu.3, ir_ratio_zhu.4, ir_ratio_zhu.3_sig,  ir_ratio_zhu.4_sig, ncol = 2, nrow = 2, labels =c("B", "E", "C", "F"))

ir_ratio_zhu_24hours

ir_volcano_plot_24hr_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_24hr, ir_volcano_plot_SRP141379_10uM_24hr, ncol = 2, labels = c("D", "G"))

Zhu_24hours_ir <- ggarrange(IR_burden_SRP1141379_24hr, ir_ratio_zhu_24hours, ir_volcano_plot_24hr_SRP141379,  ncol = 1, heights = c(1.25, 1.5, 0.7), labels = c("A", ""))
Zhu_24hours_ir

ggsave(Zhu_24hours_ir,
       filename = "Zhu_24hours_ir.PDF",
       height = 280,
       width = 180,
       units = "mm")
```


## Differentially expressed genes

```{r}
dEG_SRP141379_volcano <- ggarrange(volcano_zhu_6hr_3uM_gene, volcano_zhu_24hr_3uM_gene, volcano_zhu_6hr_10uM_gene,  volcano_zhu_24hr_10uM_gene, ncol =2, nrow = 2, labels = c("B", "D", "C", "E"))

plot_deg <- ggarrange(deg_ERP171667_plot, dEG_SRP141379_volcano, ncol = 1, labels = c("A", ""), heights = c(1, 1.5))

plot_deg

ggsave(plot = plot_deg,
       filename = "DEG_combined.PDF",
       height = 220,
       width = 180, unit = "mm")
```



