---
title: Differential intron retention in breast cancer cell lines treated with splicing
  inhibitors
author: "Monique White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(data.table)
library(ggpubr)
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
library(readxl)
source("ir_helpers.R")   # for make_ir_id()


metadata_sc5903_mw <- read_csv("metadata_sc5903_mw.csv")

delta_ir_srp_iwai <- read_tsv(
  "results/ir/SRP112607_delta_ir.tsv",
  show_col_types = FALSE
)
```

# 1. Download data

# 2. Fastqc data
Aim: identify adaptors for trimming prior to running IRFinder.

# 3. Trim data and merge data

# 4. IRFinder

## 4.1. Build reference genome

Using 150 bp mapability files for Liu et al. 2025 (Velde 2026)

Using 100 bp mapability file for Zhu et al. 2018 and ...


## 4.2. Run IRFinder

# 5. Shortlisting introns / filtering

## 5.1. 
First identify introns that have a mean IR ratio > 0.05 in at least one group.
### Velde et al: either DMSO or T025

### Zhu et al: either CC671 3uM @ 6 hours, CC671 10uM @ 6 hours, CC671 3uM @ 24 hours, or CC671 10uM @ 24 hours.

## 5.2. Within sample filtering
Met the criteria outlined in Shah et al. 2022.
1.
2.
3.
4.



# 6. Differential intron retention events (dIR)
```{r}
# Split into the two studies
Zhu <- metadata_sc5903_mw %>% filter(Citation == "Zhu et al. 2018")
Velde <- metadata_sc5903_mw %>% filter(Citation == "Velde et al. 2026") 
```

## 6.1. Velde et al. 2026 (DMSO vs T025)
```{r import dIR data for Velde et al 2026}
Velde_T025_vs_DMSO <- read_csv("results/ir/ERP171667/T025_vs_DMSO.csv")
Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |>
  mutate(
    "Hours" = 6,
    "Concentration" = "1uM",
    "Study" = "Velde"
  )

Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |> mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`,
    direction = case_when(
      log2FoldChange > 0 ~ "upregulated",
      log2FoldChange < 0 ~ "downregulated",
      TRUE ~ "no change")
  )

# Analysing introns where the delta IR ratio was > 0.05
erp_combined <- read_tsv(
  "results/ir/ERP171667_shah_filtered_combined.tsv",
  show_col_types = FALSE)

erp_combined <- erp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )
#sanity check
dim(erp_combined)
head(erp_combined)

# Delta IR
delta_erp <- read_tsv(
  "results/ir/ERP171667_delta_ir.tsv",
  show_col_types = FALSE
)

delta_erp <- delta_erp |> mutate(
    EnsemblID = str_extract(IntronID, "ENSG[0-9]+"))

delta_erp_filter <- delta_erp |> filter(delta_mean_ir_ratio_T025_vs_DMSO != "NA" &                                        delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                                        delta_mean_ir_ratio_T025_vs_DMSO <= -0.1)  #4917 introns present in both groups, but 1729 introns that have a delta IR ratio of greater than absolute 0.1

```

Aim:
Use the introns that have delta IR > absolute value of 0.1.
Then do the the mean IR ratio against each other
```{r}
erp_mean <- erp_combined |> filter(IntronID %in% delta_erp_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

# ---- 3. Now pivot_wider works, because each cell has one value ------------
erp_wide <- erp_mean %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ggplot(erp_wide, aes(x = T025, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - T025",
    y = "IR ratio - DMSO"
  ) +
  theme_minimal()
```

Aim: 
```{r}
mean_Velde <- Velde |>
  group_by(exp_design) |>
  summarise(
    mean_dir = mean(n_introns),
    se_dir   = sd(n_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )

ggplot() +
  #geom_boxplot(data = Velde, aes(x = exp_design, y = n_introns)) +
  geom_jitter(data = Velde, aes(x = exp_design, y = n_introns), width = 0.1) +
  geom_point(data = mean_Velde, aes(x = exp_design, y = mean_dir, colour = exp_design)) +
  geom_errorbar(
    data = mean_Velde,
    aes(x = exp_design, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir, colour = exp_design), width = 0.2)+
  theme_bw()+
  labs(y = "Number of intron retention events", colour = "")+
  theme(panel.grid.major.x = element_blank())
```

## 6.2. Zhu et al. 2018
```{r import dIR data for Zhu et al. 2018}
CC671_3uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")
CC671_3uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")

CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr|>
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM",
    "Study" = "Zhu")

CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr |> mutate(
    "Hours" = 24,
    "Concentration" = "3uM",
    "Study" = "Zhu")

CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr |> mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu")

CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr|> mutate(
    "Hours" = 24,
    "Concentration" = "10uM",
    "Study" = "Zhu")

# Filtering of delta IR
srp_combined <- read_tsv(
  "results/ir/SRP141379_shah_filtered_combined.tsv",
  show_col_types = FALSE)

srp_combined <- srp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )

# ---- distinct introns per gene ---------------------------------------------
srp_combined |>
  distinct(EnsemblID, IntronID) |>
  count(EnsemblID, name = "n_distinct_introns") |>
  arrange(desc(n_distinct_introns))

#Quality control data
srp_combined <- srp_combined |> 
  mutate(EnsemblID = str_extract(`Name`, "ENSG[0-9]+"))
#sanity check
dim(srp_combined)
head(srp_combined)


delta_ir_srp_zhu <- read_tsv(
  "results/ir/SRP141379_delta_ir_by_timepoint.tsv",
  show_col_types = FALSE
)
```

Aim: Number of intron retention events observed per sample.

```{r}
mean_Zhu <- Zhu |> group_by(exp_design, Hours, Concentration) |> 
  summarise(
    mean_dir = mean(n_introns),
    se_dir   = sd(n_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )

ggplot() +
  #geom_boxplot(data = Zhu, aes(x = exp_design, y = n_introns)) +
  geom_jitter(data = Zhu, aes(x = Concentration, y = n_introns), width = 0.1) +
  geom_point(data = mean_Zhu, aes(x = Concentration, y = mean_dir, colour = exp_design)) +
  geom_errorbar(
    data = mean_Zhu,
    aes(x = Concentration, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir, colour = exp_design), width = 0.2
  ) +
  facet_wrap(~ Hours)+
  theme_bw()+
  labs(y = "Number of intron retention events", colour = "", x = "Concentration [CC671]")+
  theme(panel.grid.major.x = element_blank())
```


### CC671 3uM @ 6 hours
```{r}
# 6 hours, 3uM
delta_zhu_6hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 6,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #8786 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #1268 introns

zhu_mean_6hr_3uM <- srp_combined |> filter(IntronID %in% delta_zhu_6hr_3uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

zhu_wide_6hr_3uM <- zhu_mean_6hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.1 <- ggplot(zhu_wide_6hr_3uM, aes(x = CC671_3uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 3uM",
    y = "IR ratio - DMSO",
    title = "6 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.1
```


### CC671 10uM @ 6 hours
```{r}
# 6 hours, 10uM
delta_zhu_6hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 6, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 8548 introns
                                                          delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) #1682 introns

zhu_mean_6hr_10uM <- srp_combined |> filter(IntronID %in% delta_zhu_6hr_10uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

zhu_wide_6hr_10uM <- zhu_mean_6hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.2 <- ggplot(zhu_wide_6hr_10uM, aes(x = CC671_10uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 10uM",
    y = "IR ratio - DMSO",
    title = "6 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.2
```

### CC671 3uM @ 24 hours

```{r}
# 24 hours, 3uM
delta_zhu_24hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 24,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #9302 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #773 introns 

zhu_mean_24hr_3uM <- srp_combined |> filter(IntronID %in% delta_zhu_24hr_3uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

zhu_wide_24hr_3uM <- zhu_mean_24hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.3 <- ggplot(zhu_wide_24hr_3uM, aes(x = CC671_3uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 3uM",
    y = "IR ratio - DMSO",
    title = "24 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.3
```


### CC671 10uM @ 24 hours

```{r}
# 24 hours, 10um
delta_zhu_24hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 24, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 9288 differentially expressed
                                                 delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) # 856 introns         

zhu_mean_24hr_10uM <- srp_combined |> filter(IntronID %in% delta_zhu_24hr_10uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

# ---- 3. Now pivot_wider works, because each cell has one value ------------
zhu_wide_24hr_10uM <- zhu_mean_24hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.4 <- ggplot(zhu_wide_24hr_10uM, aes(x = CC671_10uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 10uM",
    y = "IR ratio - DMSO",
    title = "24 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.4
```


# 7. Differential gene expression (dGE)

## 7.1. Velde et al. 2026 (DMSO vs T025)

## 7.2. Zhu et al. 2018
### CC671 3uM @ 6 hours

### CC671 10uM @ 6 hours

### CC671 3uM @ 24 hours

### CC671 10uM @ 24 hours

