---
title: Differential intron retention in breast cancer cell lines treated with splicing
  inhibitors
author: "Monique White"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(data.table)
library(ggpubr)
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
library(readxl)
source("ir_helpers.R")   # for make_ir_id()

metadata_sc5903_mw <- read_csv("metadata_sc5903_mw.csv")
metadata_sc5903_mw <- metadata_sc5903_mw |> 
  mutate(conc_uM = factor(conc_uM, ordered = TRUE))
```

# 1. Download data

# 2. Fastqc data
Aim: identify adaptors for trimming prior to running IRFinder.


# 3. Preparing data for IRFinder
## Trimming adaptors
Used cutadapt to trim the 15bp adaptors from all samples

## Merging lanes
Zhu et al had samples spread across multiple lanes, and thus required merging

## fastqc of cleaned data

# 4. IRFinder

## 4.1. Build reference genome

Using 150 bp mapability files for Liu et al. 2025 (Velde 2026)
```{r build reference genome with 150bp mapability files}

```


Using 100 bp mapability file for Zhu et al. 2018 and ...
```{r build reference genome with 100bp mapability files}

```

## 4.2. Run IRFinder

### Velde et al. 2026

```{r run IRFinder for ERP171667}

```

### Zhu et al. 2018
```{r run IRFinder for SRP141379}

```

### Iwai et al. 2018
```{r run IRFinder for SRP112607}

```


# 5. Shortlisting introns / filtering

## 5.1. 
First identify introns that have a mean IR ratio > 0.05 in at least one group.
### Velde et al: either DMSO or T025

### Zhu et al: either CC671 3uM @ 6 hours, CC671 10uM @ 6 hours, CC671 3uM @ 24 hours, or CC671 10uM @ 24 hours.

## 5.2. Within sample filtering
Met the criteria outlined in Shah et al. 2022.
1.
2.
3.
4.



# 6. Differential intron retention events (dIR)
```{r}
# Split into the two studies
Zhu <- metadata_sc5903_mw %>% filter(Citation == "Zhu et al. 2018")
Zhu <- Zhu |>  mutate(Concentration = factor(Concentration, levels = c("0uM", "3uM", "10uM"), ordered = TRUE))

Velde <- metadata_sc5903_mw %>% filter(Citation == "Velde et al. 2026") 

Iwai <- metadata_sc5903_mw |> filter(Citation == "Iwai et al. 2018") 
Iwai <- Iwai |> mutate(Concentration = factor(Concentration, levels = c("0nM", "30nM", "100nM", "300nM"), ordered = TRUE))

delta_ir_srp_iwai <- read_tsv(
  "results/ir/SRP112607_delta_ir.tsv",
  show_col_types = FALSE
)
```

## 6.1. Velde et al. 2026 (DMSO vs T025)
```{r import dIR data for Velde et al 2026}
Velde_T025_vs_DMSO <- read_csv("results/ir/ERP171667/T025_vs_DMSO.csv")
Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |>
  mutate(
    "Hours" = 6,
    "Concentration" = "1uM",
    "Study" = "Velde"
  )
Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |> 
  mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

# Analysing introns where the delta IR ratio was > 0.05
erp_combined <- read_tsv(
  "results/ir/ERP171667_shah_filtered_combined.tsv",
  show_col_types = FALSE)

erp_combined <- erp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )
# #sanity check
# dim(erp_combined)
# head(erp_combined)

# Delta IR
delta_erp <- read_tsv(
  "results/ir/ERP171667_delta_ir.tsv",
  show_col_types = FALSE
) #6106 rows

delta_erp <- delta_erp |> mutate(
    EnsemblID = str_extract(IntronID, "ENSG[0-9]+"))

delta_erp_filter <- delta_erp |> filter(delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                                        delta_mean_ir_ratio_T025_vs_DMSO <= -0.1)  
#2559 introns that have a delta IR ratio of greater than absolute 0.1.
```

Summary (ERP171667):
2559 introns that have a delta IR ratio of greater than or equal to |0.1|.


### Scatterplot
Aim:
Use the introns that have delta IR > absolute value of 0.1.
Then do the the mean IR ratio against each other
```{r eval=FALSE}
erp_mean <- erp_combined |> 
  filter(IntronID %in% delta_erp_filter$IntronID) |> 
  #only including introns that passed the structural filtering,
  #and have a delta IR ratio greater than or equal to |0.1|
  group_by(IntronID, condition) %>%
  summarise(
    #mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

erp_wide <- erp_mean %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio)
# 2559 Introns.

sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_erp_filter$IntronID)
#1452 intron IDs, and have a significant log2Fold Change of < 0.001 from DESeq
#1618 dIR have a base expression of 10, FDR < 0.01, and a delta IR ratio greater than or equal to |0.1|
erp_sig_wide <- erp_wide |> filter(IntronID %in% sig_dIR_erp$intron_id)

ggplot() +
  geom_point(data = erp_wide, aes(x = T025, y = DMSO), alpha = 0.4, size = 1) +
  geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "blue", alpha = 0.4, size = 1)+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - T025",
    y = "IR ratio - DMSO",
    title = "ERP171667"
  ) +
  theme_minimal()
```

Figure X. Scatterplot showing intron retention events that had a mean IR ratio of greater than 0.05 in at xx cells treated with DMSO or T025 after 24 hours. Intron retention events associated with significant dIR (p-adjust < 0.001) and delta IR ratio of > 0.1 or < -0.1 are shown in blue. dIR was tested using DESeq2.

1618 introns that have a base expression of 10, FDR < 0.01, and a delta IR ratio greater than or equal to |0.1|.

```{r volcano plot for dIR in velde et al. 2026, eval=FALSE}
erp_mean <- erp_combined |>
  mutate(
    delta_ir_filter = case_when(
      !IntronID %in% delta_erp_filter$IntronID ~ "Delta~IR~ratio<'|1|'",
      IntronID %in% delta_erp_filter$IntronID  ~ "Delta~IR~ratio>='|1|'"
    )
  ) |>
  group_by(IntronID, condition, delta_ir_filter) %>%
  summarise(
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )
  # filter(IntronID %in% delta_erp_filter$IntronID) |> 
  #only including introns that passed the structural filtering,
  #and have a delta IR ratio greater than or equal to |0.1|

erp_wide <- erp_mean %>%
  pivot_wider(
    id_cols   = c(IntronID, delta_ir_filter),
    names_from  = condition,
    values_from =  mean_ir_ratio)
# 6106 Introns.

# using the same criteria as Li et al. 2025 Sci Reports)
# padj (FDR) < 0.05
# absolute fold charge >= 2
 sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(padj < 0.05, log2FoldChange <= -2 | log2FoldChange >= 2)
 # 4694 observations

# sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(padj < 0.001, intron_id %in% delta_erp_filter$IntronID)
# #1452 intron IDs, and have a significant log2Fold Change of < 0.001 from DESeq
erp_sig_wide <- erp_wide |> filter(IntronID %in% sig_dIR_erp$intron_id)


sig_delta_and_dir <- Velde_T025_vs_DMSO |> filter(padj < 0.05, log2FoldChange <= -2 | log2FoldChange >= 2, intron_id %in% delta_erp_filter$IntronID)
# 1375 intron IDS
erp_sig_delta_and_dir <- erp_wide |> filter(IntronID %in% sig_delta_and_dir$intron_id)

ggplot() +
  geom_point(data = erp_wide, aes(x = T025, y = DMSO), colour = "black", alpha = 0.2, size = 0.5) +
  geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "pink", alpha = 0.2, size = 0.8)+
  geom_point(data = erp_sig_delta_and_dir, aes(x = T025, y = DMSO), colour = "red", alpha = 0.2, size = 0.8)+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#701705") +
  labs(
    x = "IR ratio - T025",
    y = "IR ratio - DMSO",
    title = "ERP171667"
  ) +
  theme_minimal()

ggplot() +
  #geom_point(data = erp_wide, aes(x = T025, y = DMSO), colour = "black", alpha = 0.2, size = 0.5) +
  geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "pink", alpha = 0.2, size = 0.8)+
  geom_point(data = erp_sig_delta_and_dir, aes(x = T025, y = DMSO), colour = "red", alpha = 0.2, size = 0.8)+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#701705") +
  labs(
    x = "IR ratio - T025",
    y = "IR ratio - DMSO",
    title = "ERP171667"
  ) +
  theme_minimal()


ggplot() +
  geom_point(data = erp_wide, aes(x = T025, y = DMSO), colour = "black", alpha = 0.2, size = 0.5) +
  # geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "pink", alpha = 0.2, size = 0.8)+
  # geom_point(data = erp_sig_delta_and_dir, aes(x = T025, y = DMSO), colour = "red", alpha = 0.2, size = 0.8)+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#701705") +
  labs(
    x = "IR ratio - T025",
    y = "IR ratio - DMSO",
    title = "ERP171667"
  ) +
  theme_minimal()
```

*note check with Ulf whether I should be filtering based on intron ID in deIR instead of the delta IR?*

### Intron retention events
Aim: 
```{r}
mean_Velde <- Velde |>
  group_by(conc_uM, exp_design) |>
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )

ERP171667_IR_burden <- ggplot() +
  #geom_boxplot(data = Velde, aes(x = exp_design, y = n_introns)) +
  geom_jitter(data = Velde, aes(x = conc_uM, y = sample_introns, colour = exp_design), width = 0.2) +
  geom_point(data = mean_Velde, aes(x = conc_uM, y = mean_dir), size = 2) +
  geom_errorbar(
    data = mean_Velde,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), width = 0.1)+
  
  theme_bw()+
  labs(y = "Intron retention events",
       x = expression("T-025 Concentration (" * mu * "M)")#, title = "ERP171667"
       )+
  scale_y_continuous(limits = c(0, 5000))+
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = c(0.95, 0.99),
legend.justification = c("right", "top")
        )

ggsave(filename = "ERP171667_IR_burden.PNG",
       plot = ERP171667_IR_burden,
       height = 180,
       width = 160,
       unit = "mm")

```
Figure X. Number of intron retention events passing filtering criteria (structural criteria within samples, and  IR ratio > 0.05 in at least one group after intra-sample filtering). Error bars represent standard error of the mean. Coloured dot is the mean of the samples.

### Volcano plot
```{r}
# merge delta IR and the dIR data
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))


volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  ) |> 
  filter(baseMean > 10)

volcano_velde |> group_by(significance) |> summarise(n_introns = n())


volc_data <- volcano_velde |>
  filter(baseMean > 10) |> #ensures that only introns with enough expression are included
  as.data.frame() |>
  na.omit()
# reduces down to 5447 introns 

ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "24 hours, T-025 1uM vs DMSO",
       x = expression(Delta ~ "IR ratio"))

ir_volcano_plot_ERP171667 <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())+
  labs(title = "24 hours, T-025 1uM vs DMSO",
       x = expression(Delta ~ "IR ratio (T-025/DMSO)"))

ir_volcano_plot_ERP171667
```

Figure X. Volcano plot of delta mean ir ratio in cells treated with T025 (n=3) and xx cells treated with DMSO (n =3) after 24 hours. Significant intron retention events are shown in red, where significance indicates the intron had a basemean > 10, the delta IR ratio was greater than 0.1, or less than -0.1, and the p-adjust (calculated in DESeq2 when calculating Log fold change) was < 0.001.


```{r}
ir_volcano_plot_ERP171667 <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())+
  labs(x = expression(Delta ~ "IR ratio (T-025 1" * mu * "M/DMSO)"))

erp_delta_ir_ratio_log2Fold <- left_join(
  Velde_T025_vs_DMSO, delta_erp,
  by = join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.01 ~ "#701705",
      TRUE                      ~ "#a3a8ac"
    )
  )

ir_l2fc_plot_ERP171667 <- ggplot() +
  theme_bw() +
  geom_point(
    data = erp_delta_ir_ratio_log2Fold,
    aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = log2FoldChange, color = significance),
    size = 0.6
  ) +
  scale_color_identity() +
  labs(x = expression(Delta ~ "IR ratio (T-025 1" * mu * "M/DMSO)")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())

ERP171667_volc_l2fc <- ggarrange(ir_volcano_plot_ERP171667, ir_l2fc_plot_ERP171667, labels = c("B", "C"), ncol = 1)

ERP_ir_plot <- ggarrange(ERP171667_IR_burden, ERP171667_volc_l2fc, labels = c("A", ""), ncol = 2, widths = c(1, 0.8))

ggsave(filename = "ERP171667_ir_plot.PNG",
       plot = ERP_ir_plot,
       width = 180,
       height = 160,
       unit = "mm")
```


### Delta IR ratio vs log2Fold change
```{r}
delta_erp #6106 observations
Velde_T025_vs_DMSO 


erp_delta_ir_ratio_log2Fold <- left_join(Velde_T025_vs_DMSO, delta_erp, by = join_by(intron_id == IntronID))


ggplot()+
  theme_bw()+
  geom_point(data = erp_delta_ir_ratio_log2Fold,
             aes(x=delta_mean_ir_ratio_T025_vs_DMSO, y= log2FoldChange))+
  labs(x = expression(Delta ~ "IR ratio (T-025/DMSO)"))+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())


```


Figure X. Volcano plot of log2FoldChange of intron expression in xx cells treated with T025 (n=3) and treated with DMSO (n =3) after 24 hours. Significant intron retention events are shown in red, where significance indicates the intron had a basemean > 10, the delta IR ratio was greater than 0.1, or less than -0.1, and the p-adjust (calculated in DESeq2 when calculating Log fold change) was < 0.001.

The coefficient conditionT025.IRFinderIR is testing the interaction between condition and the IR-vs-Splice factor — it's asking "does the ratio of intron-retention reads to spliced reads differ between T025 and DMSO?" A positive log2FoldChange means that ratio is higher in T025 than in DMSO — i.e., more intron retention (a higher IR ratio) in T025-treated cells relative to control.

*note - how can a padjust be so significant when the delta mean ir ratio is 0. I know padjust is calculated based on the log fold change...*

## 6.2. Zhu et al. 2018
```{r import dIR data for Zhu et al. 2018}
CC671_3uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")
CC671_3uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")

CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr|>
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM",
    "Study" = "Zhu",
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr |> mutate(
    "Hours" = 24,
    "Concentration" = "3uM",
    "Study" = "Zhu",
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr |> mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu", 
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr|> mutate(
    "Hours" = 24,
    "Concentration" = "10uM",
    "Study" = "Zhu",
    EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`)

# Filtering of delta IR
srp_combined <- read_tsv(
  "results/ir/SRP141379_shah_filtered_combined.tsv",
  show_col_types = FALSE)

srp_combined <- srp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )

# ---- distinct introns per gene ---------------------------------------------
srp_combined |>
  distinct(EnsemblID, IntronID) |>
  count(EnsemblID, name = "n_distinct_introns") |>
  arrange(desc(n_distinct_introns))

#Quality control data
srp_combined <- srp_combined |> 
  mutate(EnsemblID = str_extract(`Name`, "ENSG[0-9]+"))
#sanity check
dim(srp_combined)
head(srp_combined)


delta_ir_srp_zhu <- read_tsv(
  "results/ir/SRP141379_delta_ir_by_timepoint.tsv",
  show_col_types = FALSE
)
```

### Intron retention burden
Aim: Number of intron retention events observed per sample.

```{r}
mean_Zhu <- Zhu |> group_by(exp_design, Hours, conc_uM) |> 
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )

# Define labels for the facet
hour_labels <- c(
  "6"  = "6 hours",
  "24" = "24 hours"
)

ggplot() +
  geom_jitter(data = Zhu, aes(x = conc_uM, y = sample_introns, , colour = exp_design), width = 0.2) +
  geom_point(data = mean_Zhu, aes(x = conc_uM, y = mean_dir)) +
  geom_errorbar(
    data = mean_Zhu,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), 
    width = 0.1
  ) +
  facet_wrap(
    ~ Hours, 
    labeller = labeller(Hours = hour_labels)
  ) +
  theme_bw() +
  labs(
    y = "Intron retention events", 
    colour = "",
    x = expression("CC-671 Concentration (" * mu * "M)")
  ) +
  scale_y_continuous(limits = c(0, 5000))+
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = c(0.98, 0.99),
legend.justification = c("right", "top")
        )

IR_burden_SRP1141379 <- ggplot() +
  geom_jitter(data = Zhu, aes(x = conc_uM, y = sample_introns, , colour = exp_design), width = 0.2) +
  geom_point(data = mean_Zhu, aes(x = conc_uM, y = mean_dir)) +
  geom_errorbar(
    data = mean_Zhu,
    aes(x = conc_uM, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), 
    width = 0.1
  ) +
  facet_wrap(
    ~ Hours, 
    labeller = labeller(Hours = hour_labels)
  ) +
  theme_bw() +
  labs(
    y = "Intron retention events", 
    colour = "",
    x = expression("CC-671 Concentration (" * mu * "M)") #, title = "SRP1141379"
  ) +
  scale_y_continuous(limits = c(0, 5000))+
  theme(panel.grid.major.x = element_blank(),
        text = element_text(size = 11),
        legend.text = element_text(size = 11),
        strip.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        title = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = c(0.98, 0.99),
legend.justification = c("right", "top")
        )
  
ggsave(filename = "SRP1141379_IR_burden.PNG",
         plot = IR_burden_SRP1141379,
    units = "mm",width = 200,
         height = 180)
```


### Scatterplots of IR ratio

### CC671 3uM @ 6 hours
```{r}
# 6 hours, 3uM
delta_zhu_6hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 6,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #8786 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #1648 introns

zhu_mean_6hr_3uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_6hr_3uM_filter$IntronID, hours == 6) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")
# zhu_mean_6hr_3uM <- srp_combined |> filter(IntronID %in% delta_zhu_6hr_3uM_filter$IntronID) |>
#   group_by(IntronID, condition) %>%
#   summarise(
#     mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
#     mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
#     .groups = "drop"
#   )

zhu_wide_6hr_3uM <- zhu_mean_6hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.1 <- ggplot(zhu_wide_6hr_3uM, aes(x = CC671_3uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 3uM",
    y = "IR ratio - DMSO",
    title = "6 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.1
```




### CC671 10uM @ 6 hours
```{r}
# 6 hours, 10uM
delta_zhu_6hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 6, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 8548 introns
                                                          delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) #2215 introns

zhu_mean_6hr_10uM <- zhu_mean_6hr_3uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_6hr_10uM_filter$IntronID, hours == 6) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

zhu_wide_6hr_10uM <- zhu_mean_6hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.2 <- ggplot(zhu_wide_6hr_10uM, aes(x = CC671_10uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 10uM",
    y = "IR ratio - DMSO",
    title = "6 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.2
```

### CC671 3uM @ 24 hours

```{r}
# 24 hours, 3uM
delta_zhu_24hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 24,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #9302 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #948 introns 

zhu_mean_24hr_3uM <- zhu_mean_24hr_3uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_24hr_3uM_filter$IntronID, hours == 24) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

zhu_wide_24hr_3uM <- zhu_mean_24hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.3 <- ggplot(zhu_wide_24hr_3uM, aes(x = CC671_3uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 3uM",
    y = "IR ratio - DMSO",
    title = "24 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.3
```


### CC671 10uM @ 24 hours

```{r}
# 24 hours, 10um
delta_zhu_24hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 24, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 9288 differentially expressed
                                                 delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) # 1069 introns         

zhu_mean_24hr_10uM <- zhu_mean_24hr_10uM <- srp_combined |>
  filter(IntronID %in% delta_zhu_24hr_10uM_filter$IntronID, hours == 24) |>
  group_by(IntronID, condition) |>
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

# ---- 3. Now pivot_wider works, because each cell has one value ------------
zhu_wide_24hr_10uM <- zhu_mean_24hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.4 <- ggplot(zhu_wide_24hr_10uM, aes(x = CC671_10uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 10uM",
    y = "IR ratio - DMSO",
    title = "24 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.4
```

### Determine whether time or concentration drive difference in intron retention between control and treatment
```{r}
library(glmmTMB)

# Treat Hours and Concentration as factors if you only have a few discrete levels 

# (e.g. 0h/6h/24h, or Low/Med/High). Keep as numeric only if truly continuous/dose-response.
Zhu$Hours <- factor(Zhu$Hours)
Zhu$Concentration <- factor(Zhu$Concentration)
Zhu$group <- interaction(Zhu$exp_design, Zhu$Concentration, drop = TRUE)
# drop=TRUE removes empty combinations like Treatment.0uM

table(Zhu$group, Zhu$Hours)

model <- glmmTMB(
  directional_ir_counts ~ group * Hours,
  family = nbinom2,
  data = Zhu
)

summary(model)
drop1(model, test = "Chisq")

library(emmeans)
emmeans(model, pairwise ~ group | Hours, type = "response")
```

### Volcano plot
#### CC671 3uM @ 6 hours
```{r}
dIR_6hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_6hr_3uM <- delta_ir_srp_zhu |> filter(hours == 6)

volcano_zhu_6hr_3uM <- left_join(
  dIR_6hrs_CC671_3uM_vs_DMSO,
  delta_ir_6hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  )

volc_data <- volcano_zhu_6hr_3uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

volc_data |> group_by(significance) |> summarise(n_introns = n())

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "6 hours, CC671 3uM vs DMSO",
       x = expression(Delta ~ "IR ratio"))

ir_volcano_plot_SRP141379_3uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "6 hours",
       x = expression(Delta ~ "IR ratio (CC671 3" * mu * "M/DMSO)"))



```

```{r}
ir_l2fc_plot_SRP141379_3uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = log2FoldChange, colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "6 hours",
       x = expression(Delta ~ "IR ratio (CC671 3" * mu * "M/DMSO)")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
```


#### CC671 10uM @ 6 hours
```{r}
dIR_6hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_6hr_10uM <- delta_ir_srp_zhu |> filter(hours == 6)

volcano_zhu_6hr_10uM <- left_join(
  dIR_6hrs_CC671_10uM_vs_DMSO,
  delta_ir_6hr_10uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  )

volc_data <- volcano_zhu_6hr_10uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  theme_bw()+
  labs(title = "6 hours, CC671 10uM vs DMSO",
       x = expression(Delta ~ "IR ratio"))

ir_volcano_plot_SRP141379_10uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(
    #title = "6 hours",
       x = expression(Delta ~ "IR ratio (CC671 10" * mu * "M/DMSO)"))

# IR ratio and log2 fold change
ir_l2fc_plot_SRP141379_10uM_6hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = log2FoldChange, colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(
    #title = "6 hours",
       x = expression(Delta ~ "IR ratio (CC671 10" * mu * "M/DMSO)")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
```


#### CC671 3uM @ 24 hours

```{r}
dIR_24hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)
delta_ir_24hr_3uM <- delta_ir_srp_zhu |> filter(hours == 24)


volcano_zhu_24hr_3uM <- left_join(
  dIR_24hrs_CC671_3uM_vs_DMSO,
  delta_ir_24hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  )

volc_data <- volcano_zhu_24hr_3uM |>
  filter(baseMean > 10) |> #9332 introns
  as.data.frame() |>
  na.omit()

volc_data |> group_by(significance) |> summarise(n_introns = n())

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.7) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "24 hours",
       x= expression(Delta ~ "IR ratio (CC671 3" * mu * "M/DMSO)"))

ir_volcano_plot_SRP141379_3uM_24hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.7) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "24 hours",
       x= expression(Delta ~ "IR ratio (CC671 3" * mu * "M/DMSO)"))

# Delta IR a d log2Fold
ir_l2fc_plot_SRP141379_3uM_24hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = log2FoldChange, colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "24 hours",
       x = expression(Delta ~ "IR ratio (CC671 3" * mu * "M/DMSO)")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())

```

9332 introns have base mean > 10


#### CC671 10uM @ 24 hours

```{r}
dIR_24hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_24hr_10uM <- delta_ir_srp_zhu |> filter(hours == 24, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA")

volcano_zhu_24hr_10uM <- left_join(
  dIR_24hrs_CC671_10uM_vs_DMSO,
  delta_ir_24hr_10uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  )

volc_data <- volcano_zhu_24hr_10uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()
volc_data |> group_by(significance) |> summarise(n_introns = n())

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(title = "24 hours",
       x = expression(Delta ~ "IR ratio (CC671 10" * mu * "M/DMSO)"))


ir_volcano_plot_SRP141379_10uM_24hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(
    #title = "24 hours",
       x = expression(Delta ~ "IR ratio (CC671 10" * mu * "M/DMSO)"))

# Delta IR ratio and log2fold change
ir_l2fc_plot_SRP141379_10uM_24hr <- ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = log2FoldChange, colour = significance)) +
  geom_point(alpha = 0.6, size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  scale_x_continuous(limits = c(-1, 1)) +
  theme(panel.grid.minor = element_blank())+
  labs(
    #title = "24 hours",
       x = expression(Delta ~ "IR ratio (CC671 10" * mu * "M/DMSO)")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
```

9332 introns

```{r}
ir_volcano_plot_full_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_6hr, ir_volcano_plot_SRP141379_10uM_6hr, ir_volcano_plot_SRP141379_3uM_24hr, ir_volcano_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("A", "B", "C", "D"))


ggsave(ir_volcano_plot_full_SRP141379,
       filename = "ir_volc_SRP141379.pdf",
       height = 180,
       width = 180,
       units = "mm")


```

```{r}
ir_volcano_plot_full_SRP141379 <- ggpubr::ggarrange(ir_volcano_plot_SRP141379_3uM_6hr, ir_volcano_plot_SRP141379_3uM_24hr, ir_volcano_plot_SRP141379_10uM_6hr, ir_volcano_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("B", "C", "", ""))

ir_l2fc_plot_full_SRP141379 <- ggpubr::ggarrange(ir_l2fc_plot_SRP141379_3uM_6hr, ir_l2fc_plot_SRP141379_3uM_24hr, ir_l2fc_plot_SRP141379_10uM_6hr, ir_l2fc_plot_SRP141379_10uM_24hr, ncol = 2, nrow =2, labels = c("D", "E", "", ""))

SRP141379_ir_burden_volcano <- ggarrange(IR_burden_SRP1141379, ir_volcano_plot_full_SRP141379, ir_l2fc_plot_full_SRP141379, ncol = 1, heights = c(1, 1.2, 1.2), labels = c("A", "", ""))

ggsave(SRP141379_ir_burden_volcano,
       filename = "SRP141379_ir_burden_volcano.png",
       height = 280,
       width = 180,
       units = "mm")
```

## 6.3. Iwai et al. 2018

### Intron retention burden

```{r}
mean_iwai <- Iwai |>
  group_by(exp_design) |>
  summarise(
    mean_dir = mean(sample_introns),
    se_dir   = sd(sample_introns) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )

ggplot() +
  #geom_boxplot(data = Velde, aes(x = exp_design, y = n_introns)) +
  geom_point(data = mean_iwai, aes(x = exp_design, y = mean_dir), size = 2) +
  geom_errorbar(
    data = mean_iwai,
    aes(x = exp_design, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir), width = 0.2)+
  geom_jitter(data = Iwai, aes(x = exp_design, y = sample_introns, colour = Concentration), width = 0.1, height = 0) +
  theme_bw()+
  labs(y = "Number of intron retention events", colour = "")+
  theme(panel.grid.major.x = element_blank())


ggplot() +
  #geom_boxplot(data = Velde, aes(x = exp_design, y = n_introns)) +
  geom_point(data = Iwai, aes(x = conc_uM, y = sample_introns, colour = exp_design), size = 3, width = 0.1, height = 0) +
  theme_bw()+
  labs(y = "Intron retention events", colour = "",
       x = expression("[T-025] (" * mu * "M)"),
       title = "SRP112607")+
  scale_y_continuous(limits = c(0, 8500))+
  theme(panel.grid.major.x = element_blank())

```


# 7. Gene ontology of dIR
## 7.1. Velde et al. 2026 (DMSO vs T025)
```{r}
# merge delta IR and the dIR data
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID)) #6106 observations

significant_velde_introns <- volcano_velde |>
  filter(baseMean > 10, delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 | delta_mean_ir_ratio_T025_vs_DMSO <= -0.1 & padj < 0.001) #1994 observations

#if I change p adjust to 0.05?
# significant_velde_introns <- volcano_velde |>
#   filter(baseMean > 10, delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 | delta_mean_ir_ratio_T025_vs_DMSO <= -0.1 & padj < 0.05) #2094 only adds 100 introns

significant_velde_introns <- significant_velde_introns |>
  mutate(direction_deltaIR = case_when(
    delta_mean_ir_ratio_T025_vs_DMSO > 0 ~ "increased_retention",
    delta_mean_ir_ratio_T025_vs_DMSO < 0 ~ "decreased_retention",
    TRUE ~ NA_character_
  ))

sig_introns_up <- significant_velde_introns |> 
  filter(direction_deltaIR == "increased_retention") |> 
  pull(intron_id) |> unique() #1463

sig_genes_up   <- significant_velde_introns |> 
  filter(direction_deltaIR == "increased_retention") |> 
  pull(EnsemblID.x) |> unique() #1141 genes increased

# 1463 introns across 1141 genes were differentially expressed in the cells treated with T-025

sig_introns_down <- significant_velde_introns |> 
  filter(direction_deltaIR == "decreased_retention") |> 
  pull(intron_id) |> unique() #531

sig_genes_down <- significant_velde_introns |> 
  filter(direction_deltaIR == "decreased_retention") |> 
  pull(EnsemblID.x) |> unique() #446

```


If this DESeq2 run - like most IRFinder/IR-DESeq2 pipelines - was fit on raw intron-spanning read counts (not on the IR ratio itself), then log2FoldChange reflects the change in absolute intron read counts, normalized only by DESeq2's library-size factors. That's fundamentally different from delta_mean_ir_ratio, which reflects the change in intron reads as a proportion of total reads at that intron (intron / (intron + spliced)) - i.e. normalized within the gene, controlling for how much the gene's overall expression changed.

This produces exactly the asymmetric pattern you're seeing:

If overall gene expression drops while the proportion of unspliced transcript stays the same or increases → absolute intron read counts can still fall (-> log2FoldChange < 0, "downregulated") even though the IR ratio goes up (-> increased_retention). That's your 188 mismatched genes.
If overall gene expression is stable or rising alongside a real increase in retention -> both metrics agree (-> your clean 86/88 "upregulated" agreement).

this means your padj significance was computed on a metric (raw intron counts) that isn't the same one you're using for direction (ratio)

Delta IR going up (positive) means: the proportion of transcripts at that intron that retain it - rather than getting spliced out - has increased in the treatment condition relative to control. IR ratio itself is a ratio, not an absolute expression level: roughly speaking, IRratio approximately equal to intron-covering reads / (intron-covering reads + spliced-junction reads) at that locus. So a positive delta IR does mean the intron-retained isoform's relative share went up - yes, in that specific sense, you can say the retained-intron isoform is "increased" (more of the total transcript pool at that intron now carries the retained intron).



```{r gene ontology of dIR, eval=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

# --- 1. Define gene sets by direction (based on delta IR, not log2FoldChange) ---
sig_genes_up <- significant_velde_introns |>
  filter(delta_mean_ir_ratio_T025_vs_DMSO > 0) |>
  pull(EnsemblID.x) |>
  unique() # 251

sig_genes_down <- significant_velde_introns |>
  filter(delta_mean_ir_ratio_T025_vs_DMSO < 0) |>
  pull(EnsemblID.x) |>
  unique() #94

# --- 2. Define the background/universe ---
# This should be every gene TESTED in this comparison (baseMean > 10, non-NA padj),
# BEFORE any significance filtering — i.e. your pre-filter equivalent of volc_data
# for the Velde T025_vs_DMSO comparison. Adjust object name as needed.
volc_data <- volcano_velde |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

background_genes <- volc_data |>
  pull(EnsemblID.x) |>
  unique() #3138

# --- 3. Run GO enrichment separately for each direction ---

ego_up <- enrichGO(
  gene          = sig_genes_up,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff = 1
  #readable      = TRUE
)

ego_down <- enrichGO(
  gene          = sig_genes_down,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 1
  #readable      = TRUE
)

# --- 4. Inspect results ---

as.data.frame(ego_up)   |> head(20)
as.data.frame(ego_down) |> head(20)

# --- 5. Plot ---

dotplot(ego_up,   showCategory = 20, title = "Increased intron retention")
dotplot(ego_down, showCategory = 20, title = "Decreased intron retention")

```


## 7.2. Zhu et al. 2018

### CC671 3uM @ 6 hours

```{r}
volcano_zhu_6hr_3uM <- left_join(
  dIR_6hrs_CC671_3uM_vs_DMSO,
  delta_ir_6hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    ),
  EnsemblID = str_extract(intron_id, "ENSG[0-9]+")
  )

volc_data <- volcano_zhu_6hr_3uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

background_genes <- volc_data |>
  pull(EnsemblID) |>
  unique()

significant_zhu_6hr_3um <- volcano_zhu_6hr_3uM |> filter(significance == "red") |> 
  mutate(direction_deltaIR = case_when(
    delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0 ~ "increased_retention",
    delta_mean_ir_ratio_CC671_3uM_vs_DMSO < 0 ~ "decreased_retention",
    TRUE ~ NA_character_
  ),
  EnsemblID = str_extract(intron_id, "ENSG[0-9]+"))


sig_genes_up <- significant_zhu_6hr_3um |> 
  filter(direction_deltaIR == "increased_retention") |> 
  pull(EnsemblID) |> unique()

sig_genes_down <- significant_zhu_6hr_3um |> 
  filter(direction_deltaIR == "decreased_retention") |> 
  pull(EnsemblID) |> unique()

# GO
ego_up <- enrichGO(
  gene          = sig_genes_up,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

ego_down <- enrichGO(
  gene          = sig_genes_down,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

# --- 4. Inspect results ---

as.data.frame(ego_up)   |> head(20)
as.data.frame(ego_down) |> head(20)

# --- 5. Plot ---

dotplot(ego_up,   showCategory = 20, title = "Increased intron retention")
dotplot(ego_down, showCategory = 20, title = "Decreased intron retention")
```

```{r eval = FALSE}
# Optional: side-by-side comparison of both directions
ego_compare <- compareCluster(
  geneCluster   = list(increased_retention = sig_genes_up,
                        decreased_retention = sig_genes_down),
  fun           = "enrichGO",
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

# dotplot(ego_compare, showCategory = 10)
```

### CC671 10uM @ 6 hours

### CC671 3uM @ 24 hours

### CC671 10uM @ 24 hours

# 8. Differential gene expression (dGE)

Differential gene expression was calculated on genes where the sum of genes are kept only if the row sum of raw counts across all samples in that model is > 10. Filtering is evaluated per-model, on
whatever sample subset that model uses.

## 8.1. Velde et al. 2026 (DMSO vs T025)
```{r}
dEG_T025_vs_DMSO <- read_csv("results/genes/ERP171667/T025_vs_DMSO.csv") 
dEG_T025_vs_DMSO <- dEG_T025_vs_DMSO |> mutate("EnsemblID" = ...1)

dEG_T025_vs_DMSO <- dEG_T025_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.001 ~ "red",
      log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.01  ~ "#701705",
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    ))

dEG_T025_vs_DMSO |> group_by(significance) |> summarise(n_genes = n())

volc_data <- dEG_T025_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(x = expression("log2FoldChange (T-025 1" * mu * "M/DMSO)")
       #, title = "ERP171667"
       )+
  theme(panel.grid.minor = element_blank())
```

### Upregulated?

```{r}
dEG_T025_vs_DMSO |> 
  filter(log2FoldChange >= 3,
         padj < 0.01)

#2151 genes of 17 724 are have increased expression in T025 compared to DMSO
```

2151 genes of 17 724 are have increased expression in T025 compared to DMSO


### Downregulated

### Upregualted genes the same genes with increased IR?

Are the genes upregulated in expression the same genes from which the isoforms have increasing IR ratio
```{r}
dEG_ERP171667_upreg_filter <- dEG_T025_vs_DMSO |> 
  filter(log2FoldChange >= 3,
         padj <= 0.05) |> 
  pull(EnsemblID)

volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  )

volcano_velde |> filter(significance != "#a3a8ac") |>
  #1988 introns
  filter(EnsemblID.x %in% dEG_ERP171667_upreg_filter) # 64 genes that have upregulated introns have the genes differentially expressed

ERP171667_upreg_genes_and_increased_ir <- volcano_velde |> filter(significance != "#a3a8ac") |>
  #1988 introns
  filter(EnsemblID.x %in% dEG_ERP171667_upreg_filter) 
```

### Intron retention events that are significant but the expression of the gene is not upregulated (dEG) (where upregulated is defined as a log2FoldChange of >= 3)
```{r}
volcano_velde |> filter(significance != "#a3a8ac") |>
  #1988 introns
  filter(!EnsemblID.x %in% dEG_ERP171667_upreg_filter) #1924 introns 
```


## 8.2. Zhu et al. 2018

### CC671 3uM @ 6 hours
```{r}
dEG_6hr_CC671_3uM_vs_DMSO <- read_csv("results/genes/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")

dEG_6hr_CC671_3uM_vs_DMSO <- dEG_6hr_CC671_3uM_vs_DMSO |> mutate("EnsemblID" = ...1)

dEG_6hr_CC671_3uM_vs_DMSO <- dEG_6hr_CC671_3uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.001 ~ "red",
      log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.01  ~ "#701705",
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    ))

volc_data <- dEG_6hr_CC671_3uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(x = "log2FoldChange (CC671 3uM/DMSO)",
       title = "6 hours")

volcano_zhu_6hr_3uM_gene <-ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(
    x = expression("log2FoldChange (CC671 3" * mu * "M/DMSO)"),
       title = "6 hours") +
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(limits = c(-5, 11))
```

#### Upregulated genes
```{r}
dEG_6hr_CC671_3uM_vs_DMSO |> 
  filter(log2FoldChange >= 3,
         padj <= 0.05)

#351 genes of 18114 are have increased expression in cells treated with CC671 compared to DMSO
```

#### Increased gene expression == increased intron retention?
```{r}
SRP141379_3uM_6hr_upreg_filter <- dEG_6hr_CC671_3uM_vs_DMSO |> 
  filter(log2FoldChange >= 3,
         padj <= 0.05)|> 
  pull(EnsemblID)
# 351 introns

SRP141379_3uM_6hr_upreg_genes_and_increased_ir <-  volcano_zhu_6hr_3uM |>
  filter(baseMean > 10) |> filter(significance != "#a3a8ac") |>
  #1079 introns
  filter(EnsemblID %in% SRP141379_3uM_6hr_upreg_filter) 
# 19 introns with significant delta IR, and log2Fold change were derived from genes that had upregulated gene expression in treated cells
```



### CC671 10uM @ 6 hours
```{r}
dEG_6hr_CC671_10uM_vs_DMSO <- read_csv("results/genes/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
dEG_6hr_CC671_10uM_vs_DMSO <- dEG_6hr_CC671_10uM_vs_DMSO |> mutate("EnsemblID" = ...1)

dEG_6hr_CC671_10uM_vs_DMSO <- dEG_6hr_CC671_10uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.001 ~ "red",
      log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.01  ~ "#701705",
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    ))

volc_data <- dEG_6hr_CC671_10uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(
     x = expression("log2FoldChange (CC671 10" * mu * "M/DMSO)"),
       title = "6 hours")

volcano_zhu_6hr_10uM_gene <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(
     x = expression("log2FoldChange (CC671 10" * mu * "M/DMSO)")
     #, title = "6 hours"
     )+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(limits = c(-5, 11))
```

#### Upregulated genes
```{r}
dEG_6hr_CC671_10uM_vs_DMSO |> 
  filter(log2FoldChange >= 3,
         padj <= 0.05)

#475 genes of 18114 are have increased expression in cells treated with CC671 compared to DMSO
```

### CC671 3uM @ 24 hours
```{r}
dEG_24hr_CC671_3uM_vs_DMSO <- read_csv("results/genes/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
dEG_24hr_CC671_3uM_vs_DMSO <- dEG_24hr_CC671_3uM_vs_DMSO |> mutate("EnsemblID" = ...1)

# Volcano Plot
dEG_24hr_CC671_3uM_vs_DMSO <- dEG_24hr_CC671_3uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.001 ~ "red",
      log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.01  ~ "#701705",
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    ))

volc_data <- dEG_24hr_CC671_3uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(
     x = expression("log2FoldChange (CC671 3" * mu * "M/DMSO)"),
       title = "24 hours")

volcano_zhu_24hr_3uM_gene <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(x = expression("log2FoldChange (CC671 3" * mu * "M/DMSO)"),
       title = "24 hours")+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(limits = c(-5, 7.5))
```

#### Upregulated genes
```{r}
dEG_24hr_CC671_3uM_vs_DMSO |> 
  filter(log2FoldChange >= 3,
         padj <= 0.05)

#88 genes of 17923 are have increased expression in cells treated with CC671 compared to DMSO
```

### CC671 10uM @ 24 hours

```{r}
dEG_24hr_CC671_10uM_vs_DMSO <- read_csv("results/genes/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")
dEG_24hr_CC671_10uM_vs_DMSO <- dEG_24hr_CC671_10uM_vs_DMSO |> mutate("EnsemblID" = ...1)


dEG_24hr_CC671_10uM_vs_DMSO <- dEG_24hr_CC671_10uM_vs_DMSO |> filter(baseMean > 10) |> #indicates enough expression for it to be meaningful
  mutate(
    significance = case_when(
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.001 ~ "red",
      log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.01  ~ "#701705",
      #log2FoldChange >= 3 | log2FoldChange <= -3 & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    ))

volc_data <- dEG_24hr_CC671_10uM_vs_DMSO |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs(x = expression("log2FoldChange (CC671 10" * mu * "M/DMSO)")#,title = "24 hours"
       )

volcano_zhu_24hr_10uM_gene <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj), colour = significance)) +
  geom_point(size = 0.6) +
  scale_colour_identity() +
  theme_bw()+
  labs( x = expression("log2FoldChange (CC671 10" * mu * "M/DMSO)")
        #, title = "24 hours"
        )+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(limits = c(-5, 7.5))
```

```{r}
dEG_SRP141379_volcano <- ggarrange(volcano_zhu_6hr_3uM_gene, volcano_zhu_24hr_3uM_gene, volcano_zhu_6hr_10uM_gene,  volcano_zhu_24hr_10uM_gene, ncol =2, nrow = 2, labels = c("A", "B", "", ""))

ggsave(filename = "dEG_SRP141379_volcano.PNG",
       plot = dEG_SRP141379_volcano,
       height = 180,
       width = 180,
       unit = "mm")

```


# 9. Cross study comparisons
## 9.1. Are intron events consistent between T-025 and CC-671 after 24 hours?
```{r}
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))
volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.1,
    significance = case_when(
      #meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "#701705",
      #meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "#a3a8ac"
    )
  )

sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(baseMean > 10,
  padj < 0.01, 
  intron_id %in% delta_erp_filter$IntronID)

#1618 dIR have a base expression of 10, FDR < 0.01, and a delta IR ratio greater than or equal to |0.1|
```


```{r}
# name each Zhu dataset with its delta column, so the loop knows which column to use
zhu_datasets <- list(
  "6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  "24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)


filter_hits <- function(df, delta_col, delta_thresh = 0.1) {
  df |>
    filter(
      .data[[delta_col]] >= delta_thresh,   # only increases in IR ratio
      significance != "#a3a8ac", # FDR < 0.01
      baseMean > 10 #expression sufficient for differential analysis
    ) |>
    pull(intron_id)
}

# ERP171667 hits (baseline, computed once)

velde_hits <- filter_hits(volcano_velde, "delta_mean_ir_ratio_T025_vs_DMSO")

# Loop over the four Zhu comparisons
comparison_results <- map_dfr(names(zhu_datasets), function(nm) {
  entry    <- zhu_datasets[[nm]]
  zhu_hits <- filter_hits(entry$df, entry$delta_col)
  shared   <- intersect(velde_hits, zhu_hits)

  tibble(
    comparison   = nm,
    n_velde_hits = length(velde_hits),
    n_zhu_hits   = length(zhu_hits),
    n_shared     = length(shared),
    shared_ids   = list(shared)   # list-column so you can inspect IDs later
  )
})

comparison_results
```

```{r}
library(UpSetR)

all_hits <- list(
  `T-025 1uM 24hr`    = velde_hits,
  `CC-671 3uM 6hr`   = filter_hits(volcano_zhu_6hr_3uM,   "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  `CC-671 10uM 6hr`  = filter_hits(volcano_zhu_6hr_10uM,  "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  `CC-671 3uM 24hr`  = filter_hits(volcano_zhu_24hr_3uM,  "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  `CC-671 10uM 24hr` = filter_hits(volcano_zhu_24hr_10uM, "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)

# upset(fromList(all_hits), order.by = "freq")

set_sizes <- sapply(all_hits, length)
sorted_names <- names(sort(set_sizes, decreasing = TRUE))

# rebuild bar_colors in the same sorted order
color_lookup <- c(
  `T-025 1uM 24hr`    = "#005b96",
  `CC-671 3uM 6hr`    = "#ee5d6c",
  `CC-671 10uM 6hr`   = "#ee5d6c",
  `CC-671 3uM 24hr`   = "#ce4993",
  `CC-671 10uM 24hr`  = "#ce4993"
)
sorted_colors <- color_lookup[sorted_names]

dIR_significant_shared_plot <- upset(fromList(all_hits),
      order.by = "freq",
      sets = sorted_names,
      keep.order = TRUE,
      sets.bar.color = sorted_colors,
      text.scale = 1.5)


```

```{r}
library(dplyr)
library(purrr)
library(tibble)

zhu_datasets <- list(
  "6hr_3uM"   = list(df = volcano_zhu_6hr_3uM,   delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "6hr_10uM"  = list(df = volcano_zhu_6hr_10uM,  delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO"),
  "24hr_3uM"  = list(df = volcano_zhu_24hr_3uM,  delta_col = "delta_mean_ir_ratio_CC671_3uM_vs_DMSO"),
  "24hr_10uM" = list(df = volcano_zhu_24hr_10uM, delta_col = "delta_mean_ir_ratio_CC671_10uM_vs_DMSO")
)

# direction = "up", "down", or "both" (both = union of up & down, not requiring same sign in comparison)
filter_hits <- function(df, delta_col, direction = "both", delta_thresh = 0.1) {
  d <- .data <- NULL  # avoid lint noise
  vals <- df[[delta_col]]

  keep <- switch(direction,
    up   = vals >= delta_thresh,
    down = vals <= -delta_thresh,
    both = vals >= delta_thresh | vals <= -delta_thresh
  )

  df |>
    filter(keep,
           significance != "#a3a8ac",
           baseMean > 10) |>
    pull(intron_id)
}

# Velde hits, split by direction
velde_up   <- filter_hits(volcano_velde, "delta_mean_ir_ratio_T025_vs_DMSO", "up")
velde_down <- filter_hits(volcano_velde, "delta_mean_ir_ratio_T025_vs_DMSO", "down")

comparison_results <- map_dfr(names(zhu_datasets), function(nm) {
  entry <- zhu_datasets[[nm]]

  zhu_up   <- filter_hits(entry$df, entry$delta_col, "up")
  zhu_down <- filter_hits(entry$df, entry$delta_col, "down")

  shared_up      <- intersect(velde_up, zhu_up)          # both increase
  shared_down    <- intersect(velde_down, zhu_down)      # both decrease
  shared_opposite_ud <- intersect(velde_up, zhu_down)     # Velde up, Zhu down
  shared_opposite_du <- intersect(velde_down, zhu_up)     # Velde down, Zhu up

  tibble(
    comparison        = nm,
    n_velde_up        = length(velde_up),
    n_velde_down      = length(velde_down),
    n_zhu_up          = length(zhu_up),
    n_zhu_down        = length(zhu_down),
    n_shared_up       = length(shared_up),
    n_shared_down     = length(shared_down),
    n_shared_opposite = length(shared_opposite_ud) + length(shared_opposite_du),
    n_shared_same_dir = length(shared_up) + length(shared_down),
    shared_up_ids      = list(shared_up),
    shared_down_ids     = list(shared_down)
  )
})

comparison_results |>
  dplyr::select(comparison, n_velde_up, n_velde_down, n_zhu_up, n_zhu_down,
         n_shared_up, n_shared_down, n_shared_same_dir, n_shared_opposite)
```

```{r}
up_only <- list(
  `T-025 24hr 1uM` = velde_up,
  `CC-671 6hr 3uM`    = filter_hits(volcano_zhu_6hr_3uM,   "delta_mean_ir_ratio_CC671_3uM_vs_DMSO",  "up"),
  `CC-671 6hr 10uM`   = filter_hits(volcano_zhu_6hr_10uM,  "delta_mean_ir_ratio_CC671_10uM_vs_DMSO", "up"),
  `CC-671 24hr 3uM`   = filter_hits(volcano_zhu_24hr_3uM,  "delta_mean_ir_ratio_CC671_3uM_vs_DMSO",  "up"),
  `CC-671 24hr 10uM`  = filter_hits(volcano_zhu_24hr_10uM, "delta_mean_ir_ratio_CC671_10uM_vs_DMSO", "up")
)

upset(fromList(up_only), order.by = "freq", nsets = 5)
```

22 introns were shared in all 5 samples.

```{r}
shared_all <- Reduce(intersect, up_only)
length(shared_all)   # should be 22
shared_all           # the intron IDs themselves

shared_all_annotated <- purrr::reduce(
  list(
    volcano_velde |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_velde   = delta_mean_ir_ratio_T025_vs_DMSO,
             padj_velde    = padj,
             baseMean_velde = baseMean,
             log2FC_velde   = log2FoldChange),

    volcano_zhu_6hr_3uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_6hr_3uM   = delta_mean_ir_ratio_CC671_3uM_vs_DMSO,
             padj_6hr_3uM    = padj,
             baseMean_6hr_3uM = baseMean,
             log2FC_6hr_3uM   = log2FoldChange),

    volcano_zhu_6hr_10uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_6hr_10uM   = delta_mean_ir_ratio_CC671_10uM_vs_DMSO,
             padj_6hr_10uM    = padj,
             baseMean_6hr_10uM = baseMean,
             log2FC_6hr_10uM   = log2FoldChange),

    volcano_zhu_24hr_3uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_24hr_3uM   = delta_mean_ir_ratio_CC671_3uM_vs_DMSO,
             padj_24hr_3uM    = padj,
             baseMean_24hr_3uM = baseMean,
             log2FC_24hr_3uM   = log2FoldChange),

    volcano_zhu_24hr_10uM |>
      filter(intron_id %in% shared_all) |>
      dplyr::select(intron_id,
             delta_24hr_10uM   = delta_mean_ir_ratio_CC671_10uM_vs_DMSO,
             padj_24hr_10uM    = padj,
             baseMean_24hr_10uM = baseMean,
             log2FC_24hr_10uM   = log2FoldChange)
  ),
  left_join, by = "intron_id"
)

shared_all_annotated
```

```{r}
library(tidyr)
library(dplyr)
library(stringr)

shared_all_long <- shared_all_annotated |>
  pivot_longer(
    cols = -intron_id,
    names_to = c(".value", "condition"),
    names_pattern = "^(delta|padj|baseMean|log2FC)_(.*)$"
  )

shared_all_long
shared_all_long <- shared_all_long |>
  mutate(condition = factor(condition,
                             levels = c("velde", "6hr_3uM", "6hr_10uM", "24hr_3uM", "24hr_10uM")))
shared_all_long <- shared_all_long |>
  mutate(
    condition = recode(condition,
      "velde"      = "T-025 1uM 24hrs",
      "6hr_3uM"    = "CC-671 3uM 6hrs",
      "6hr_10uM"   = "CC-671 10uM 6hrs",
      "24hr_3uM"   = "CC-671 3uM 24hrs",
      "24hr_10uM"  = "CC-671 10uM 24hrs"
    ),
    condition = factor(condition,
      levels = c("T-025 1uM 24hrs",
                 "CC-671 3uM 6hrs",
                 "CC-671 10uM 6hrs",
                 "CC-671 3uM 24hrs",
                 "CC-671 10uM 24hrs")
    )
  )

library(ggplot2)

ggplot(shared_all_long, aes(x = condition, y = intron_id, fill = delta)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

