---
title: Differential intron retention in breast cancer cell lines treated with splicing
  inhibitors
author: "Monique White"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(data.table)
library(ggpubr)
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
library(readxl)
source("ir_helpers.R")   # for make_ir_id()

metadata_sc5903_mw <- read_csv("metadata_sc5903_mw.csv")

delta_ir_srp_iwai <- read_tsv(
  "results/ir/SRP112607_delta_ir.tsv",
  show_col_types = FALSE
)
```

# 1. Download data

# 2. Fastqc data
Aim: identify adaptors for trimming prior to running IRFinder.


# 3. Preparing data for IRFinder
## Trimming adaptors
Used cutadapt to trim the 15bp adaptors from all samples

## Merging lanes
Zhu et al had samples spread across multiple lanes, and thus required merging

## fastqc of cleaned data

# 4. IRFinder

## 4.1. Build reference genome

Using 150 bp mapability files for Liu et al. 2025 (Velde 2026)
```{r build reference genome with 150bp mapability files}

```


Using 100 bp mapability file for Zhu et al. 2018 and ...
```{r build reference genome with 100bp mapability files}

```

## 4.2. Run IRFinder

### Velde et al. 2026

```{r run IRFinder for ERP171667}

```

### Zhu et al. 2018
```{r run IRFinder for SRP141379}

```

### Iwai et al. 2018
```{r run IRFinder for SRP112607}

```


# 5. Shortlisting introns / filtering

## 5.1. 
First identify introns that have a mean IR ratio > 0.05 in at least one group.
### Velde et al: either DMSO or T025

### Zhu et al: either CC671 3uM @ 6 hours, CC671 10uM @ 6 hours, CC671 3uM @ 24 hours, or CC671 10uM @ 24 hours.

## 5.2. Within sample filtering
Met the criteria outlined in Shah et al. 2022.
1.
2.
3.
4.



# 6. Differential intron retention events (dIR)
```{r}
# Split into the two studies
Zhu <- metadata_sc5903_mw %>% filter(Citation == "Zhu et al. 2018")
Zhu <- Zhu |>  mutate(Concentration = factor(Concentration, levels = c("0uM", "3uM", "10uM"), ordered = TRUE))

Velde <- metadata_sc5903_mw %>% filter(Citation == "Velde et al. 2026") 
```

## 6.1. Velde et al. 2026 (DMSO vs T025)
```{r import dIR data for Velde et al 2026}
Velde_T025_vs_DMSO <- read_csv("results/ir/ERP171667/T025_vs_DMSO.csv")
Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |>
  mutate(
    "Hours" = 6,
    "Concentration" = "1uM",
    "Study" = "Velde"
  )

Velde_T025_vs_DMSO <- Velde_T025_vs_DMSO |> mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`,
    direction = case_when(
      log2FoldChange > 0 ~ "upregulated",
      log2FoldChange < 0 ~ "downregulated",
      TRUE ~ "no change")
  )

# Analysing introns where the delta IR ratio was > 0.05
erp_combined <- read_tsv(
  "results/ir/ERP171667_shah_filtered_combined.tsv",
  show_col_types = FALSE)

erp_combined <- erp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )
# #sanity check
# dim(erp_combined)
# head(erp_combined)

# Delta IR
delta_erp <- read_tsv(
  "results/ir/ERP171667_delta_ir.tsv",
  show_col_types = FALSE
)

delta_erp <- delta_erp |> mutate(
    EnsemblID = str_extract(IntronID, "ENSG[0-9]+"))

delta_erp_filter <- delta_erp |> filter(delta_mean_ir_ratio_T025_vs_DMSO != "NA" &                                        delta_mean_ir_ratio_T025_vs_DMSO >= 0.1 |
                                        delta_mean_ir_ratio_T025_vs_DMSO <= -0.1)  #4917 introns present in both groups, but 1729 introns that have a delta IR ratio of greater than absolute 0.1

```

Summary (ERP171667):
4917 introns present in both groups (DMSO and T025), but 1729 introns that have a delta IR ratio of greater than absolute 0.1.


### Scatterplot
Aim:
Use the introns that have delta IR > absolute value of 0.1.
Then do the the mean IR ratio against each other
```{r volcano plot for dIR in velde et al. 2026}
erp_mean <- erp_combined |> 
  filter(IntronID %in% delta_erp_filter$IntronID) |> #only including introns that passed the structural filtering, and the group mean IR ratio filtering
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop")

erp_wide <- erp_mean %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio)

sig_dIR_erp <- Velde_T025_vs_DMSO |> filter(padj < 0.001, intron_id %in% delta_erp_filter$IntronID)
erp_sig_wide <- erp_wide |> filter(IntronID %in% sig_dIR_erp$intron_id)

ggplot() +
  geom_point(data = erp_wide, aes(x = T025, y = DMSO), alpha = 0.4, size = 1) +
  geom_point(data = erp_sig_wide, aes(x = T025, y = DMSO), colour = "blue", alpha = 0.4, size = 1)+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - T025",
    y = "IR ratio - DMSO"
  ) +
  theme_minimal()
```

Figure X. Scatterplot showing intron retention events that had a mean IR ratio of greater than 0.05 in at xx cells treated with DMSO or T025 after 24 hours. Intron retention events associated with significant dIR (p-adjust < 0.001) and delta IR ratio of > 0.1 or < -0.1 are shown in blue. dIR was tested using DESeq2.


*note check with Ulf whether I should be filtering based on intron ID in deIR instead of the delta IR?*

### Intron retention events
Aim: 
```{r}
mean_Velde <- Velde |>
  group_by(exp_design) |>
  summarise(
    mean_dir = mean(n_introns_sample_after_group_filter),
    se_dir   = sd(n_introns_sample_after_group_filter) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )

ggplot() +
  #geom_boxplot(data = Velde, aes(x = exp_design, y = n_introns)) +
  geom_point(data = mean_Velde, aes(x = exp_design, y = mean_dir, colour = exp_design), size = 2) +
  geom_errorbar(
    data = mean_Velde,
    aes(x = exp_design, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir, colour = exp_design), width = 0.2)+
  geom_jitter(data = Velde, aes(x = exp_design, y = n_introns_sample_after_group_filter), width = 0.1) +
  theme_bw()+
  labs(y = "Number of intron retention events", colour = "")+
  theme(panel.grid.major.x = element_blank())
```
Figure X. Number of intron retention events passing filtering criteria (structural criteria within samples, and  IR ratio > 0.05 in at least one group after intra-sample filtering). Error bars represent standard error of the mean. Coloured dot is the mean of the samples.

### Volcano plot
```{r}
# merge delta IR and the dIR data
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))

volcano_velde <- volcano_velde |> 
  mutate(
    meets_delta = delta_mean_ir_ratio_T025_vs_DMSO >= 0.25 |
                  delta_mean_ir_ratio_T025_vs_DMSO <= -0.25,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    )
  )


volc_data <- volcano_velde |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = delta_mean_ir_ratio_T025_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point() +
  scale_colour_identity() +
  theme_bw()

```

Figure X. Volcano plot of delta mean ir ratio in cells treated with T025 (n=3) and xx cells treated with DMSO (n =3) after 24 hours. Significant intron retention events are shown in red, where significance indicates the intron had a basemean > 10, the delta IR ratio was greater than 0.25, or less than -0.25, and the p-adjust (calculated in DESeq2 when calculating Log fold change) was < 0.001.

```{r eval=FALSE}
# merge delta IR and the dIR data
volcano_velde <- left_join(Velde_T025_vs_DMSO, delta_erp, join_by(intron_id == IntronID))

  # # filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  # as.data.frame() %>% 
  # na.omit() %>% 
  # #rownames_to_column("GeneID") %>% 
  # arrange(padj) %>% 
  # filter(padj<0.001)

#1344 introns if using delta ir of 0.1
#254 if using delta ir of 0.25


volc_data <- volcano_velde%>% 
  filter(baseMean > 10#, 
         #log2FoldChange > 1 | log2FoldChange < -1
         ) |>
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_velde_introns, color="red")+
  theme_bw()
```

Figure X. Volcano plot of log2FoldChange of intron expression in xx cells treated with T025 (n=3) and treated with DMSO (n =3) after 24 hours. Significant intron retention events are shown in red, where significance indicates the intron had a basemean > 10, the delta IR ratio was greater than 0.25, or less than -0.25, and the p-adjust (calculated in DESeq2 when calculating Log fold change) was < 0.001.

The coefficient conditionT025.IRFinderIR is testing the interaction between condition and the IR-vs-Splice factor — it's asking "does the ratio of intron-retention reads to spliced reads differ between T025 and DMSO?" A positive log2FoldChange means that ratio is higher in T025 than in DMSO — i.e., more intron retention (a higher IR ratio) in T025-treated cells relative to control.

*note - how can a padjust be so significant when the delta mean ir ratio is 0. I know padjust is calculated based on the log fold change...*

## 6.2. Zhu et al. 2018
```{r import dIR data for Zhu et al. 2018}
CC671_3uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")
CC671_3uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_6hr <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
CC671_10uM_vs_DMSO_24hr <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")

CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr|>
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM",
    "Study" = "Zhu")

CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr |> mutate(
    "Hours" = 24,
    "Concentration" = "3uM",
    "Study" = "Zhu")

CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr |> mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu")

CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr|> mutate(
    "Hours" = 24,
    "Concentration" = "10uM",
    "Study" = "Zhu")

# Filtering of delta IR
srp_combined <- read_tsv(
  "results/ir/SRP141379_shah_filtered_combined.tsv",
  show_col_types = FALSE)

srp_combined <- srp_combined |>
  mutate(
    EnsemblID = str_extract(Name, "ENSG[0-9]+"),
    IntronID  = make_ir_id(pick(everything()))   # Name|Chr|Start|End|Strand
  )

# ---- distinct introns per gene ---------------------------------------------
srp_combined |>
  distinct(EnsemblID, IntronID) |>
  count(EnsemblID, name = "n_distinct_introns") |>
  arrange(desc(n_distinct_introns))

#Quality control data
srp_combined <- srp_combined |> 
  mutate(EnsemblID = str_extract(`Name`, "ENSG[0-9]+"))
#sanity check
dim(srp_combined)
head(srp_combined)


delta_ir_srp_zhu <- read_tsv(
  "results/ir/SRP141379_delta_ir_by_timepoint.tsv",
  show_col_types = FALSE
)
```

### Intron retention burden
Aim: Number of intron retention events observed per sample.

```{r}
mean_Zhu <- Zhu |> group_by(exp_design, Hours, Concentration) |> 
  summarise(
    mean_dir = mean(n_introns_sample_after_group_filter),
    se_dir   = sd(n_introns_sample_after_group_filter) / sqrt(n()),   # sd / sqrt(n)
    .groups  = "drop"
  )


ggplot() +
  #geom_boxplot(data = Zhu, aes(x = exp_design, y = n_introns)) +
  geom_jitter(data = Zhu, aes(x = Concentration, y = n_introns_sample_after_group_filter), width = 0.1) +
  geom_point(data = mean_Zhu, aes(x = Concentration, y = mean_dir, colour = exp_design)) +
  geom_errorbar(
    data = mean_Zhu,
    aes(x = Concentration, ymin = mean_dir - se_dir, ymax = mean_dir + se_dir, colour = exp_design), width = 0.2
  ) +
  facet_wrap(~ Hours)+
  theme_bw()+
  labs(y = "Number of intron retention events", colour = "", x = "Concentration [CC671]")+
  theme(panel.grid.major.x = element_blank())
```


### Scatterplots of IR ratio

### CC671 3uM @ 6 hours
```{r}
# 6 hours, 3uM
delta_zhu_6hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 6,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #8786 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #1268 introns

zhu_mean_6hr_3uM <- srp_combined |> filter(IntronID %in% delta_zhu_6hr_3uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

zhu_wide_6hr_3uM <- zhu_mean_6hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.1 <- ggplot(zhu_wide_6hr_3uM, aes(x = CC671_3uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 3uM",
    y = "IR ratio - DMSO",
    title = "6 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.1
```


### CC671 10uM @ 6 hours
```{r}
# 6 hours, 10uM
delta_zhu_6hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 6, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 8548 introns
                                                          delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) #1682 introns

zhu_mean_6hr_10uM <- srp_combined |> filter(IntronID %in% delta_zhu_6hr_10uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

zhu_wide_6hr_10uM <- zhu_mean_6hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.2 <- ggplot(zhu_wide_6hr_10uM, aes(x = CC671_10uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 10uM",
    y = "IR ratio - DMSO",
    title = "6 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.2
```

### CC671 3uM @ 24 hours

```{r}
# 24 hours, 3uM
delta_zhu_24hr_3uM_filter <- delta_ir_srp_zhu |> filter(hours == 24,
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO != "NA" & #9302 introns present in both groups
                                                       delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.1) #773 introns 

zhu_mean_24hr_3uM <- srp_combined |> filter(IntronID %in% delta_zhu_24hr_3uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

zhu_wide_24hr_3uM <- zhu_mean_24hr_3uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.3 <- ggplot(zhu_wide_24hr_3uM, aes(x = CC671_3uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 3uM",
    y = "IR ratio - DMSO",
    title = "24 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.3
```


### CC671 10uM @ 24 hours

```{r}
# 24 hours, 10um
delta_zhu_24hr_10uM_filter <- delta_ir_srp_zhu |> filter(hours == 24, delta_mean_ir_ratio_CC671_10uM_vs_DMSO != "NA" & # 9288 differentially expressed
                                                 delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.1 | delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.1) # 856 introns         

zhu_mean_24hr_10uM <- srp_combined |> filter(IntronID %in% delta_zhu_24hr_10uM_filter$IntronID) |>
  group_by(IntronID, condition) %>%
  summarise(
    mean_intron_depth = mean(IntronDepth, na.rm = TRUE),
    mean_ir_ratio      = mean(IRratio, na.rm = TRUE),
    .groups = "drop"
  )

# ---- 3. Now pivot_wider works, because each cell has one value ------------
zhu_wide_24hr_10uM <- zhu_mean_24hr_10uM %>%
  pivot_wider(
    id_cols   = IntronID,
    names_from  = condition,
    values_from =  mean_ir_ratio
  )

ir_ratio_zhu.4 <- ggplot(zhu_wide_24hr_10uM, aes(x = CC671_10uM, y = DMSO)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "IR ratio - CC671 10uM",
    y = "IR ratio - DMSO",
    title = "24 hours"
  ) +
  theme_minimal()

ir_ratio_zhu.4
```

### Determine whether time or concentration drive difference in intron retention between control and treatment
```{r}
library(glmmTMB)

# Treat Hours and Concentration as factors if you only have a few discrete levels 

# (e.g. 0h/6h/24h, or Low/Med/High). Keep as numeric only if truly continuous/dose-response.
Zhu$Hours <- factor(Zhu$Hours)
Zhu$Concentration <- factor(Zhu$Concentration)
Zhu$group <- interaction(Zhu$exp_design, Zhu$Concentration, drop = TRUE)
# drop=TRUE removes empty combinations like Treatment.0uM

table(Zhu$group, Zhu$Hours)

model <- glmmTMB(
  directional_ir_counts ~ group * Hours,
  family = nbinom2,
  data = Zhu
)

summary(model)
drop1(model, test = "Chisq")

library(emmeans)
emmeans(model, pairwise ~ group | Hours, type = "response")
```

### Volcano plot
#### CC671 3uM @ 6 hours
```{r}
dIR_6hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_6hr_3uM <- delta_ir_srp_zhu |> filter(hours == 6)

volcano_zhu_6hr_3uM <- left_join(
  dIR_6hrs_CC671_3uM_vs_DMSO,
  delta_ir_6hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.25 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.25,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    )
  )

volc_data <- volcano_zhu_6hr_3uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point() +
  scale_colour_identity() +
  theme_bw()

```


#### CC671 10uM @ 6 hours
```{r}
dIR_6hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/6hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)


delta_ir_6hr_10uM <- delta_ir_srp_zhu |> filter(hours == 6)

volcano_zhu_6hr_10uM <- left_join(
  dIR_6hrs_CC671_10uM_vs_DMSO,
  delta_ir_6hr_10uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.25 |
                  delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.25,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    )
  )

volc_data <- volcano_zhu_6hr_10uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point() +
  scale_colour_identity() +
  theme_bw()
```


#### CC671 3uM @ 24 hours

```{r}
dIR_24hrs_CC671_3uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_3uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_24hr_3uM <- delta_ir_srp_zhu |> filter(hours == 24)


volcano_zhu_24hr_3uM <- left_join(
  dIR_24hrs_CC671_3uM_vs_DMSO,
  delta_ir_24hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.25 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.25,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    )
  )

volc_data <- volcano_zhu_24hr_3uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_3uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point() +
  scale_colour_identity() +
  theme_bw()

```

#### CC671 10uM @ 24 hours

```{r}
dIR_24hrs_CC671_10uM_vs_DMSO <- read_csv("results/ir/SRP141379/24hr/CC671_10uM_vs_DMSO.csv") |> 
  mutate(intron_id = `...1`)

delta_ir_24hr_10uM <- delta_ir_srp_zhu |> filter(hours == 24)

volcano_zhu_24hr_10uM <- left_join(
  dIR_24hrs_CC671_10uM_vs_DMSO,
  delta_ir_24hr_10uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_10uM_vs_DMSO >= 0.25 |
                  delta_mean_ir_ratio_CC671_10uM_vs_DMSO <= -0.25,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    )
  )

volc_data <- volcano_zhu_24hr_10uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

ggplot(volc_data, aes(x = delta_mean_ir_ratio_CC671_10uM_vs_DMSO, y = -log10(padj), colour = significance)) +
  geom_point() +
  scale_colour_identity() +
  theme_bw()
```

# 7. Gene ontology of dIR
## 7.1. Velde et al. 2026 (DMSO vs T025)
```{r}
significant_velde_introns <- volcano_velde |>
  filter(baseMean > 10, delta_mean_ir_ratio_T025_vs_DMSO >= 0.25 | delta_mean_ir_ratio_T025_vs_DMSO <= -0.25 & padj < 0.001)


significant_velde_introns <- significant_velde_introns |>
  mutate(direction_deltaIR = case_when(
    delta_mean_ir_ratio_T025_vs_DMSO > 0 ~ "increased_retention",
    delta_mean_ir_ratio_T025_vs_DMSO < 0 ~ "decreased_retention",
    TRUE ~ NA_character_
  ))

sig_genes_up   <- significant_velde_introns |> 
  filter(direction_deltaIR == "increased_retention") |> 
  pull(EnsemblID.x) |> unique()

sig_genes_down <- significant_velde_introns |> 
  filter(direction_deltaIR == "decreased_retention") |> 
  pull(EnsemblID.x) |> unique()

significant_velde_introns |> 
  count(direction, direction_deltaIR)
```


If this DESeq2 run — like most IRFinder/IR-DESeq2 pipelines — was fit on raw intron-spanning read counts (not on the IR ratio itself), then log2FoldChange reflects the change in absolute intron read counts, normalized only by DESeq2's library-size factors. That's fundamentally different from delta_mean_ir_ratio, which reflects the change in intron reads as a proportion of total reads at that intron (intron / (intron + spliced)) — i.e. normalized within the gene, controlling for how much the gene's overall expression changed.

This produces exactly the asymmetric pattern you're seeing:

If overall gene expression drops while the proportion of unspliced transcript stays the same or increases → absolute intron read counts can still fall (→ log2FoldChange < 0, "downregulated") even though the IR ratio goes up (→ increased_retention). That's your 188 mismatched genes.
If overall gene expression is stable or rising alongside a real increase in retention → both metrics agree (→ your clean 86/88 "upregulated" agreement).

this means your padj significance was computed on a metric (raw intron counts) that isn't the same one you're using for direction (ratio)

```{r gene ontology of dIR}
library(clusterProfiler)
library(org.Hs.eg.db)

# --- 1. Define gene sets by direction (based on delta IR, not log2FoldChange) ---

sig_genes_up <- significant_velde_introns |>
  filter(delta_mean_ir_ratio_T025_vs_DMSO > 0) |>
  pull(EnsemblID.x) |>
  unique() # 251

sig_genes_down <- significant_velde_introns |>
  filter(delta_mean_ir_ratio_T025_vs_DMSO < 0) |>
  pull(EnsemblID.x) |>
  unique() #94

# --- 2. Define the background/universe ---
# This should be every gene TESTED in this comparison (baseMean > 10, non-NA padj),
# BEFORE any significance filtering — i.e. your pre-filter equivalent of volc_data
# for the Velde T025_vs_DMSO comparison. Adjust object name as needed.
volc_data <- volcano_velde |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

background_genes <- volc_data |>
  pull(EnsemblID.x) |>
  unique() #2780

# --- 3. Run GO enrichment separately for each direction ---

ego_up <- enrichGO(
  gene          = sig_genes_up,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

ego_down <- enrichGO(
  gene          = sig_genes_down,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

# --- 4. Inspect results ---

as.data.frame(ego_up)   |> head(20)
as.data.frame(ego_down) |> head(20)

# --- 5. Plot ---

dotplot(ego_up,   showCategory = 20, title = "Increased intron retention")
dotplot(ego_down, showCategory = 20, title = "Decreased intron retention")

# Optional: side-by-side comparison of both directions
ego_compare <- compareCluster(
  geneCluster   = list(increased_retention = sig_genes_up,
                        decreased_retention = sig_genes_down),
  fun           = "enrichGO",
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

dotplot(ego_compare, showCategory = 10)
```


## 7.2. Zhu et al. 2018
### CC671 3uM @ 6 hours
```{r}
volcano_zhu_6hr_3uM <- left_join(
  dIR_6hrs_CC671_3uM_vs_DMSO,
  delta_ir_6hr_3uM,
  join_by(intron_id == IntronID)
) |>
  mutate(
    meets_delta = delta_mean_ir_ratio_CC671_3uM_vs_DMSO >= 0.25 |
                  delta_mean_ir_ratio_CC671_3uM_vs_DMSO <= -0.25,
    significance = case_when(
      meets_delta & padj < 0.001 ~ "red",
      meets_delta & padj < 0.01  ~ "pink",
      meets_delta & padj < 0.05  ~ "salmon",
      TRUE                       ~ "black"
    ),
  EnsemblID = str_extract(intron_id, "ENSG[0-9]+")
  )

volc_data <- volcano_zhu_6hr_3uM |>
  filter(baseMean > 10) |>
  as.data.frame() |>
  na.omit()

background_genes <- volc_data |>
  pull(EnsemblID) |>
  unique()

significant_zhu_6hr_3um <- volcano_zhu_6hr_3uM |> filter(significance == "red") |> 
  mutate(direction_deltaIR = case_when(
    delta_mean_ir_ratio_CC671_3uM_vs_DMSO > 0 ~ "increased_retention",
    delta_mean_ir_ratio_CC671_3uM_vs_DMSO < 0 ~ "decreased_retention",
    TRUE ~ NA_character_
  ),
  EnsemblID = str_extract(intron_id, "ENSG[0-9]+"))


sig_genes_up <- significant_zhu_6hr_3um |> 
  filter(direction_deltaIR == "increased_retention") |> 
  pull(EnsemblID) |> unique()

sig_genes_down <- significant_zhu_6hr_3um |> 
  filter(direction_deltaIR == "decreased_retention") |> 
  pull(EnsemblID) |> unique()



# GO
ego_up <- enrichGO(
  gene          = sig_genes_up,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

ego_down <- enrichGO(
  gene          = sig_genes_down,
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE
)

# --- 4. Inspect results ---

as.data.frame(ego_up)   |> head(20)
as.data.frame(ego_down) |> head(20)

# --- 5. Plot ---

dotplot(ego_up,   showCategory = 20, title = "Increased intron retention")
dotplot(ego_down, showCategory = 20, title = "Decreased intron retention")

# Optional: side-by-side comparison of both directions
ego_compare <- compareCluster(
  geneCluster   = list(increased_retention = sig_genes_up,
                        decreased_retention = sig_genes_down),
  fun           = "enrichGO",
  universe      = background_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

dotplot(ego_compare, showCategory = 10)
```

### CC671 10uM @ 6 hours

### CC671 3uM @ 24 hours

### CC671 10uM @ 24 hours

# 8. Differential gene expression (dGE)

Differential gene expression was calculated on genes where the sum of genes are kept only if the row sum of raw counts across all samples in that model is > 10. Filtering is evaluated per-model, on
whatever sample subset that model uses.

## 8.1. Velde et al. 2026 (DMSO vs T025)
```{r}
dEG_T025_vs_DMSO <- read_csv("results/genes/ERP171667/T025_vs_DMSO.csv")
```

How many are upregulated
How many are downregulated


## 8.2. Zhu et al. 2018


### CC671 3uM @ 6 hours
```{r}
dEG_6hr_CC671_3uM_vs_DMSO <- read_csv("results/genes/SRP141379/6hr/CC671_3uM_vs_DMSO.csv")
```


### CC671 10uM @ 6 hours
```{r}
dEG_6hr_CC671_10uM_vs_DMSO <- read_csv("results/genes/SRP141379/6hr/CC671_10uM_vs_DMSO.csv")
```


### CC671 3uM @ 24 hours
```{r}
dEG_24hr_CC671_3uM_vs_DMSO <- read_csv("results/genes/SRP141379/24hr/CC671_3uM_vs_DMSO.csv")
```


### CC671 10uM @ 24 hours
```{r}
dEG_24hr_CC671_10uM_vs_DMSO <- read_csv("results/genes/SRP141379/24hr/CC671_10uM_vs_DMSO.csv")
```



# 9. Gene ontology of dEGs

## 9.1. Velde et al. 2026 (DMSO vs T025)

## 9.2. Zhu et al. 2018
### CC671 3uM @ 6 hours

### CC671 10uM @ 6 hours

### CC671 3uM @ 24 hours

### CC671 10uM @ 24 hours