---
title: "Special Topics Differential analysis"
author: "Monique White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)

Velde_T025_vs_DMSO_1uM_24h <- read_csv("T025_vs_DMSO_1uM_24h(in).csv")
```

Genes with an average read count > 10 in all samples were selected for differential gene expression analysis (n = 23,072). Genes with an absolute log2 fold change > 1 and FDR-adjusted p < 0.05 were considered significant. 

```{r}

signif_genes <- Velde_T025_vs_DMSO_1uM_24h %>%
  filter(baseMean > 10, log2FoldChange > 1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)

volc_data <- Velde_T025_vs_DMSO_1uM_24h %>% 
  filter(baseMean > 10, log2FoldChange > 1) |>
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

