---
title: "Special Topics Differential analysis"
author: "Monique White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(data.table)

Velde_T025_vs_DMSO_1uM_24h <- read_csv("T025_vs_DMSO_1uM_24h(2).csv")
Velde_T025_vs_DMSO_1uM_24h <- Velde_T025_vs_DMSO_1uM_24h |>
  mutate(
    "Hours" = 6,
    "Concentration" = "1uM",
    "Study" = "Velde"
  )
```

Genes with an average read count > 10 in all samples were selected for differential gene expression analysis (n = 23,072). Genes with an absolute log2 fold change > 1 and FDR-adjusted p < 0.05 were considered significant. 

```{r}
significant_velde_genes <- Velde_T025_vs_DMSO_1uM_24h %>%
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)

volc_data <- Velde_T025_vs_DMSO_1uM_24h %>% 
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |>
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_velde_genes, color="red")+
  theme_bw()
```

The number of IR events in a sample was determined based on introns with an IR ratio > 0.1 and meeting the filtering criteria described above. 

```{r eval = FALSE}
# Set the folder once
data_dir <- "C:/Users/Monique/OneDrive/Documents/SC5901 Biomedicine Lab/BrCa.Data/BrCa-Special-Topics"

files <- c(
  "30nM"  = file.path(data_dir, "30nM_vs_0nM.txt"),
  "100nM" = file.path(data_dir, "100nM_vs_0nM.txt"),
  "300nM" = file.path(data_dir, "300nM_vs_0nM.txt")
)

read_irfinder <- function(path, label) {
  dt <- fread(path, skip = 2, header = TRUE, sep = "\t")   # <-- use path, not a fixed string
  dt[, c("GeneName", "GeneID", "IntronType") := tstrsplit(`Intron-GeneName/GeneID`, "/", fixed = TRUE, keep = 1:3)]
  dt[, Comparison := label]
  dt
}

ir_list <- Map(read_irfinder, files, names(files))
ir_all <- rbindlist(ir_list, use.names = TRUE, fill = TRUE)

head(ir_all)
table(ir_all$Comparison)
```

```{r}
ir_all |> 
  filter(`p-increased` < 0.05 | `p-decreased` < 0.05) #filters from 22,072 to 56,785

ir_all |> 
  filter(`B-IRratio`>0.1, `p-increased` < 0.05 | `p-decreased` < 0.05) #1,457


ir_all |> 
  filter(Comparison == "30nM",
         `p-increased` < 0.05 | `p-decreased` < 0.05) #2,151 

ir_all |> 
  filter(Comparison == "100nM",
         `p-increased` < 0.05 | `p-decreased` < 0.05) #6,978

ir_all |> 
  filter(Comparison == "300nM",
         `p-increased` < 0.05 | `p-decreased` < 0.05) #12,943 

## Filtering based on IR ratio

ir_all |> 
  filter(Comparison == "30nM",
         `B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05) #306 

ir_all |> 
  filter(Comparison == "100nM",
         `B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05) #518

ir_all |> 
  filter(Comparison == "300nM",
         `B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05) #633

# Filtered dats to compare to other studies
Iwai_ir_filtered <- ir_all |> filter(`B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05)
```

Seems to be a dose dependent effect there, with more introns



# Zhu et al
```{r}
library(readr)
CC671_3uM_vs_DMSO_6hr <- read_csv("CC671_3uM_vs_DMSO_6hr.csv")
CC671_3uM_vs_DMSO_24hr <- read_csv("CC671_3uM_vs_DMSO_24hr.csv")
CC671_10uM_vs_DMSO_6hr <- read_csv("CC671_10uM_vs_DMSO_6hr.csv")
CC671_10uM_vs_DMSO_24hr <- read_csv("CC671_10uM_vs_DMSO_24hr.csv")

CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr|>
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM",
    "Study" = "Zhu"
  )

CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr |> mutate(
    "Hours" = 24,
    "Concentration" = "3uM",
    "Study" = "Zhu"
  )

CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr |> mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu"
  )

CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr|> mutate(
    "Hours" = 24,
    "Concentration" = "10uM",
    "Study" = "Zhu"
  )
```

```{r}
significant_DE_CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr %>%
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)|> 
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM", 
    "Study" = "Zhu"
  )
volc_data <- CC671_3uM_vs_DMSO_6hr %>% 
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_3uM_6hr <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_3uM_vs_DMSO_6hr, color="red")+
  theme_bw()+
  labs(title = "3uM, 6hr")
```

```{r}
significant_DE_CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr %>%
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)|> 
  mutate(
    "Hours" = 24,
    "Concentration" = "3uM", 
    "Study" = "Zhu"
  )
volc_data <- CC671_3uM_vs_DMSO_24hr %>% 
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_3uM_24hr <- ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_3uM_vs_DMSO_24hr, color="red")+
  theme_bw()+
  labs(title = "3uM, 24hr")
```

```{r}
significant_DE_CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr %>%
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)|> 
  mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu"
  )

volc_data <- CC671_10uM_vs_DMSO_6hr %>% 
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_10uM_6hr <- ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_10uM_vs_DMSO_6hr, color="red")+
  theme_bw()+
  labs(title = "10uM, 6hr")
```

```{r}
significant_DE_CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr %>%
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05) |> 
  mutate(
    "Hours" = 24,
    "Concentration" = "10uM", 
    "Study" = "Zhu"
  )

volc_data <- CC671_10uM_vs_DMSO_24hr %>% 
  filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_10uM_24hr <- ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_10uM_vs_DMSO_24hr, color="red")+
  theme_bw()+
  theme(panel.grid.minor = element_blank())+
  labs(title = "10uM, 24hr")
```


```{r}
library(ggpubr)

ggarrange(volc_plot_3uM_6hr, volc_plot_10uM_6hr, volc_plot_3uM_24hr, volc_plot_10uM_24hr,
          ncol =2, nrow = 2,
          labels = c("A", "B", "C", "D"))
```

```{r}
Zhu_DE_IR <- bind_rows(significant_DE_CC671_3uM_vs_DMSO_6hr, significant_DE_CC671_3uM_vs_DMSO_24hr, significant_DE_CC671_10uM_vs_DMSO_6hr, significant_DE_CC671_10uM_vs_DMSO_24hr)
library(dplyr)
library(stringr)

Zhu_DE_IR <- Zhu_DE_IR %>%
  mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"))

Zhu_DE_IR <- Zhu_DE_IR %>% rename(intron_id = `...1`)

# how many are repeated

DE_iR_among_treatment_filter <- Zhu_DE_IR |> 
  group_by(intron_id) |> 
  summarise(Occurrences = n()) |> 
  filter(Occurrences > 1) |> 
  pull(intron_id)

Zhu_DE_IR |> 
  filter(intron_id %in% DE_iR_among_treatment_filter) |> 
  arrange(intron_id, Hours)
```

1. Extract clean Ensembl IDs
Use R's gsub() or stringr to isolate the Ensembl ID from your string pattern (SAMD11/ENSG00000187634/clean/1:924948-925921:+).

# Example vector of your IRFinder/DESeq2 row names
raw_ids <- c("SAMD11/ENSG00000187634/clean/1:924948-925921:+")

# Extract the Ensembl ID (matches ENSG followed by numbers)
clean_genes <- gsub(".*(ENSG[0-9]+).*", "\\1", raw_ids)
print(clean_genes) 
# Output: "ENSG00000187634"
2. Prepare your DESeq2 data
Filter your DESeq2 results to get your significant genes (e.g., \(p_{\text{adj}} < 0.05\)
# Load libraries
```{r}
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
keytypes(org.Hs.eg.db)  # confirm "ENSEMBL" is listed
AnnotationDbi::select(org.Hs.eg.db, keys = head(sig_genes, 10),
                       keytype = "ENSEMBL", columns = "SYMBOL")

sig_genes <- Zhu_DE_IR %>%
  filter(padj < 0.05) %>%
  pull(EnsemblID) %>%
  unique() %>%
  na.omit()

ego <- enrichGO(
  gene          = sig_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",       # or "MF", "CC", "ALL"
  pAdjustMethod = "BH",
  #qvalueCutoff  = 0.05,
  readable      = TRUE
)

dotplot(ego, showCategory=15)

dotplot(ego, showCategory = 20) + 
  ggplot2::ggtitle("GO Biological Processes for Intron Retention Targets")
```

```{r}
ego_test <- enrichGO(
     gene          = sig_genes,
    OrgDb         = org.Hs.eg.db,
     keyType       = "ENSEMBL",
     ont           = "BP",
     pAdjustMethod = "BH",
     pvalueCutoff  = 1,
     qvalueCutoff  = 1
)

nrow(as.data.frame(ego_test))

head(as.data.frame(ego_test)[, c("Description", "pvalue", "p.adjust", "Count")])
#                                                                     Description      pvalue  p.adjust Count
# GO:1901072                    glucosamine-containing compound catabolic process 0.005610739 0.1330355     1
# GO:0034145        positive regulation of toll-like receptor 4 signaling pathway 0.007846944 0.1330355     1
# GO:0046348                                        amino sugar catabolic process 0.007846944 0.1330355     1
# GO:0007601                                                    visual perception 0.007906551 0.1330355     2
# GO:0050953                                 sensory perception of light stimulus 0.008159344 0.1330355     2
# GO:1901533 negative regulation of hematopoietic progenitor cell differentiation 0.008405274 0.1330355     1
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
# > keytypes(org.Hs.eg.db)  # confirm "ENSEMBL" is listed
#  [1] "ACCNUM"       "ALIAS"        "ENSEMBL"      "ENSEMBLPROT"  "ENSEMBLTRANS" "ENTREZID"     "ENZYME"       "EVIDENCE"    
#  [9] "EVIDENCEALL"  "GENENAME"     "GENETYPE"     "GO"           "GOALL"        "IPI"          "MAP"          "OMIM"        
# [17] "ONTOLOGY"     "ONTOLOGYALL"  "PATH"         "PFAM"         "PMID"         "PROSITE"      "REFSEQ"       "SYMBOL"      
# [25] "UCSCKG"       "UNIPROT"     
AnnotationDbi::select(org.Hs.eg.db, keys = head(sig_genes, 10),
                        keytype = "ENSEMBL", columns = "SYMBOL")
 
sig_genes <- Zhu_DE_IR %>%
   filter(padj < 0.05) %>%
   pull(EnsemblID) %>%
   unique() %>%
   na.omit()
 
 ego <- enrichGO(
   gene          = sig_genes,
   OrgDb         = org.Hs.eg.db,
   keyType       = "ENSEMBL",
   ont           = "BP",       # or "MF", "CC", "ALL"
   pAdjustMethod = "BH",
   #qvalueCutoff  = 0.05,
   readable      = TRUE
)

dotplot(ego_test, showCategory=15)

```

```{r}
combined_data <- bind_rows(significant_DE_CC671_10uM_vs_DMSO_24hr, significant_DE_CC671_10uM_vs_DMSO_6hr, significant_DE_CC671_3uM_vs_DMSO_24hr, significant_DE_CC671_3uM_vs_DMSO_6hr, significant_velde_genes)

combined_data <- combined_data |> 
  mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`) 
  
```


```{r}
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database
keytypes(org.Hs.eg.db)  # confirm "ENSEMBL" is listed
AnnotationDbi::select(org.Hs.eg.db, keys = head(sig_genes, 10),
                       keytype = "ENSEMBL", columns = "SYMBOL")

combined_sig_genes <- combined_data %>%
  filter(padj < 0.05) %>%
  pull(EnsemblID) %>%
  unique() %>%
  na.omit()

ego_test <- enrichGO(
     gene          = combined_sig_genes,
    OrgDb         = org.Hs.eg.db,
     keyType       = "ENSEMBL",
     ont           = "BP",
     pAdjustMethod = "BH",
     pvalueCutoff  = 1,
     qvalueCutoff  = 1
)

nrow(as.data.frame(ego_test))

head(as.data.frame(ego_test)[, c("Description", "pvalue", "p.adjust", "Count")])
#                                                                     Description      pvalue  p.adjust Count
# GO:1901072                    glucosamine-containing compound catabolic process 0.005610739 0.1330355     1
# GO:0034145        positive regulation of toll-like receptor 4 signaling pathway 0.007846944 0.1330355     1
# GO:0046348                                        amino sugar catabolic process 0.007846944 0.1330355     1
# GO:0007601                                                    visual perception 0.007906551 0.1330355     2
# GO:0050953                                 sensory perception of light stimulus 0.008159344 0.1330355     2
# GO:1901533 negative regulation of hematopoietic progenitor cell differentiation 0.008405274 0.1330355     1

# > keytypes(org.Hs.eg.db)  # confirm "ENSEMBL" is listed
#  [1] "ACCNUM"       "ALIAS"        "ENSEMBL"      "ENSEMBLPROT"  "ENSEMBLTRANS" "ENTREZID"     "ENZYME"       "EVIDENCE"    
#  [9] "EVIDENCEALL"  "GENENAME"     "GENETYPE"     "GO"           "GOALL"        "IPI"          "MAP"          "OMIM"        
# [17] "ONTOLOGY"     "ONTOLOGYALL"  "PATH"         "PFAM"         "PMID"         "PROSITE"      "REFSEQ"       "SYMBOL"      
# [25] "UCSCKG"       "UNIPROT"     
AnnotationDbi::select(org.Hs.eg.db, keys = head(combined_sig_genes, 10),
                        keytype = "ENSEMBL", columns = "SYMBOL")
 
dotplot(ego_test, showCategory=15)

ego_df <- as.data.frame(ego_test)
```

```{r}
cg <- compareCluster(
  EnsemblID ~ Study,
  data          = combined_data,
  fun           = "enrichGO",
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 1
)

cg_df <- as.data.frame(cg)
head(cg_df)

dotplot(cg)
```


# Intron Retention

```{r}
IR_quantification <- read.delim("IR_summary_all_samples.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

library(readxl)
metadata_sc5903_mw <- read_csv("metadata_sc5903_mw.csv")

IR_quantification <- inner_join(
  IR_quantification, 
  metadata_sc5903_mw, 
  by = c("Sample" = "Irfinder file")
)
```

```{r}
IR_quantification |> 
  ggplot(aes(x= `Sequencing depth (FASTQC)`, y = Mean_IR_ratio, colour = Therapy)) + 
  geom_point()+
  facet_wrap(~Study)+
  theme_bw()
```
What I observe here is that DMSO (Control) typically has lower mean IR ratio than ones treated with the disease. Suggests that drugs work.

```{r}
IR_quantification |> 
  ggplot(aes(x= N_reliable_introns, y = Mean_IR_ratio, colour = Therapy, shape = Study)) + 
  geom_point()+
  theme_bw()
```

Doesn't appear to be strong correlation between number of reliable introns and the mean IR ratio

```{r}
IR_quantification |> 
  ggplot(aes(y= N_reliable_introns, x = `Sequencing depth (FASTQC)`, colour = Therapy, shape = Study)) + 
  geom_point()+
  theme_bw()
```
Number of reliable introns seems to be more dependant on the study rather than on the sequencing depth... 


```{r}
IR_quantification |> 
  filter(Therapy != c("0.1% DMSO", "DMSO")) |> 
  ggplot(aes(x= Median_IR_ratio, y = Mean_IR_ratio, colour = Hours, shape = Study)) + 
  geom_point()+
  theme_bw()
```

