---
title: "Special Topics Differential analysis"
author: "Monique White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)

Velde_T025_vs_DMSO_1uM_24h <- read_csv("T025_vs_DMSO_1uM_24h(in).csv")
```

Genes with an average read count > 10 in all samples were selected for differential gene expression analysis (n = 23,072). Genes with an absolute log2 fold change > 1 and FDR-adjusted p < 0.05 were considered significant. 

```{r}

signif_genes <- Velde_T025_vs_DMSO_1uM_24h %>%
  filter(baseMean > 10, log2FoldChange > 1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)

volc_data <- Velde_T025_vs_DMSO_1uM_24h %>% 
  filter(baseMean > 10, log2FoldChange > 1) |>
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

The number of IR events in a sample was determined based on introns with an IR ratio > 0.1 and meeting the filtering criteria described above. 

```{r}
library(data.table)

# Set the folder once
data_dir <- "C:/Users/Monique/OneDrive/Documents/SC5901 Biomedicine Lab/BrCa.Data"

files <- c(
  "30nM"  = file.path(data_dir, "30nM_vs_0nM.txt"),
  "100nM" = file.path(data_dir, "100nM_vs_0nM.txt"),
  "300nM" = file.path(data_dir, "300nM_vs_0nM.txt")
)

read_irfinder <- function(path, label) {
  dt <- fread(path, skip = 2, header = TRUE, sep = "\t")   # <-- use path, not a fixed string
  dt[, c("GeneName", "GeneID", "IntronType") := tstrsplit(`Intron-GeneName/GeneID`, "/", fixed = TRUE, keep = 1:3)]
  dt[, Comparison := label]
  dt
}

ir_list <- Map(read_irfinder, files, names(files))
ir_all <- rbindlist(ir_list, use.names = TRUE, fill = TRUE)

head(ir_all)
table(ir_all$Comparison)
```

```{r}
ir_all |> 
  filter(`p-increased` < 0.05 | `p-decreased` < 0.05) #filters from 22,072 to 56,785

ir_all |> 
  filter(`B-IRratio`>0.1, `p-increased` < 0.05 | `p-decreased` < 0.05) #1,457


ir_all |> 
  filter(Comparison == "30nM",
         `p-increased` < 0.05 | `p-decreased` < 0.05) #2,151 

ir_all |> 
  filter(Comparison == "100nM",
         `p-increased` < 0.05 | `p-decreased` < 0.05) #6,978

ir_all |> 
  filter(Comparison == "300nM",
         `p-increased` < 0.05 | `p-decreased` < 0.05) #12,943 

## Filtering based on IR ratio

ir_all |> 
  filter(Comparison == "30nM",
         `B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05) #306 

ir_all |> 
  filter(Comparison == "100nM",
         `B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05) #518

ir_all |> 
  filter(Comparison == "300nM",
         `B-IRratio`>0.1,
         `p-increased` < 0.05 | `p-decreased` < 0.05) #633
```

Seems to be a dose dependent effect there, with more introns