---
title: "Special Topics"
author: "Monique White"
output: html_document
date: "2026-03-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
# BiocManager::install("DESeq2")
# BiocManager::install("tximport")
#BiocManager::install("edgeR")
#BiocManager::install("clusterProfiler")
#BiocManager::install("org.Hs.eg.db")
library(clusterProfiler)
library(org.Hs.eg.db)
library(readr)
library(tidyverse)
library(edgeR)

library(DESeq2)
library(tximport)

bcv <- 0.1
```

Data downloaded from the https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE101540 https://pmc.ncbi.nlm.nih.gov/articles/PMC5991599/

```{r}
T025 <- read_tsv("GSE101540_raw_counts_GRCh38.p13_NCBI(1).tsv") # read in raw count tsv
```
GSM2705897 	COLO320_0nM
GSM2705898 	COLO320_30nM
GSM2705899 	COLO320_100nM
GSM2705900 	COLO320_300nM
GSM2705901 	MB468_0nM
GSM2705902 	MB468_30nM
GSM2705903 	MB468_100nM
GSM2705904 	MB468_300nM
GSM2705905 	H1048_0nM
GSM2705906 	H1048_30nM
GSM2705907 	H1048_100nM
GSM2705908 	H1048_300nM
GSM2705909 	786O_0nM
GSM2705910 	786O_30nM
GSM2705911 	786O_100nM
GSM2705912 	786O_300nM

I am interested in the breast cancer cells in particular, so I may select those columns

```{r}
T025_BR <-  T025 %>% 
  select(GeneID, GSM2705901, GSM2705902, GSM2705903, GSM2705904) #select all the breast cancer cells

# T025_BR <- T025_BR %>% 
#   mutate(
#     "MB468_0nM" = GSM2705901,
#     "MB468_30nM" = GSM2705902,
#     "MB468_100nM" = GSM2705903,
#     "MB468_300nM" = GSM2705904
#   )
# 
# #checking differece between reads between treatment concentration
# 
# T025_BR_calc <- T025_BR %>% 
#   mutate(
#     "300nm" = GSM2705904 - GSM2705901,
#     "100nm" = GSM2705903 - GSM2705901,
#     "30nm" = GSM2705902 - GSM2705901
#   )

```

edgeR is primarily intended for use with data including biological replication. Nevertheless,
RNA-Seq and ChIP-Seq are still expensive technologies, so it sometimes happens that only
one library can be created for each treatment condition. In these cases there are no replicate
libraries from which to estimate biological variability. In this situation, the data analyst is
faced with the following choices, none of which are ideal. We do not recommend any of
these choices as a satisfactory alternative for biological replication. Rather, they are the best
that can be done at the analysis stage, and options 2Ű4 may be better than assuming that biological variability is absent.
1. Be satisĄed with a descriptive analysis, that might include an MDS plot and an analysis
of fold changes. Do not attempt a signiĄcance analysis. This may be the best advice.
2. Simply pick a reasonable dispersion value, based on your experience with similar data,
and use that for exactTest or glmFit. Typical values for the common BCV (square-rootdispersion) for datasets arising from well-controlled experiments are 0.4 for human data,
0.1 for data on genetically identical model organisms or 0.01 for technical replicates.
Here is a toy example with simulated data:

*The typical BCV value of 0.4 for human data.*

```{r}
bcv <- 0.1
```

# Breast Cancer Cells

```{r}
# convert to matrix
count_mat_br <- as.matrix(T025_BR[, -(1)])

# set gene IDs as rownames
rownames(count_mat_br) <- T025_BR$GeneID

group <- factor(c("0nM", "30nM", "100nM", "300nM"))
```


## 0nM vs 30nM

```{r}
y <- DGEList(counts = count_mat_br, group = group)

#perform exact test
et <- exactTest(y, dispersion = bcv^2)

et_table <- et$table
```
```{r}
# 0nm vs 30nm
et_MB4680nm_30nm <- exactTest(y, pair = c("0nM","30nM"), dispersion = bcv^2)

res_MB4680nm_30nm <- et_MB4680nm_30nm$table %>%
  rownames_to_column("GeneID")

signif_genes <- res_MB4680nm_30nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(PValue<0.05)

volc_data <- res_MB4680nm_30nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

###  Gene Enrichment

```{r}
library(clusterProfiler)
library(org.Hs.eg.db) 
```

```{r}
gene_list <- signif_genes$GeneID
gene_list <- as.character(gene_list)

ego <- enrichGO(gene          = gene_list,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",   # try also "MF" and "CC"
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
results_df <- as.data.frame(ego)
head(results_df)
```

```{r}
background_genes <- rownames(res_MB4680nm_30nm)

ego <- enrichGO(gene          = gene_list,
                universe      = background_genes,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH")

results_df <- as.data.frame(ego)
head(results_df)
```

## 0nM vs 100nM

```{r}
# 0nm vs 100nm
et_MB468_0nm_100nm <- exactTest(y, pair = c("0nM","100nM"), dispersion = bcv^2)

res_MB468_0nm_100nm <- et_MB468_0nm_100nm$table %>%
  rownames_to_column("GeneID")

signif_genes <- res_MB468_0nm_100nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(PValue<0.05)

volc_data <- res_MB468_0nm_100nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

```{r}
signif_genes <- res_MB468_0nm_100nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(-log10(PValue) > 30) # don't do this because it then goes back to light perception which isn't really what we want to look at

volc_data <- res_MB468_0nm_100nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

```{r}
gene_list <- signif_genes$GeneID
gene_list <- as.character(gene_list)

ego <- enrichGO(gene          = gene_list,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",   # try also "MF" and "CC"
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

results_df <- as.data.frame(ego)
head(results_df)
```



```{r}
background_genes <- rownames(res_MB468_0nm_100nm)

ego <- enrichGO(gene          = gene_list,
                universe      = background_genes,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH")
results_df <- as.data.frame(ego)
head(results_df)
```

## 0nM vs 300nM
```{r}
# 0nm vs 300nm
et_MB4680nm_300nm <- exactTest(y, pair = c("0nM","300nM"), dispersion = bcv^2)

res_MB4680nm_300nm <- et_MB4680nm_300nm$table %>%
  rownames_to_column("GeneID")

signif_genes <- res_MB4680nm_300nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(PValue<0.05)

volc_data <- res_MB4680nm_300nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```
```{r}
gene_list <- signif_genes$GeneID
gene_list <- as.character(gene_list)

ego <- enrichGO(gene          = gene_list,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",   # try also "MF" and "CC"
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

results_df <- as.data.frame(ego)
head(results_df)
```

```{r}
gene_list <- signif_genes$GeneID
gene_list <- as.character(gene_list)

ego <- enrichGO(gene          = gene_list,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "MF",   # try also "MF" and "CC"
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

results_df <- as.data.frame(ego)
head(results_df)
```


Looking at AFAP1

```{r}
T025_BR %>% 
  filter(GeneID %in% c("24017", "60312"))
```


Hmm.. That is interesting. Its expression seems to get lower as the concentration of T025 is increased

```{r}
# et_table

et_tibble <- rownames_to_column(et_table, var = "GeneID")

# et_tibble %>% 
#   filter(GeneID == "60312")
```


# 786O cancer cell line
```{r}
T025_786O <- T025 %>% select(GeneID,GSM2705909,GSM2705910,GSM2705911, GSM2705912)
# convert to matrix
count_mat_786O <- as.matrix(T025_786O[, -(1)])

# set gene IDs as rownames
rownames(count_mat_786O) <- T025_786O$GeneID

group <- factor(c("0nM", "30nM", "100nM", "300nM"))
```

```{r}
y <- DGEList(counts = count_mat_786O, group = group)

#perform exact test
et <- exactTest(y, dispersion = bcv^2)

et_table <- et$table
```

## 0nm vs 30nm
```{r}
et_786O_30nm <- exactTest(y, pair = c("0nM","30nM"), dispersion = bcv^2)

res_786O_30nm <- et_786O_30nm$table %>%
  rownames_to_column("GeneID")

signif_genes <- res_786O_30nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(PValue<0.05)

volc_data <- res_786O_30nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

## 0nM vs 100nM
```{r}
et_786O_100nm <- exactTest(y, pair = c("0nM","100nM"), dispersion = bcv^2)

res_786O_100nm <- et_786O_100nm$table %>%
  rownames_to_column("GeneID")

signif_genes <- res_786O_100nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(PValue<0.05)

volc_data <- res_786O_100nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```

## 0nM vs 300nM
```{r}
et_786O_300nm <- exactTest(y, pair = c("0nM","300nM"), dispersion = bcv^2)

res_786O_300nm <- et_786O_300nm$table %>%
  rownames_to_column("GeneID")

signif_genes <- res_786O_300nm %>% 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(PValue) %>% 
  filter(PValue<0.05)

volc_data <- res_786O_300nm %>% 
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = logFC, y = -log10(PValue))) + 
  geom_point() +
  geom_point(data = signif_genes, color="red")
```


## Compare baseline levels of AFAP1-AS1 in Different cell lines

GSM2705897 	COLO320_0nM
GSM2705901 	MB468_0nM
GSM2705905 	H1048_0nM
GSM2705909 	786O_0nM

```{r}
T025_0nm <-  T025 %>% 
  select(GeneID, GSM2705897, GSM2705901, GSM2705905, GSM2705909) #select all cell lines at 0nm
```


```{r}
# convert to matrix
count_mat_0nm <- as.matrix(T025_0nm[, -(1)])

# set gene IDs as rownames
rownames(count_mat_0nm) <- T025_0nm$GeneID

group_0nm <- factor(c("COLO320", "MB468", "H1048", "786O"))
```



```{r}
y_0nm <- DGEList(counts = count_mat_0nm, group = group_0nm)

# normalization
y_0nm <- calcNormFactors(y_0nm)

# # estimate dispersion
# y_0nm <- estimateDisp(y_0nm)

#perform exact test
et_0nm <- exactTest(y_0nm, dispersion = bcv^2)

et_table_0nm <- et_0nm$table
```

```{r}
et_tibble_0nm <- rownames_to_column(et_table_0nm, var = "GeneID")

et_tibble_0nm %>% 
  filter(GeneID == "60312")
```


```{r}
# MB468 vs H1048
et_MB468_H1048 <- exactTest(y_0nm, pair = c("MB468","H1048"), dispersion = bcv^2)

res_MB468_H1048 <- et_MB468_H1048$table %>%
  rownames_to_column("GeneID")

# MB468 vs 786O
et_MB468_786O <- exactTest(y_0nm, pair = c("MB468","786O"), dispersion = bcv^2)

res_MB468_786O <- et_MB468_786O$table %>%
  rownames_to_column("GeneID")
```

```{r}
# MB468 vs COLO320
et_MB468_COLO320 <- exactTest(y_0nm, pair = c("MB468","COLO320"), dispersion = bcv^2)

res_MB468_COLO320 <- et_MB468_COLO320$table %>%
  rownames_to_column("GeneID")


# MB468 vs H1048
et_MB468_H1048 <- exactTest(y_0nm, pair = c("MB468","H1048"), dispersion = bcv^2)

res_MB468_H1048 <- et_MB468_H1048$table %>%
  rownames_to_column("GeneID")


# MB468 vs 786O
et_MB468_786O <- exactTest(y_0nm, pair = c("MB468","786O"), dispersion = bcv^2)

res_MB468_786O <- et_MB468_786O$table %>%
  rownames_to_column("GeneID")
```


```{r}
res_MB468_COLO320 %>% filter(GeneID == "60312")
res_MB468_H1048 %>% filter(GeneID == "60312")
res_MB468_786O %>% filter(GeneID == "60312")
```
AFAP1-AS1 is highest in Renal Cell Carcinoma cell line (7860), followed by Breast Cancer cell line (MB468).
See if treatment with T025 also reduces the expression of this gene as well, and whether the differential expression is greater in Breast Cancer cells or Not

