---
title: "Meeting Ulf 21/07/2026"
author: "Monique White"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(data.table)
library(ggpubr)
library(clusterProfiler)
library(org.Hs.eg.db) # Human annotation database

## Velde et al. 2025
Velde_T025_vs_DMSO_1uM_24h <- read_csv("ir_results/ERP171667/T025_vs_DMSO_1uM_24h.csv")
Velde_T025_vs_DMSO_1uM_24h <- Velde_T025_vs_DMSO_1uM_24h |>
  mutate(
    "Hours" = 6,
    "Concentration" = "1uM",
    "Study" = "Velde"
  )

Velde_T025_vs_DMSO_1uM_24h <- Velde_T025_vs_DMSO_1uM_24h |> mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`,
    direction = case_when(
      log2FoldChange > 0 ~ "upregulated",
      log2FoldChange < 0 ~ "downregulated",
      TRUE ~ "no change"
    )
  )

## Zhu et al. 2018 Data

#### Pooled meaning that the IR filtering was calculated on a global list where the intron had to be present in 50% of all 16 samples...
# CC671_3uM_vs_DMSO_6hr <- read_csv("ir_results/SRP141379/CC671_3uM_vs_DMSO_6hr.csv")
# CC671_3uM_vs_DMSO_24hr <- read_csv("ir_results/SRP141379/CC671_3uM_vs_DMSO_24hr.csv")
# CC671_10uM_vs_DMSO_6hr <- read_csv("ir_results/SRP141379/CC671_10uM_vs_DMSO_6hr.csv")
# CC671_10uM_vs_DMSO_24hr <- read_csv("ir_results/SRP141379/CC671_10uM_vs_DMSO_24hr.csv")

#Thus, using time specific because the dispersion diagnostics

CC671_3uM_vs_DMSO_6hr <- read_csv("ir_results/SRP141379/6hr/CC671_3uM_vs_DMSO_6hr.csv")
CC671_3uM_vs_DMSO_24hr <- read_csv("ir_results/SRP141379/24hr/CC671_3uM_vs_DMSO_24hr.csv")
CC671_10uM_vs_DMSO_6hr <- read_csv("ir_results/SRP141379/6hr/CC671_10uM_vs_DMSO_6hr.csv")
CC671_10uM_vs_DMSO_24hr <- read_csv("ir_results/SRP141379/24hr/CC671_10uM_vs_DMSO_24hr.csv")



CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr|>
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM",
    "Study" = "Zhu"
  )

CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr |> mutate(
    "Hours" = 24,
    "Concentration" = "3uM",
    "Study" = "Zhu"
  )

CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr |> mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu"
  )

CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr|> mutate(
    "Hours" = 24,
    "Concentration" = "10uM",
    "Study" = "Zhu"
  )

#Quality control data
IR_quantification <- read.delim("IR_summary_all_samples.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

library(readxl)
metadata_sc5903_mw <- read_csv("metadata_sc5903_mw.csv")

IR_quantification <- inner_join(
  IR_quantification, 
  metadata_sc5903_mw, 
  by = c("Sample" = "Irfinder file")
)
```

# Intron Retention

```{r echo = FALSE}
IR_quantification |> 
  ggplot(aes(x= log(`Sequencing depth (FASTQC)`), y = Mean_IR_ratio, colour = exp_design)) + 
  geom_point()+
  facet_wrap(~Study, ncol =1)+
  theme_bw()
```


What I observe here is that DMSO (Control) typically has lower mean IR ratio than ones treated with the disease. Suggests that drugs work.

```{r echo = FALSE}
IR_quantification |> 
  ggplot(aes(x= N_reliable_introns, y = Mean_IR_ratio, colour = exp_design, shape = Study)) + 
  geom_point(size =3)+
  theme_bw()+
  labs(colour = "Treatment",
       x = "Reliable introns (n)",
       y = "Mean IR ration")
```

Doesn't appear to be strong correlation between number of reliable introns and the mean IR ratio. Perhaps weak negative.

```{r echo = FALSE}
IR_quantification |> 
  ggplot(aes(x=N_reliable_introns, y=Mean_IR_ratio, colour = `Cell line`))+
  geom_point(size = 3)+
  theme_bw()+
  labs(y= "Mean IR ratio",
       x = "Reliable introns")+
  facet_wrap(~exp_design)
```



# Intron Retention

## Velde et al. 2026 ERR...

```{r}
significant_velde_genes <- Velde_T025_vs_DMSO_1uM_24h |> 
  filter(baseMean > 10, log2FoldChange > 5 | log2FoldChange < -5) |>
  # filter(baseMean > 10, log2FoldChange > 1 | log2FoldChange < -1) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.05)

volc_data <- Velde_T025_vs_DMSO_1uM_24h %>% 
  filter(baseMean > 10#, 
         #log2FoldChange > 1 | log2FoldChange < -1
         ) |>
  as.data.frame() %>% 
  na.omit()

ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_velde_genes, color="red")+
  theme_bw()

# ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
#   geom_point() +
#   geom_point(data = significant_velde_genes, color="red")+
#   theme_bw() |> 
#   ggsave(filename = "velde_ir_volc.pdf", 
#          height = 150, width = 150, unit = "mm")
```

## Zhu et al. 2018 SRP141379

Significant de ir 
filter(baseMean > 10, log2FoldChange > 3 | log2FoldChange < -3)
filter(padj<0.01)

ir events included in the plot:
baseMean > 10 

```{r echo = FALSE}
significant_DE_CC671_3uM_vs_DMSO_6hr <- CC671_3uM_vs_DMSO_6hr %>%
  filter(baseMean > 10, log2FoldChange > 3 | log2FoldChange < -3) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.01)|> 
  mutate(
    "Hours" = 6,
    "Concentration" = "3uM", 
    "Study" = "Zhu"
  )
volc_data <- CC671_3uM_vs_DMSO_6hr %>% 
  filter(baseMean > 10 #, 
         #log2FoldChange > 1 | log2FoldChange < -1
         ) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_3uM_6hr <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_3uM_vs_DMSO_6hr, color="red")+
  theme_bw()+
  labs(title = "3uM, 6hr")

significant_DE_CC671_3uM_vs_DMSO_24hr <- CC671_3uM_vs_DMSO_24hr %>%
  filter(baseMean > 10, log2FoldChange > 3 | log2FoldChange < -3) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.01)|> 
  mutate(
    "Hours" = 24,
    "Concentration" = "3uM", 
    "Study" = "Zhu"
  )
volc_data <- CC671_3uM_vs_DMSO_24hr %>% 
  filter(baseMean > 10 #, 
         #log2FoldChange > 1 | log2FoldChange < -1
         ) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_3uM_24hr <- ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_3uM_vs_DMSO_24hr, color="red")+
  theme_bw()+
  labs(title = "3uM, 24hr")

significant_DE_CC671_10uM_vs_DMSO_6hr <- CC671_10uM_vs_DMSO_6hr %>%
  filter(baseMean > 10, log2FoldChange > 3 | log2FoldChange < -3) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.01)|> 
  mutate(
    "Hours" = 6,
    "Concentration" = "10uM",
    "Study" = "Zhu"
  )

volc_data <- CC671_10uM_vs_DMSO_6hr %>% 
  filter(baseMean > 10#, 
         #log2FoldChange > 3 | log2FoldChange < -3
         ) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_10uM_6hr <- ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_10uM_vs_DMSO_6hr, color="red")+
  theme_bw()+
  labs(title = "10uM, 6hr")


significant_DE_CC671_10uM_vs_DMSO_24hr <- CC671_10uM_vs_DMSO_24hr %>%
  filter(baseMean > 10, log2FoldChange > 3 | log2FoldChange < -3) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.01) |> 
  mutate(
    "Hours" = 24,
    "Concentration" = "10uM", 
    "Study" = "Zhu"
  )

volc_data <- CC671_10uM_vs_DMSO_24hr %>% 
  filter(baseMean > 10#, 
         #log2FoldChange > 1 | log2FoldChange < -1
         ) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_10uM_24hr <- ggplot(volc_data,aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = significant_DE_CC671_10uM_vs_DMSO_24hr, color="red")+
  theme_bw()+
  theme(panel.grid.minor = element_blank())+
  labs(title = "10uM, 24hr")
```


```{r}
ggarrange(volc_plot_3uM_6hr, volc_plot_10uM_6hr, volc_plot_3uM_24hr, volc_plot_10uM_24hr,
          ncol =2, nrow = 2,
          labels = c("A", "B", "C", "D"))
```

### how intron retention events are diferentially expressed at both 6 hours and 24 hours?

```{r}
Zhu_DE_IR <- bind_rows(significant_DE_CC671_3uM_vs_DMSO_6hr, significant_DE_CC671_3uM_vs_DMSO_24hr, significant_DE_CC671_10uM_vs_DMSO_6hr, significant_DE_CC671_10uM_vs_DMSO_24hr)

library(stringr)

Zhu_DE_IR <- Zhu_DE_IR %>%
  mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = ...1)

# how many are present in both 6 hours and 24 hours
DE_iR_among_treatment_filter <- Zhu_DE_IR |> 
  group_by(intron_id) |> 
  summarise(Occurrences = n()) |> 
  filter(Occurrences > 1) |> 
  pull(intron_id) #468 are repeated in all combinations of drug and timepoint

Zhu_DE_IR |> 
  group_by(intron_id, Concentration) |> 
  summarise(Occurrences = n_distinct(Hours)) |> 
  filter(Occurrences > 1) #|> 
  #pull(intron_id) 
#103 are repeated in more than one timepoint, within a concentration

Zhu_DE_IR |> 
  group_by(intron_id, Concentration) |> 
  summarise(Occurrences = n_distinct(Hours)) |> 
  filter(Occurrences == 1) #|> 
  #pull(intron_id) 
#1116 introns that occur in only one timepoint for a given concentration
```


```{r}
combined_data <- bind_rows(significant_DE_CC671_10uM_vs_DMSO_24hr, significant_DE_CC671_10uM_vs_DMSO_6hr, significant_DE_CC671_3uM_vs_DMSO_24hr, significant_DE_CC671_3uM_vs_DMSO_6hr, significant_velde_genes)

combined_data <- combined_data |> 
  mutate(EnsemblID = str_extract(`...1`, "ENSG[0-9]+"),
         intron_id = `...1`,
    direction = case_when(
      log2FoldChange > 0 ~ "upregulated",
      log2FoldChange < 0 ~ "downregulated",
      TRUE ~ "no change"
    )
  )

shared_de_ir <- combined_data |> group_by(intron_id) |>
  summarise(count = n_distinct(Study)) |> 
  filter(count > 1) |> 
  pull(intron_id) #shared de ir is 199 genes

#combined_data |> filter(intron_id %in% shared_de_ir) |> write.csv(file = "shared_de_ir.csv")
length(shared_de_ir)
```


## shared de intron retentions between studies gene ontology

Not sure why this is not working. there are 199 IR events, should be many more genes 
```{r}
shared_de_ir_df <- combined_data |> filter(intron_id %in% shared_de_ir)

combined_sig_ir <- shared_de_ir_df %>%
  filter(padj < 0.05) %>%
  pull(EnsemblID) %>%
  unique() %>%
  na.omit()

ego_test <- enrichGO(
     gene          = combined_sig_ir,
    OrgDb         = org.Hs.eg.db,
     keyType       = "ENSEMBL",
     ont           = "BP",
     pAdjustMethod = "BH",
     pvalueCutoff  = 1,
     qvalueCutoff  = 1
)

sig_ir <- as.data.frame(ego_test)

dotplot(ego_test)
```

# Gene Expression

Gene count matrices were generated using the Sorted.bam files from IRFinder, which aligned to the Human-GRCh38-release 100.
These bam files were converted into a gene count matrix using featureCounts.
The flags included,
-p
-P
-B
Meaning only reads where both forward and reverse reads aligned to the genome.


## Velde et al. 2026

DEseq was run for these using the following code
```{r eval = FALSE}
# ── Custom IRFinder DESeq2 function (Kept for Raw File Parsing) ─────────────
DESeqDataSetFromIRFinder = function(filePaths, designMatrix, designFormula){
  res=c()
  libsz=c()
  spl=c()
  irtest=read.table(filePaths[1])
  if (irtest[1,1]=="Chr"){irtest=irtest[-1,]}
  irnames=unname(apply(as.matrix(irtest),1,FUN=function(x){return(paste0(x[4],"/",x[1],":",x[2],"-",x[3],":",x[6]))}))
  n=1
  for (i in filePaths){
    print(paste0("processing file ",n," at ",i))
    irtab=read.table(i)
    if (irtab[1,1]=="Chr"){irtab=irtab[-1,]}
    tmp1=as.numeric(as.vector(irtab[,9]))
    tmp2=as.numeric(as.vector(irtab[,19]))
    tmp3=tmp1+tmp2
    tmp4=as.numeric(as.vector(irtab[,17]))
    tmp5=as.numeric(as.vector(irtab[,18]))
    tmp6=pmax(tmp4,tmp5, na.rm=T)
    res=cbind(res,tmp1)
    libsz=cbind(libsz,tmp2)
    spl=cbind(spl,tmp6)
    n=n+1
  }
  res.rd=round(res)
  libsz.rd=round(libsz)
  spl.rd=round(spl)
  colnames(res.rd)=paste("intronDepth",as.vector(designMatrix[,1]),sep=".")
  rownames(res.rd)=irnames
  colnames(libsz.rd)=paste("totalSplice",as.vector(designMatrix[,1]),sep=".")
  rownames(libsz.rd)=irnames
  colnames(spl.rd)=paste("maxSplice",as.vector(designMatrix[,1]),sep=".")
  rownames(spl.rd)=irnames
  
  ir=c(rep("IR",dim(designMatrix)[1]),rep("Splice",dim(designMatrix)[1]))
  group=rbind(designMatrix,designMatrix)
  group$IRFinder=ir
  group$IRFinder=factor(group$IRFinder,levels=c("Splice","IR"))
  
  counts.IRFinder=cbind(res.rd,spl.rd)
  
  dd = DESeqDataSetFromMatrix(countData = counts.IRFinder, colData = group, design = designFormula)
  sizeFactors(dd)=rep(1,dim(group)[1])
  rownames(dd)=irnames
  final=list(dd,res,libsz,spl)
  names(final)=c("DESeq2Object","IntronDepth","SpliceDepth","MaxSplice")
  return(final)
}

# ── Helpers ──────────────────────────────────────────────────
# Automatically maps raw paths to their corresponding filtered text file counterpart
filtered_path_for <- function(raw_path) {
  dir <- dirname(raw_path)
  sample <- sub("_irfinder$", "", basename(dir))
  file.path(dir, paste0("filtered_ir_nondir_", sample, ".txt"))
}

get_testable_introns <- function(raw_paths, min_frac = 0.5) {
  filtered_paths <- vapply(raw_paths, filtered_path_for, character(1))
  missing <- filtered_paths[!file.exists(filtered_paths)]
  if (length(missing) > 0) {
    stop("Missing filtered file(s):\n  ", paste(missing, collapse = "\n  "),
         "\nRun filter_irfinder.R first for these samples.")
  }
  id_lists <- lapply(filtered_paths, function(fp) {
    df <- read.delim(fp, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
    paste0(df$Gene, "/", df$Chr, ":", df$Start, "-", df$End, ":", df$Strand)
  })
  all_ids <- unlist(id_lists)
  tab <- table(all_ids)
  n <- length(filtered_paths)
  names(tab)[(tab / n) >= min_frac]
}

dir.create("results/ERP171667", recursive = TRUE, showWarnings = FALSE)

# ==============================================================================
# ERP171667 DATA ANALYSIS
# ==============================================================================
# CORRECTED: Point these directly to the raw files so matrix building functions properly
paths <- c(
  file.path("/bioinformatics/moniquew/BrCa-Special-Topics/raw_data/IR_quantification/ERP171667/ERR14856082_irfinder/IRFinder-IR-nondir.txt"),
  file.path("/bioinformatics/moniquew/BrCa-Special-Topics/raw_data/IR_quantification/ERP171667/ERR14856083_irfinder/IRFinder-IR-nondir.txt"),
  file.path("/bioinformatics/moniquew/BrCa-Special-Topics/raw_data/IR_quantification/ERP171667/ERR14856084_irfinder/IRFinder-IR-nondir.txt"),
  file.path("/bioinformatics/moniquew/BrCa-Special-Topics/raw_data/IR_quantification/ERP171667/ERR14856085_irfinder/IRFinder-IR-nondir.txt"),
  file.path("/bioinformatics/moniquew/BrCa-Special-Topics/raw_data/IR_quantification/ERP171667/ERR14856086_irfinder/IRFinder-IR-nondir.txt"),
  file.path("/bioinformatics/moniquew/BrCa-Special-Topics/raw_data/IR_quantification/ERP171667/ERR14856087_irfinder/IRFinder-IR-nondir.txt")
)

experiment <- data.frame(
  condition = factor(
    c("DMSO","DMSO","DMSO","T025","T025","T025"),
    levels = c("DMSO","T025")
  )
)

# Initialize with interaction formula layout
metaList <- DESeqDataSetFromIRFinder( 
  filePaths     = paths,
  designMatrix  = experiment,
  designFormula = ~ condition + IRFinder + condition:IRFinder
)

# ==============================================================================
# TWO-STAGE FILTERING STEP
# ==============================================================================
# Stage 1: Get high quality whitelisted introns from filtered files (min_frac check)
testable_introns <- get_testable_introns(paths, min_frac = 0.5)

# Map valid row names back to matrices to prevent bounds errors
intron_raw_mat <- metaList$IntronDepth
rownames(intron_raw_mat) <- rownames(metaList$DESeq2Object)
intron_filtered_mat <- intron_raw_mat[testable_introns, ]

splice_raw_mat <- metaList$SpliceDepth
rownames(splice_raw_mat) <- rownames(metaList$DESeq2Object)
splice_filtered_mat <- splice_raw_mat[testable_introns, ]

# Stage 2: Calculate actual IR ratios and filter for active signal (IR > 0.1)
all_ir_ratios <- intron_filtered_mat / (intron_filtered_mat + splice_filtered_mat + 0.01)

# Retain introns where at least 3 samples (50% of cohort) show an active IR ratio > 0.1
high_ir_introns <- rownames(all_ir_ratios)[rowSums(all_ir_ratios > 0.1) >= 3]

message("ERP171667: ", length(testable_introns), " introns pass quality screening.")
message("Value Filtering: ", length(high_ir_introns), " / ", length(testable_introns), 
        " introns also pass baseline expression threshold (IR ratio > 0.1 in >=3 samples).")

# Final clean subset of the dds object
dds <- metaList$DESeq2Object[rownames(metaList$DESeq2Object) %in% high_ir_introns, ]
dds <- DESeq(dds)

print("Available coefficients for ERP171667:")
print(resultsNames(dds))

# Extract the interaction term
res <- results(dds, name = "conditionT025.IRFinderIR")

write.csv(as.data.frame(res),
          file = "results/ERP171667/T025_vs_DMSO_1uM_24h.csv")

message("T-025 1uM 24h vs DMSO (Differential IR): ",
        sum(res$padj < 0.05, na.rm = TRUE),
        " significant introns")
```

Then csv was saved, and the remaining analysis was done on my laptop.
```{r echo = FALSE}
Velde_gene_de_T025_DMSO <- read_csv("deseq_gene/deseq2_results/ERP171667/T025_vs_DMSO.csv")
```

Gene included had a base mean expression greater than 10
Significant (shown in red) represent genes whose base mean was greater than 10, and had a log2FoldChange of 3 or above, or below -3. 

```{r}
Velde_gene_de_T025_DMSO <- Velde_gene_de_T025_DMSO |> 
  mutate(
    "EnsemblID" = `...1`,
    "Contrast" = "T025_vs_DMSO",
    "Study" = "Velde",
    "Hours" = 6,
    "Concentration" = "1uM")

sig_Velde_gene_de_T025_DMSO <- Velde_gene_de_T025_DMSO |> 
  filter(baseMean > 10, log2FoldChange > 3 | log2FoldChange < -3) |> 
  as.data.frame() %>% 
  na.omit() %>% 
  #rownames_to_column("GeneID") %>% 
  arrange(padj) %>% 
  filter(padj<0.01)

volc_data <- Velde_gene_de_T025_DMSO  %>% 
  filter(baseMean > 10 #, 
         #log2FoldChange > 1 | log2FoldChange < -1
         ) |>
  as.data.frame() %>% 
  na.omit()

volc_plot_velde <- ggplot(volc_data, aes(x = log2FoldChange, y = -log10(padj))) + 
  geom_point() +
  geom_point(data = sig_Velde_gene_de_T025_DMSO, color="red")+
  theme_bw()+
  labs(title = "Differential gene expression, T-025 1uM")

volc_plot_velde
```


```{r}
upreg_ir_velde <- Velde_T025_vs_DMSO_1uM_24h |> 
  filter(direction == "upregulated") |> 
  pull(EnsemblID)

Velde_gene_de_T025_DMSO |> filter(EnsemblID %in% upreg_ir_velde) |> summarise(n = n_distinct(EnsemblID)) 
```
1567 genes that have differential gene expression have an intron retention event that is upregulated in treated cells compared to non-treated cells.

```{r}
Velde_gene_de_T025_DMSO |> filter(EnsemblID %in% upreg_ir_velde,
                                  log2FoldChange > 0) |> 
  summarise(n = n_distinct(EnsemblID))
```

513 genes that are upregulated in treated cells have intron retention events that are upregulated in treated cells compared to non-treated cells.

```{r}
Velde_gene_de_T025_DMSO |> filter(EnsemblID %in% upreg_ir_velde,
                                  log2FoldChange < 0) |> 
  summarise(n = n_distinct(EnsemblID))
```

1054 genes that are downregulated in treated cells have upregulation of intron retention in them in treated cells compared to non-treated cells.

## Zhu gene expression
A little more complex, as there are timepoints and concentrations

